### 2018-10-02 unpublished/pedestrian/pedestrianInteraction.tex
adding in writeup on pedestrian stuff

% Document settings
\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{amsmath,amssymb,gensymb,epsfig, epstopdf}
\usepackage{xcolor}
\pagestyle{plain}
\setlength\parindent{0pt}
\setlength{\parskip}{1em}

\begin{document}

% Header
\textbf{Nitin Kapania} \\
\textbf{Modeling Pedestrian - AV Interaction at a Mid-Block Intersection} \\

\section{Problem Formulation}
Consider the problem of an autonomous vehicle attempting to drive past an unsignalized, ``mid-block" crosswalk. A pedestrian is simultaneously walking towards the intersection.

\begin{center}
   \includegraphics[width=.8\textwidth]{Images/example.eps}
   \label{img:example}
\end{center}

In this case, what should the autonomous vehicle do? Should the vehicle always stop as soon as a pedestrian is observed near a cross-walk? Can the AV make the decision to speed up and cross before the pedestrian is in the intersection? If so, how can it communicate intent to the pedestrian? Likewise, can it interpret signals from the pedestrian to understand whether the pedestrian is set on crossing?

Uncontrolled mid-block intersections, like the one shown in Fig.~\ref{img:example}, pose the ambiguity of right of way. These laws are dictated on a state-by-state basis. While several states (e.g. Minnesota, Georgia, Maryland) have "must-stop" rules for uncontrolled intersections, other states have ``must yield" rules. For example, vehicles in Arkansas must yield once a pedestrian is in any portion of the roadway, while vehicles in Alabama require motorists to yield to a pedestrian either on the same half of the road or approaching from the opposite side of the roadway.

Given this ambiguity, there is a clear need to study the interaction between the pedestrian and the autonomous vehicle. Most human drivers and pedestrians do not know the explicit laws of their state, but safely navigate pedestrian interactions nonetheless.

Several research questions can be posed for the following situation:

\begin{itemize}  
	\item Can we come up with a mathematical formulation describing this situation? 
	\item Can we superimpose this mathematical model with real-world data of pedestrian intersection crossings? 
	\item Can this formulation, combined with real-world data, be used to predict realistic 
	interaction scenarios?
	\item Can we design control policies for the vehicle that maintain a high standard of safety while also enabling robust traffic flow 
\end{itemize}

\section{Prior Work}

There is a wide body of literature segmenting the work on pedestrian behavior. In terms of purely modeling the dynamics of pedestrian motion, Batkovic et al. \cite{Batkovic} presented a mathematical model to predict pedestrian motion over a finite horizon, with a focus on low computational complexity for autonomous driving. Karasev et al. modeled pedestrian behavior as intent using a Markov Decision Process \cite{Karasev2016}.

Another segment of the literature is focused on developing predictive models for pedestrian crossing. These approaches typically study the parallel issues of pedestrian crossing behavior or driver yielding behavior and date as early as 1975 \cite{Katz1975}. Schroeder and Rouphail \cite{Schroeder2011} explored factors associated with driver yielding behavior at unsignalized pedestrian crossings. Using logistic regression, the authors found that drivers are more likely to yield to assertive pedestrians who walk briskly in their approach to a crosswalk. Kadali and Perumal \cite{RaghuramKadali2012} studied the ``gap acceptance" behavior of pedestrians at mid-block crosswalks through a video graphic survey, and found that the gap accepted by crossing was explained by factors such as crossing direction, vehicle speed, and pedestrian age. Yannis et al. \cite{Yannis2013} also found that gap acceptance was influenced by the size of the oncoming vehicle and the presence of other pedestrians. Gap acceptance models were also proposed by \cite{Sun2002}. Lee and Aty \cite{Lee2005} studied interactions in the form of crashes, and found crashes were linked with higher daily traffic.

Another branch of literature specifically studies the interaction between pedestrians and autonomous vehicles. A review of these studies was conducted by \cite{Rasouli}. Rothenbucher et al. \cite{Rothenbucher2016} studied the interaction between pedestrians and driverless vehicles by constructing a car seat costume to disguise a driver. The authors noted that pedestrians overall managed interactions at crosswalks effectively, but later mentioned increased uncertainty about the car's behavior. To improve the issue of trust, several researchers have developed external interfaces to more clearly broadcast the intent of the autonomous vehicle \cite{Matthews}, \cite{Lagstrom2015}.

Several real-world studies of AV-pedestrian interaction noted that once a local population learned a vehicle was programmed to be perfectly safe, pedestrians would regularly take advantage of the AV and walk in front of it. \cite{Madigan}. This illustrates an issue with automated driving in that an overly conservative crossing algorithm will often be taken advantage of or cause confusion among pedestrians. Camara et al \cite{Camara2018} attempted to model the natural negotation for priority between a pedestrian and an AV at an intersection using game theory.

\section{Proposed Research}

\bibliographystyle{plain}
\bibliography{Bibliography}

\end{document}

### 2019-01-09 conference/iv2019/manuscript.tex
adding in iv2019 manuscript

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%2345678901234567890123456789012345678901234567890123456789012345678901234567890
%        1         2         3         4         5         6         7         8

\documentclass[letterpaper, 10 pt, conference]{ieeeconf}  % Comment this line out if you need a4paper

%\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4 paper

\IEEEoverridecommandlockouts                              % This command is only needed if 
                                                          % you want to use the \thanks command

\overrideIEEEmargins                                      % Needed to meet printer requirements.

%In case you encounter the following error:
%Error 1010 The PDF file may be corrupt (unable to open PDF file) OR
%Error 1000 An error occurred while parsing a contents stream. Unable to analyze the PDF file.
%This is a known problem with pdfLaTeX conversion filter. The file cannot be opened with acrobat reader
%Please use one of the alternatives below to circumvent this error by uncommenting one or the other
%\pdfobjcompresslevel=0
%\pdfminorversion=4

% See the \addtolength command later in the file to balance the column lengths
% on the last page of the document

% The following packages can be found on http:\\www.ctan.org
\usepackage{graphics} % for pdf, bitmapped graphics files
\usepackage{epsfig} % for postscript graphics files
%\usepackage{mathptmx} % assumes new font selection scheme installed
%\usepackage{times} % assumes new font selection scheme installed
%\usepackage{amsmath} % assumes amsmath package installed
%\usepackage{amssymb}  % assumes amsmath package installed

\title{\LARGE \bf
A Hybrid Control Framework for Navigating Pedestrian Interaction at Mid-Block Intersections}

\author{Nitin R. Kapania$^{1}$, Sarah M Thornton, Francesco Borrelli, J Christian Gerdes% <-this % stops a space
\thanks{*This work was supported by the Hyundai Center of Excellence}% <-this % stops a space
\thanks{$^{1}$Nitin Kapania is with the Dept. of Mechanical Engineering, Stanford University, and is a visiting scholar at UC Berkeley}%
}

\begin{document}

\maketitle
\thispagestyle{empty}
\pagestyle{empty}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}

As autonomous vehicles (AVs) inch closer to reality, a central requirement for acceptance will be earning the trust of humans in everyday driving situations. In particular, the interaction between AVs and pedestrians is of
high importance, as every human is a pedestrian at some point of the day. This paper considers the interaction of a pedestrian and an autonomous vehicle at a mid-block, unsignalized intersection where there is ambiguity
over when the pedestrian should cross and when and how the vehicle should yield. By modeling pedestrian behavior through the concept of gap acceptance, the authors show that a hybrid controller with just four states allows an autonomous vehicle to successfully interact with a pedestrian across a continuous spectrum of possible crosswalk-entry behaviors. The controller is validated in simulation through comparison with a previously published control policy obtained through solution of a POMDP, and experimental results are provided on a Hyundai Genesis vehicle for a virtual pedestrian.

\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

\subsection{Motivation}

While autonomous vehicles have the potential to save thousands of lives every year and create tremendous societal benefits \cite{Fagnant2015}, widespread adoption is unlikely until AVs gain the broad trust of society. Given that every human is a pedestrian at some point during the day, one of the central ways that autonomous vehicles will be evaluated is through their interactions with pedestrians. In fact, recent events with industry leader Waymo highlight that AV-pedestrian interaction has still not been perfected\cite{MattRichtelandConorDougherty2015}. In general, interactions with pedestrians are complex, even for experienced human drivers. Issues such as lack of visibility, improper communication, poorly marked roads, and distraction on both the driver or pedestrian side can lead to accidents and fatalities. From 2015-2016, pedestrian fatalaties increased by 9\% to 5987, representing the highest number since 1990, and also representing 16\% of all automotive fatalities \cite{HighwayTrafficSafetyAdministration2016}. As autonomous vehicles inch closer to widespread adoption, they must have a clear control strategy for pedestrian interaction that can handle a wide variety of pedestrian behaviors while maintaining a reasonable flow of traffic.

\subsection{Prior Art}

The need to further understand pedestrian behavior for autonomous driving has created growing body of interdisciplinary research. One branch of research has focused around modeling pedestrian behavior given various sensor inputs. Keller et al \cite{Keller2014} presented a study on pedestrian path prediction and action classification (e.g. crossing vs. waiting) using Gaussian process dynamical models and trajectory matching from optical data. Several other techniques for pedestrian trajectory prediction have been proposed, including LQR \cite{Batkovic}, set-based reachability analysis \cite{Koschi2018}, and Markov processes \cite{Karasev2016}

Another branch of literature is focused on developing binary predictive models for pedestrian crossing. While typically studied for purposes of road design, this body of literature holds promising insights for AV designers. For example, Schroeder and Rouphail \cite{Schroeder2011} explored factors associated with driver yielding behavior at unsignalized pedestrian crossings. Using logistic regression, the authors found that drivers are more likely to yield to assertive pedestrians who walk briskly in their approach to a crosswalk. Kadali and Perumal \cite{RaghuramKadali2012} studied the ``gap acceptance" behavior of pedestrians at mid-block crosswalks through a video graphic survey, and found that the gap accepted for crossing was explained by factors such as crossing direction, vehicle speed, and pedestrian age. Yannis et al. \cite{Yannis2013} and \cite{Sun2002} also found that gap acceptance was influenced by the size of the oncoming vehicle and the presence of other pedestrians.  Lee and Aty \cite{Lee2005} studied interactions in the form of crashes, and found crashes were linked with higher daily traffic.

Finally, a small but rapidly growing body of literature specifically studies the interaction between pedestrians and autonomous vehicles at crosswalks. An excellent review of these studies was conducted by \cite{Rasouli}. As an example, Rothenbucher et al. \cite{Rothenbucher2016} studied the interaction between pedestrians and driverless vehicles by constructing a car seat costume to disguise a driver. The authors noted that pedestrians overall managed interactions at crosswalks effectively, but later mentioned increased uncertainty about the autonomous vehicles behavior. To improve the issue of trust, several researchers have developed external interfaces to more clearly broadcast the intent of the autonomous vehicle \cite{Matthews}, \cite{Lagstrom2015}.

Several real-world studies of AV-pedestrian interaction note that once a local population learned a vehicle was programmed to be perfectly safe, pedestrians would regularly take advantage of the AV and walk in front of it. \cite{Madigan}. This illustrates an issue with automated driving in that an overly conservative crossing algorithm will often be taken advantage of or cause confusion among pedestrians. Camara et al \cite{Camara2018} attempted to model the natural negotation for priority between a pedestrian and an AV at an intersection using the framework of ``sequential chicken" adopted from game theory. Chen et al. \cite{Chen} notes the need for a tradeoff between passive and aggressive driving behavior, and developed a stochastic model of pedestrian behavior to evaluate proposed AV control policies. Control polices for pedestrian interaction are also developed by \cite{Bandyopadhyay}, who proposes a Mixed-observable Markov Decision Process (MOMDP) to incorporate the intention uncertainty of a pedestrian. A POMDP formulation is also proposed by Thornton \cite{Thornton2018}, who proposes navigation of the pedestrian uncertainty through value-sensitive design.

\subsection{Statement of Contributions}
While there exists a wide body of research on the distinct subjects of pedestrian modeling and AV control, there is a lack of research that combines the two. In other words, significant research has been conducted to understand the likelihood of pedestrian crossing given certain traffic conditions and pedestrian demographics \cite{Schroeder2011} -\cite{Lee2005}, and a small but growing body of literature develops control strategies for pedestrian avoidance \cite{Bandyopadhyay}-\cite{Thornton2018}. However, what is missing is a contribution that explicitly considers tests whether a proposed control strategy is robust to the variety of pedestrian behaviors that have been observed from experimental studies on real roads. What is also missing is analysis of how this controller should behave across multiple traffic scenarios - for example, the navigation problem is different for the pedestrian crossing on the opposing side of traffic, and also depends on which particular lane the vehicle is in.

This paper aims to address this gap by developing a state-machine based control architecture that accounts for several distinct pedestrian modes of behavior at an unsignalized crosswalk. An unsignalized crosswalk is chosen as it is the trickiest crosswalk for a pedestrian and vehicle to navigate, as there is no explicit declaration of whose turn it is cross. Simulations show that the state machine controller is able to handle a continuous spectrum of pedestrian \textit{gap acceptance} behavior, tolerating a range of highly conservative to highly aggressive pedestrians. Moreover, this is possible with just four distinct states. Simulations are also used to compare the closed-loop controller with the controller developed by \cite{Thornton2018}. Finally, experimental results are shown on a real vehicle to demonstrate the feasibility of the proposed state machine controller.

\section{Problem Description and System Requirements}

\subsection{Unsignalized Intersections}

In general, pedestrian crosswalks may be controlled or uncontrolled. In the case of the former, control devices such as stop signs or walk signals guide the interaction of vehicles and pedestrians explicitly. Fig.~\ref{fig:schematic} shows the latter example, in which a pedestrian must select a gap in traffic flow and cross.

\begin{figure}
\centering
\includegraphics[width=3.5in]{figures/example.eps}
\caption{Schematic of pedestrian approaching a stream of traffic at a mid-block intersection.}
\label{fig:schematic}
\end{figure}

In general, right-of-way for uncontrolled intersections is complex. For example, nine states and the District of Columbia require motorists to stop when approaching a pedestrian in an uncontrolled crosswalk. Six states require a motorist to stop when a pedestrian is upon the same half of the roadway or within one lane of the lane that the motorist is traveling upon. Another nineteen states require a motorist to yield when a pedestrian is upon any portion of the roadway, and another 20 states mandate that motorists yield when a pedestrian is upon the same half of the roadway or approaching closely from the opposite side of the roadway \cite{NCSL}.

\subsection{Pedestrian Gap Acceptance}

One of the major factors that determines pedestrian crossing behavior is \textit{gap acceptance}, defined in this paper as:

\begin{align}
\mathrm{gap} = \frac{\mathrm{distance\hspace{1mm}to\hspace{1mm}crosswalk}}{\mathrm{vehicle\hspace{1mm}speed}}
\end{align}

Also known as time to collision (TTC) - the safety gap is a measure of how much time there is before the vehicle would enter the crosswalk if it kept it's current speed constant. When faced with a stream of traffic at a crosswalk, a pedestrian inherently decides how much of a gap to accept before crossing. The average gap acceptance is reported in the literature to be between 3-7 seconds, meaning pedestrians usually do not cross if the vehicle would enter the crosswalk in under three seconds \cite{DiPietroCharlesMandKing1970}, and are very likely to cross when they have more than seven seconds \cite{Schmidt2009}.

\subsection{Controller Objectives}

Given the multiple stakeholders present (the autonomous vehicle and its passengers, the pedestrian(s), and the other vehicles in the traffic stream), it is important to consider the values and needs of all participants when determining the objectives of a control framework. This paper determines controller objectives based on the three-part value sensitive design (VSD) framework proposed by Thornton et al. 
\cite{Thornton2018a}

\section{Proposed Control Architecture}

\section{Evaluation Methodology and Simulation Results}

\section{Experimental Results}

\section{CONCLUSIONS}

\addtolength{\textheight}{-12cm}   % This command serves to balance the column lengths
                                  % on the last page of the document manually. It shortens
                                  % the textheight of the last page by a suitable amount.
                                  % This command does not take effect until the next page
                                  % so it should come on the page before the last. Make
                                  % sure that you do not shorten the textheight too much.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{APPENDIX}

Appendixes should appear before the acknowledgment.

\section*{ACKNOWLEDGMENT}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliographystyle{IEEEtran}
\bibliography{Bibliography}

\end{document}

### 2019-01-09 unpublished/pedestrian/pedestrianInteraction.tex
adding in iv2019 manuscript

@@
 \textbf{Nitin Kapania} \\
-\textbf{Modeling Pedestrian - AV Interaction at a Mid-Block Intersection} \\
+\textbf{Modeling the Interaction Between a Pedestrian and Autonomous Vehicle at a Mid-Block Intersection} \\
 
 \section{Problem Formulation}
-Consider the problem of an autonomous vehicle attempting to drive past an unsignalized, ``mid-block" crosswalk. A pedestrian is simultaneously walking towards the intersection.
+Consider the problem of an autonomous vehicle attempting to drive past an unsignalized, ``mid-block" crosswalk, shown in Fig.~\ref{img:example}.
+A pedestrian is simultaneously walking towards the intersection, but has not entered.
 
@@
 
-In this case, what should the autonomous vehicle do?
+What should the autonomous vehicle do?
 Should the vehicle always stop as soon as a pedestrian is observed near a cross-walk?
@@
 While several states (e.g.
-Minnesota, Georgia, Maryland) have "must-stop" rules for uncontrolled intersections, other states have ``must yield" rules.
+Minnesota, Georgia, Maryland) have ``must-stop" rules for uncontrolled intersections, other states have ``must yield" rules.
 For example, vehicles in Arkansas must yield once a pedestrian is in any portion of the roadway, while vehicles in Alabama require motorists to yield to a pedestrian either on the same half of the road or approaching from the opposite side of the roadway.
@@
 	interaction scenarios?
-	\item Can we design control policies for the vehicle that maintain a high standard of safety while also enabling robust traffic flow 
+	\item Can we design control policies for the vehicle that maintain a high standard of safety while also enabling robust traffic flow? 
 \end{itemize}
@@
 
-There is a wide body of literature segmenting the work on pedestrian behavior.
-In terms of purely modeling the dynamics of pedestrian motion, Batkovic et al.
-\cite{Batkovic} presented a mathematical model to predict pedestrian motion over a finite horizon, with a focus on low computational complexity for autonomous driving.
+There is a wide body of literature focused on understanding pedestrian behavior.
+In terms of microscopically modeling the dynamics of pedestrian motion, Batkovic et al.
+\cite{Batkovic} presented a mathematical model to predict pedestrian motion over a finite horizon based on a map of the road configuration, with a focus on low computational complexity for autonomous driving.
 Karasev et al.
-modeled pedestrian behavior as intent using a Markov Decision Process \cite{Karasev2016}.
+modeled pedestrian intent using a Markov Decision Process \cite{Karasev2016}.
 
-Another segment of the literature is focused on developing predictive models for pedestrian crossing.
-These approaches typically study the parallel issues of pedestrian crossing behavior or driver yielding behavior and date as early as 1975 \cite{Katz1975}.
+Another segment of the literature is focused on developing macroscopic predictive models for pedestrian crossing.
+These approaches typically study the parallel issues of pedestrian crossing behavior or driver yielding behavior based on experimental observation and date as early as 1975 \cite{Katz1975}.
 Schroeder and Rouphail \cite{Schroeder2011} explored factors associated with driver yielding behavior at unsignalized pedestrian crossings.
 Using logistic regression, the authors found that drivers are more likely to yield to assertive pedestrians who walk briskly in their approach to a crosswalk.
-Kadali and Perumal \cite{RaghuramKadali2012} studied the ``gap acceptance" behavior of pedestrians at mid-block crosswalks through a video graphic survey, and found that the gap accepted by crossing was explained by factors such as crossing direction, vehicle speed, and pedestrian age.
+Kadali and Perumal \cite{RaghuramKadali2012} studied the ``gap acceptance" behavior of pedestrians at mid-block crosswalks through a video graphic survey, and found that the gap accepted for crossing was explained by factors such as crossing direction, vehicle speed, and pedestrian age.
 Yannis et al.
-\cite{Yannis2013} also found that gap acceptance was influenced by the size of the oncoming vehicle and the presence of other pedestrians.
-Gap acceptance models were also proposed by \cite{Sun2002}.
+\cite{Yannis2013} and \cite{Sun2002} also found that gap acceptance was influenced by the size of the oncoming vehicle and the presence of other pedestrians.
 Lee and Aty \cite{Lee2005} studied interactions in the form of crashes, and found crashes were linked with higher daily traffic.
 
-Another branch of literature specifically studies the interaction between pedestrians and autonomous vehicles.
-A review of these studies was conducted by \cite{Rasouli}.
-Rothenbucher et al.
+Another branch of literature specifically studies the interaction between pedestrians and autonomous vehicles at crosswalks.
+An excellent review of these studies was conducted by \cite{Rasouli}.
+As an example, Rothenbucher et al.
 \cite{Rothenbucher2016} studied the interaction between pedestrians and driverless vehicles by constructing a car seat costume to disguise a driver.
-The authors noted that pedestrians overall managed interactions at crosswalks effectively, but later mentioned increased uncertainty about the car's behavior.
+The authors noted that pedestrians overall managed interactions at crosswalks effectively, but later mentioned increased uncertainty about the autonomous vehicles behavior.
 To improve the issue of trust, several researchers have developed external interfaces to more clearly broadcast the intent of the autonomous vehicle \cite{Matthews}, \cite{Lagstrom2015}.
@@
 This illustrates an issue with automated driving in that an overly conservative crossing algorithm will often be taken advantage of or cause confusion among pedestrians.
-Camara et al \cite{Camara2018} attempted to model the natural negotation for priority between a pedestrian and an AV at an intersection using game theory.
+Camara et al \cite{Camara2018} attempted to model the natural negotation for priority between a pedestrian and an AV at an intersection using the framework of ``sequential chicken" adopted from game theory.
+Chen et al.
+\cite{Chen} notes the need for a tradeoff between passive and aggressive driving behavior, and developed a stochastic model of pedestrian behavior to evaluate proposed AV control policies.
+Policies for pedestrian interaction are also developed by \cite{Bandyopadhyay}, who proposes a Mixed-observable Markov Decision Process (MOMDP) to incorporate the intention uncertainty of a pedestrian.
+A POMDP formulation is also proposed by Thornton \cite{Goldenstein2013}, who proposes navigation of the pedestrian uncertainty through value-sensitive design.
 
 \section{Proposed Research}
+
+I have a couple of ideas to get into this field, but could use more inspiration
+
+\begin{itemize}
+	\item Collecting a dataset of ambiguous pedestrian-vehicle interaction exemplars from the VAIL simulator. 
+	\subitem
+		It is likely that we can use our simulator capabilities to set up an environment where two humans can negotiate a tricky mid-block intersection. The 
+		initial settings can be set up so it is ambiguous whether the car should yield or go forward
+
+	\subitem
+		One human can be operating a vehicle, while another human can operate as a pedestrian in the world using a joystick.
+
+	\item Showing the inherent uncertainty of the problem by simulating a pedestrian and vehicle acting independently at a mid-block intersection. Given a traffic law 
+	for a certain state, we can show that given uncertainty, the traffic flow is inherently suboptimal.
+
+	\subitem 
+		However, given some communication of intent (either through a signaling device on the AV) or through assertive or intent-giving motion from the AV, we can optimize traffic flow while avoiding collision probability.
+
+\end{itemize}
 

### 2019-01-16 conference/iv2019/manuscript.tex
chugging along through IV2019 paper

@@
 \usepackage{epsfig} % for postscript graphics files
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+%\usepackage{align}
 %\usepackage{mathptmx} % assumes new font selection scheme installed
@@
 \title{\LARGE \bf
-A Hybrid Control Framework for Navigating Pedestrian Interaction at Mid-Block Intersections}
+A Hybrid Control Framework for Autonomous Vehicles at Uncontrolled Intersections}
 
-\author{Nitin R. Kapania$^{1}$, Sarah M Thornton, Francesco Borrelli, J Christian Gerdes% <-this % stops a space
+\author{Nitin R. Kapania$^{1}$, Francesco Borrelli, J Christian Gerdes% <-this % stops a space
 \thanks{*This work was supported by the Hyundai Center of Excellence}% <-this % stops a space
@@
 high importance, as every human is a pedestrian at some point of the day.
-This paper considers the interaction of a pedestrian and an autonomous vehicle at a mid-block, unsignalized intersection where there is ambiguity
+This paper considers the interaction of a pedestrian and an autonomous vehicle at a mid-block, uncontrolled intersection where there is ambiguity
 over when the pedestrian should cross and when and how the vehicle should yield.
-By modeling pedestrian behavior through the concept of gap acceptance, the authors show that a hybrid controller with just four states allows an autonomous vehicle to successfully interact with a pedestrian across a continuous spectrum of possible crosswalk-entry behaviors.
+By modeling pedestrian behavior through the concept of gap acceptance, the authors show that a hybrid controller with just four distinct modes allows an autonomous vehicle to successfully interact with a pedestrian across a continuous spectrum of possible crosswalk-entry behaviors.
 The controller is validated in simulation through comparison with a previously published control policy obtained through solution of a POMDP, and experimental results are provided on a Hyundai Genesis vehicle for a virtual pedestrian.
@@
 
-The need to further understand pedestrian behavior for autonomous driving has created growing body of interdisciplinary research.
+The need to further understand pedestrian behavior for autonomous driving has created a growing body of interdisciplinary research.
 One branch of research has focused around modeling pedestrian behavior given various sensor inputs.
@@
 
-Another branch of literature is focused on developing binary predictive models for pedestrian crossing.
+Another branch of literature is focused on developing predictive models for pedestrian crossing.
 While typically studied for purposes of road design, this body of literature holds promising insights for AV designers.
@@
 \subsection{Statement of Contributions}
-While there exists a wide body of research on the distinct subjects of pedestrian modeling and AV control, there is a lack of research that combines the two.
-In other words, significant research has been conducted to understand the likelihood of pedestrian crossing given certain traffic conditions and pedestrian demographics \cite{Schroeder2011} -\cite{Lee2005}, and a small but growing body of literature develops control strategies for pedestrian avoidance \cite{Bandyopadhyay}-\cite{Thornton2018}.
-However, what is missing is a contribution that explicitly considers tests whether a proposed control strategy is robust to the variety of pedestrian behaviors that have been observed from experimental studies on real roads.
+Significant research has been conducted to understand the likelihood of pedestrian crossing given certain traffic conditions and pedestrian demographics \cite{Schroeder2011} -\cite{Lee2005}, and a small but growing body of literature develops control strategies for pedestrian avoidance \cite{Bandyopadhyay}-\cite{Thornton2018}.
+However, what is missing is a contribution that explicitly  tests whether a proposed control strategy is robust to the variety of pedestrian behaviors that have been observed from experimental studies on real roads.
 What is also missing is analysis of how this controller should behave across multiple traffic scenarios - for example, the navigation problem is different for the pedestrian crossing on the opposing side of traffic, and also depends on which particular lane the vehicle is in.
 
-This paper aims to address this gap by developing a state-machine based control architecture that accounts for several distinct pedestrian modes of behavior at an unsignalized crosswalk.
+This paper aims to address this gap by developing a hybrid control architecture that accounts for several distinct pedestrian modes of behavior at an unsignalized crosswalk.
 An unsignalized crosswalk is chosen as it is the trickiest crosswalk for a pedestrian and vehicle to navigate, as there is no explicit declaration of whose turn it is cross.
@@
 
-\section{Problem Description and System Requirements}
+\section{Problem Overview}
 
@@
 
-One of the major factors that determines pedestrian crossing behavior is \textit{gap acceptance}, defined in this paper as:
+One of the major factors that determines pedestrian crossing behavior at uncontrolled intersections is \textit{gap acceptance}, defined in this paper as:
 
-\begin{align}
+\begin{equation}
 \mathrm{gap} = \frac{\mathrm{distance\hspace{1mm}to\hspace{1mm}crosswalk}}{\mathrm{vehicle\hspace{1mm}speed}}
-\end{align}
+\end{equation}
 
-Also known as time to collision (TTC) - the safety gap is a measure of how much time there is before the vehicle would enter the crosswalk if it kept it's current speed constant.
+Similar to time to collision (TTC), the safety gap is a measure of how much time there is before the vehicle would enter the crosswalk if it kept it's current speed constant.
 When faced with a stream of traffic at a crosswalk, a pedestrian inherently decides how much of a gap to accept before crossing.
@@
 
-\subsection{Controller Objectives}
+\subsection{Problem Formalization and System Requirements}
 
-Given the multiple stakeholders present (the autonomous vehicle and its passengers, the pedestrian(s), and the other vehicles in the traffic stream), it is important to consider the values and needs of all participants when determining the objectives of a control framework.
-This paper determines controller objectives based on the three-part value sensitive design (VSD) framework proposed by Thornton et al.
-\cite{Thornton2018a}
+A diagram of the relevant state variables is shown below in Fig.~\ref{fig:diagram1}.
+We assume that knowledge of distance $d$ to a fixed stopping point 3-5 meters ahead of the crosswalk can be estimated by a perception system, and that the velocity of the vehicle $\dot{d}$ is known as well.
+Furthermore, we also assume the perception system is able to determine the position of the pedestrian $x_p$ within the crosswalk, and is able to determine a velocity estimate $\dot{x_p}$ for the pedestrian as well.
+
+\begin{figure}
+\centering
+\includegraphics[width=1in]{figures/diagram.eps}
+\caption{Description of relevant problem variables.}
+\label{fig:diagram1}
+\end{figure}
+
+The relevant control problem is therefore to determine an appropriate longitudinal acceleration command for the vehicle, $u = \ddot{d}$, as a closed-loop function of the vehicle and pedestrian states:
+
+\begin{equation}
+\ddot{d} = f(d, \dot{d}, x_p, \dot{x_p})
+\end{equation}
+
+In general, this control problem is difficult as there are multiple stakeholders and therefore multiple competing objectives that must be considered.
+For example, relevant stakeholders for this problem are the autonomous vehicle and its passengers, the pedestrian(s), and the other vehicles in the traffic stream.
+
+As a result, an algorithm that simply yields every time a pedestrian is near the crosswalk is likely to be too conservative, as there is a likelihood the pedestrian is waiting for a larger gap in the traffic stream to cross or is walking through the sidewalk without intending to cross in the first place.
+To formalize the system requirements for this multi-stake holder problem, we propose the following \textit{design constraints}, extended from the engineering specifications proposed by Thornton \cite{Thornton2018}.
+
+\begin{center}
+\begin{tabular}{ l| l }
+\label{tb:specs}
+ \textbf{Engineering} & \textbf{Design} \\
+ \textbf{Specification} & \textbf{Constraint}\\\hline 
+ 1. Safety and Legality & a. Avoid collisions for all\\
+                     & possible accepted gaps                        \\
+                     &b. Ability to follow stop/yield    \\
+                     &laws for all 50 U.S. states\\
+
+&\\
+ 2. Efficiency & a. Average speed: Minimize  \\ 
+               & deviation from speed of traffic \\    
+               & b. Wait for pedestrian crossing \\
+               & event to materialize before yielding. \\
+
+               & \\
+ 3. Smoothness & a. Low accel: Nominal brake / \\ 
+ 			   & accel limits of 2 m/s/s \\
+ 			   & b. Stop within 3-5 meters\\
+ 			   & of crosswalk
+
+\end{tabular}
+\end{center}
 
@@
 
+The proposed control architecture is shown below in Fig.~\ref{fig:hybridController}, and the algorithm for operation is shown in Algorithm \ref{alg:hybridController}.
+At heart, the control algorithm uses a feedback-feedforward methodology to compute desired acceleration commands $\ddot{d}$, but computes these commands differently depending on which of four discrete states the controller is in.
+The next sections will describe the four states.
+
+\begin{figure}
+\centering
+\includegraphics[width=3.5in]{figures/stateDiagram.eps}
+\caption{Schematic of hybrid control structure.}
+\label{fig:hybridController}
+\end{figure}
+
+\begin{algorithm} % enter the algorithm environment
+\caption{Calculate $\ddot{d} = f(d, \dot{d}, x_p, \dot{x}_p)$} % give the algorithm a caption
+\label{alg:hybridController} % and a label for \ref{} commands later in the document
+\begin{algorithmic}
+\State $state \leftarrow DRIVING$
+\While {$\mathbf{true}$}
+\If {$state = DRIVING$}
+	\State $\ddot{d} \gets k_s(\dot{d}-v_\mathrm{speed limit})$ 
+	\If{$d > 0 \hspace{1mm} \mathbf{and} \hspace{1mm} \mathrm{inCrosswalk}(x_p)$}
+		\If{$\mathrm{timeAdvantage}(x_p, \dot{x_p}, d, \dot{d}) > t_{max}$} 
+		\State $state \leftarrow DRIVING$
+
+		\ElsIf{$d > \frac{\dot{d}^2}{2a_\mathrm{cmfrt}}$}
+		\State $state \leftarrow YIELDING$
+
+	\ElsIf{$d < \frac{\dot{d}^2}{2a_\mathrm{cmfrt}} \hspace{1mm} \mathbf{and} \hspace{1mm} d > \frac{\dot{d}^2}{2a_\mathrm{max}}$}
+	\State $state \leftarrow HARD\hspace{1mm}BRAKING$
+
+	\Else
+	\State $state \leftarrow SPEED\hspace{1mm}UP$
+
+	\EndIf
+	\EndIf
+	\EndIf
+\vspace{3mm}
+\If {$state = YIELDING$}
+\If {$d > \frac{\dot{d}^2}{2a_\mathrm{cmfrt}} + t_{delay}\dot{d}$}
+	\State $\ddot{d}_{des} = 0$
+	\State $\dot{d}_{des} = v_\mathrm{speed limit}$
+\Else 
+	\State $\ddot{d}_{des} = -a_\mathrm{cmf}$
+	\State $\dot{d}_{des} = \mathrm{getDesiredSpeed}(d, \dot{d})$
+\EndIf 
+	\State $\ddot{d} = \ddot{d}_{des} + k_s(\dot{d} - \dot{d}_{des})$
+
+\If {$\textbf{not} \hspace{1mm} \mathrm{inCrosswalk}(x_p)$} \State $state \leftarrow DRIVING$
+
+\EndIf
+\EndIf
+
+\vspace{3mm}
+
+\If {$state = HARD\hspace{1 mm}BRAKING$}
+	\State $\dot{d}_{des} = \mathrm{getDesiredSpeed}(d, \dot{d})$
+	\State $\ddot{d} = -\frac{\dot{d}^2}{2d} + k_s(\dot{d} - \dot{d}_{des})$ 	
+	\If {$\textbf{not} \hspace{1 mm} \mathrm{inCrosswalk}(x_p)$} \State $state \leftarrow DRIVING$
+
+	\EndIf
+	\EndIf
+
+\EndIf
+\If {$state = SPEED\hspace{1mm}UP$}
+	\State $\ddot{d} = a_\mathrm{cmf}$ 	
+	\If {$\textbf{not} \hspace{1mm} \mathrm{inCrosswalk}(x_p)$} \State $state \leftarrow DRIVING$
+
+	\EndIf
+	\EndIf
+
+\EndWhile
+
+\end{algorithmic}
+\end{algorithm}
+
+\subsection{Nominal Driving}
+In the nominal state, the vehicle attempts to drive through the crosswalk at the speed limit $v_\mathrm{speedlimit}$.
+A simple proportional speed control is applied to keep the vehicle at the speed limit, achieving design constraint 2.a.:
+
+\begin{equation}
+	\ddot{d} = k_s(\dot{d} - v_\mathrm{speedlimit})
+\end{equation}
+
+The controller remains in the driving mode unless the pedestrian begins to enter the crosswalk while $d > 0$.
+To allow more time for the controller to respond, we allow the $\mathrm{inCrosswalk}(x_p, \dot{x}_p)$ indicator function to be triggered the moment the pedestrian approaches the crosswalk from the sidewalk - e.g.
+once the pedestrian velocity $\dot{x}_p$ becomes non-zero.
+
+ Note that according to design constraint 1.b, the definition of entering the crosswalk will vary according to which of the fifty US states the vehicle is in.
+For example, in California, the pedestrian entering any portion of the crosswalk should trigger the control mode to change - while in Louisiana, the pedestrian entering the same half of the roadway as the vehicle should trigger the change \cite{NCSL2018}.
+
+Once the pedestrian enters the crosswalk, the algorithm decides which mode to enter next.
+
+\subsubsection{Time Advantage}
+
+Frequently he vehicle will pass through the intersection well before the pedestrian will reach the vehicle's lane.
+Assuming the vehicle is operating in a US state where stopping is not explicitly required, design constraint 2.a suggests it is desirable for the vehicle to continue through the intersection.
+
+This can be formalized by calculating the \textit{time advantage} $t_adv$, also known as the \textit{time to collision} as follows:
+
+\begin{equation}
+t_{adv} = \frac{d}{\dot{d}} - \frac{x_v - x_p}{\dot{x_p}}
+\end{equation}
+
+Where $x_v$ is the $x$ position of the vehicle in the crosswalk.
+The control algorithm in this work will explicitly allow the vehicle to continue driving through the intersection as long as $t_{adv}$ exceeds a specfiied threshold $t_{max}$.
+
+\subsubsection{Yielding}
+
+If the time advantage is not sufficient for the vehicle to pass through, the autonomous vehicle must slow down.
+The preferred option to meet design constraint 3.a is to brake at a low, comfortable deceleration $-a_{cmf}$ for the passenger.
+Note that design constraint 3.b is satisfied by having $d=0$ be defined to be a point several meters ahead of the crosswalk.
+
+Comfortable braking at a deceleration $-a_{cmf}$ is possible if sufficient braking distance exists. The braking distance is derived kinematically as
+
+\begin{equation}
+d_{brake} = \frac{\dot{d}^2}{2a_{cmf}}
+\end{equation}
+
+If this holds true when the pedestrian starts to enter the crosswalk, the car enters a $Yielding$ state.
+
+\subsubsection{Hard Braking}
+
+In the unlikely event a pedestrian begins to cross when the time gap is low (e.g.
+under 3 seconds), the vehicle will need to prioritize design constraint 1.a and come to a stop at a deceleration higher than $-a_{cmf}$.
+This occurs when the following conditions holds:
+
+\begin{equation}
+\frac{\dot{d}}{2a_{max}} < d < \frac{\dot{d}^2}{2a_{cmf}}
+\end{equation}
+
+Where $a_{max}$ is the largest deceleration magnitude allowed by the tire-road friction.
+
+\subsection{Speeding Up}
+
+As a final condition, consider the case where the pedestrian enters the crosswalk just as the vehicle is crossing $d = 0$ - in this case, it makes little sense to decelerate, as the vehicle has insufficient space to stop before the crosswalk and will risk being rear-ended (Design Constraint 2.a) or stopping in the crosswalk.
+It makes more sense in this condition for the vehicle to speed up and exit the crosswalk quickly.
+As a result, if $d < \frac{\dot{d}}{2a_{max}}$, the vehicle speeds up.
+
+\subsection{Yielding}
+
+In the yielding state, the vehicle follows a different version of the feedback-feedforward dynamics.
+The vehicle first follows the speed limit until it reaches the critical value of $\frac{\dot{d}^2}{2a_{cmf}}$.
+In practice, an additional term $t_delay\dot{d}$ is added to compensate for the delay in brake communication.
+
+At the critical istance, the vehicle decelerates at the uniform yielding deceleration of $-a_{cmf}$, with an additional feedback term $k_s(\dot{d} - \dot{d}_des$), where $\dot{d}_des$ is the desired velocity.
+The feedback term helps bring the vehicle speed to 0 at the desired stopping point.
+
+The desired velocity profile $\dot{d}(d)$ for a vehicle decelerating at constant acceleration is given via integration:
+
+\begin{equation}
+\dot{d}(d) = \sqrt{2a_{cmf}(d-d_o) + \dot{d}_o^2}
+\end{equation}
+
+Where $d_o$ and $\dot{d}_o$ is the value of $d$ and $\dot{d}$ when the controller first enters the yielding state.
+Simple inspection shows that the desired velocity is $\dot{d}_o$ when $d = d_o$ and is 0 when $d = 0$ (recall that we enter the yielding state when $d_o = \frac{\dot{d_o}^2}{2a_{cmf}}$.
+
+The controller exits the yielding mode and returns to the driving mode once the pedestrian clears the crosswalk.
+
+\subsection{Hard Braking}
+
+In the hard braking state, the vehicle follows a similar feedback-feedforward algorithm, but the desired feedforward deceleration in this case is given by $\ddot{d} = -\frac{d^2}{2d}$.
+Integrating to find the desired speed profile $\dot{d}(d)$ yields the following deceleration:
+
+\begin{equation}
+\dot{d}(d) = \frac{\dot{d}_o}{\sqrt{d_o}}\sqrt{d}
+\end{equation}
+
+Where $d_o$ and $\dot{d}_o$ is the value of $d$ and $\dot{d}$ when the controller first enters the braking state.
+Again, the controller exits the braking mode and returns to the driving mode once the pedestrian clears the crosswalk.
+
+\subsection{Speed Up}
+
+In the speed up state, there is no need to follow an exact speed profile as the vehicle is merely trying to exit the crosswalk area slightly faster.
+In this case, the commanded acceleration $\ddot{d} = a_{cmf}$ until the pedestrian clears the crosswalk.
+
 \section{Evaluation Methodology and Simulation Results}
 
+\subsection{Evaluation Methodology}
+
+In order to check performance of the algorithm, both simulation and experimental studies were undertaken.
+The simulation results focused on testing Algorithm 1 against the design requirements from Table \ref{tb:specs} over many pedestrian crossings.
+
+The algorithm is validated in simulation via a built-from-scratch crosswalk simulation environment (Fig.
+\ref{fig:simFramework}) developed in Python.
+The simulation environment is built on the ROS \cite{ROS} robotics interface to enable easy porting of the control architecture to the experimental vehicle in Section \ref{sec:expres}.
+The simulation includes features such as multiple lanes, the pedestrian crossing on the right or the left, and animation capabilities.
+
+\begin{figure}
+\centering
+\includegraphics[width=3.5in]{figures/simFramework.png}
+\caption{Crosswalk simulation environment for algorithm validation.}
+\label{fig:simFramework}
+\end{figure}
+
+To simulate random pedestrian crossing events, a pedestrian is spawned in an area near the crosswalk.
+The pedestrian accepts a randomly sized time gap in the traffic stream to cross the road at constant speed $\dot{x}_p$.
+The size of the time gap is drawn from a normal distribution with mean $\mu_{gap}$ and standard deviation $\sigma_{gap}$.
+Statistics for $\mu_{gap}$ and $\sigma_{gap}$ are obtained from a video graphic analysis by Feliciani et al.
+\cite{Feliciani2017}.
+
+The ego vehicle's response to the pedestrian is simulated over Algorithm 1 for the parameters shown in Table \ref{tb:params}.
+
+\begin{table}[h]
+\footnotesize
+\begin{center}
+\caption{Simulation and Controller Parameters}\label{tb:params}
+\begin{tabular}{lccc}
+Parameter & Symbol & Value & Units \\\hline\hline
+Number of lanes & $n$ & 4 & - \\
+Gap variance & $\sigma_{gap}$ & 2.5 & sec\\
+Mean accepted gap & $\mu_{gap}$ & 4.0 & sec\\\hline
+Speed feedback gain & $k_s$ & 2.0 & 1 / sec\\
+Brake time delay    & $t_\mathrm{delay}$ & 0 & sec \\ 
+Speed limit & $v_\mathrm{speedlimit}$ & 4.5 & m/s \\
+Comfort acceleration & $a_\mathrm{cmf}$ & 2 & $\mathrm{m/s^2}$ \\
+Maximum acceleration & $a_\mathrm{max}$ & 9 & $\mathrm{m/s^2}$ \\\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+\subsection{Simulation Results}
+
+Figure \ref{fig:velPlot1} shows simulation results for the case where the pedestrian enters the crosswalk on the same side of the road as the vehicle.
+The left side of the figure is for the case where the vehicle is on the right-most lane of a four way road, while the right side of the figure is for the case where the vehicle is in the second to right lane.
+
+\begin{figure}
+\centering
+\includegraphics[width=3.5in]{figures/velPlot.eps}
+\caption{Simulation results for case where pedestrian enters the crosswalk on the same side of the road as the vehicle.}
+\label{fig:velPlot1}
+\end{figure}
+
+Figure \ref{fig:velPlot1} shows simulation results for the case where the pedestrian enters the crosswalk on the same side of the road as the vehicle.
+The left side of the figure is for the case where the vehicle is on the right-most lane of a four way road, while the right side of the figure is for the case where the vehicle is in the second to right lane.
+
 \section{Experimental Results}
+\label{sec:expres}
 

### 2019-01-16 conference/iv2019/plotScripts/velocityPlot.m

@@ -1,0 +1,129 @@
+clear all; close all; clc;
+dat1 =load('/home/nkapania/catkin_ws/src/harpagornis/bags/out1.mat');
+dat1.lane = 3;
+
+dat2 =load('/home/nkapania/catkin_ws/src/harpagornis/bags/out2.mat');
+dat2.lane = 2;
+
+laneWidth = 3.7;
+width = 1.5;
+
+dat1.vehicleX = dat1.lane * laneWidth + laneWidth /2 - width / 2;
+dat2.vehicleX = dat2.lane * laneWidth + laneWidth /2 - width / 2;
+
+pedestrianY = 28.0;
+
+%%
+N1 = length(dat1.inCrosswalk_bool);
+N2 = length(dat2.inCrosswalk_bool);
+
+dat1.pedXY = zeros(N1,2);
+dat1.vehXY = zeros(N1,2);
+dat1.distXY = zeros(N1,1);
+
+for i = 1:N1
+    dat1.pedXY(i,:) = [dat1.xP(i) pedestrianY];
+    dat1.vehXY(i,:) = [dat1.vehicleX dat1.xV(i)];
+    dat1.distXY(i) = norm(dat1.pedXY(i,:) - dat1.vehXY(i,:));
+end
+
+
+dat2.pedXY = zeros(N2,2);
+dat2.vehXY = zeros(N2,2);
+dat2.distXY = zeros(N2,1);
+
+for i = 1:N2
+    dat2.pedXY(i,:)  = [dat2.xP(i) pedestrianY];
+    dat2.vehXY(i,:)  = [dat2.vehicleX dat2.xV(i)];
+    dat2.distXY(i) = norm(dat2.pedXY(i,:) - dat2.vehXY(i,:));
+end
+
+%%
+
+dat1.numTrials = max(dat1.trialNumber);
+dat1.avgVelocity = zeros(dat1.numTrials,1);
+dat1.maxAccel    = zeros(dat1.numTrials,1); 
+dat1.minDistance = zeros(dat1.numTrials,1);
+dat1.acceptedGapDownsample = zeros(dat1.numTrials,1);
+
+
+for i = 1:dat1.numTrials
+    ind = find(dat1.trialNumber == i);
+    dat1.avgVelocity(i) = mean(dat1.velocity_mps(ind)); 
+    dat1.maxAccel(i) = max(abs(dat1.ddxV(ind)));
+    dat1.acceptedGapDownsample(i) = mean(dat1.acceptedGap(ind));
[…]

### 2019-01-16 conference/iv2019/manuscript.tex
wrapping up simulation section

@@
 \subsection{Problem Formalization and System Requirements}
+\label{sec:probform}
 
@@
 \centering
-\includegraphics[width=3.5in]{figures/simFramework.png}
-\caption{Crosswalk simulation environment for algorithm validation.}
+\includegraphics[width=2.5in]{figures/simFramework.png}
+\caption{Crosswalk simulation environment for algorithm validation. Ego vehicle is shown in black, following vehicle in traffic stream shown in blue, and pedestrian shown in red.}
 \label{fig:simFramework}
@@
 Comfort acceleration & $a_\mathrm{cmf}$ & 2 & $\mathrm{m/s^2}$ \\
+Maximum time advantage & $t_\mathrm{max}$ & 4 & sec \\
 Maximum acceleration & $a_\mathrm{max}$ & 9 & $\mathrm{m/s^2}$ \\\hline
@@
 The left side of the figure is for the case where the vehicle is on the right-most lane of a four way road, while the right side of the figure is for the case where the vehicle is in the second to right lane.
+Every marker in the figure represents a simulated vehicle crossing.
 
@@
 
-Figure \ref{fig:velPlot1} shows simulation results for the case where the pedestrian enters the crosswalk on the same side of the road as the vehicle.
-The left side of the figure is for the case where the vehicle is on the right-most lane of a four way road, while the right side of the figure is for the case where the vehicle is in the second to right lane.
+Figure \ref{fig:velPlot2} shows the same simulation results, but for the case where the pedestrian enters the crosswalk on the opposite side of the road as the vehicle.
+Again, the total number of pedestrian crossing trials conducted was 750, with 375 trials for each side.
+
+\begin{figure}
+\centering
+\includegraphics[width=3.5in]{figures/velPlot2.eps}
+\caption{Simulation results for case where pedestrian enters the crosswalk on the opposite side of the road as the vehicle.}
+\label{fig:velPlot2}
+\end{figure}
+
+\subsubsection{Safety and Legality}
+
+To meet the collision avoidance design requirement, we generated enough simulations to ensure errant pedestrian crossing behavior at "risky" accepted gaps (< 3 seconds).
+The minimum distance between the vehicle and the pedestrian for each trial is shown in the bottom rows of Fig.
+\ref{fig:velPlot1} and Fig.\ref{fig:velPlot2}.
+In general, the vehicle is able to maintain a safe distance of at least four meters from the pedestrian for the case where the vehicle is in the second to right lane.
+This is because the algorithm has enough time to react to a pedestrian crossing no matter what size gap is accepted.
+
+The riskiest case occurs when the vehicle is in the right-most lane and the pedestrian crosses from the right side, accepting a gap between 1.25 and 1.75 seconds.
+Note that this is in general unlikely given gap acceptance statistics reported in the literature \cite{Feliciani2017}\cite{Rasouli}, but would represent a safety risk given the vehicle has limited time to react.
+However, since the algorithm recognizes that braking will not clear the crosswalk, the vehicle is able to speed up and maintain at least 2 meters of lateral distance from the pedestrian at all times.
+Note that a better way to confirm robustness of the controller to all reasonable pedestrian behaviors would be through a verification analysis - this will be the focus of future work.
+
+In terms of the fifty state legality requirement, we conducted our simulation for the twenty states where the vehicle is required to yield, not stop, to a pedestrian in the same side of the crosswalk or approaching from the opposite side.
+For the US states where the vehicle is explicitly required to stop, the change to the controller is trivial - the controller may not stay in the driving mode if sufficient time advantage exits (see Algorithm 1).
+
+\subsubsection{Efficiency}
+
+The requirement to minimize deviation from speed of traffic is validated in the top row of Fig.\ref{fig:velPlot1} and Fig.\ref{fig:velPlot2}, which shows the average velocity of the vehicle for each simulated crossing.
+For high accepted gaps, the pedestrian simply waits for the ego vehicle to exit the crosswalk before beginning to walk, and the algorithm does not need to slow down from the speed of traffic.
+For low accepted gaps, the vehicle cannot stop, and must speed up or continue driving through the intersection, also keeping the speed of traffic high.
+
+The most disruption to traffic flow occurs when the vehicle must come to a complete stop, which occurs for accepted gaps between 2-6 seconds.
+In this case, the average velocity increases as the accepted gap increases.
+This is because as the pedestrian enters the crosswalk earlier, they are closer to exiting when the vehicle must yield, resulting in less time stopped and higher average velocity.
+
+Note that a special case occurs in Fig.\ref{fig:velPlot2} where the vehicle is in the right-most lane of the four lane road and the pedestrian crosses from the other side.
+Because the vehicle is at a far distance from the pedestrian, the time advantage remains sufficiently high for the vehicle to comfortably continue through the crosswalk without needing to slow down.
+
+\subsubsection{Smoothness}
 

### 2019-01-16 conference/iv2019/plotScripts/velocityPlot.m

@@ -1,8 +1,8 @@
 clear all; close all; clc;
-dat1 =load('/home/nkapania/catkin_ws/src/harpagornis/bags/out1.mat');
+dat1 =load('/home/nkapania/catkin_ws/src/harpagornis/bags/out4.mat');
 dat1.lane = 3;
 
-dat2 =load('/home/nkapania/catkin_ws/src/harpagornis/bags/out2.mat');
+dat2 =load('/home/nkapania/catkin_ws/src/harpagornis/bags/out3.mat');
 dat2.lane = 2;
 
 laneWidth = 3.7;
@@ -108,7 +108,7 @@ ylabel('Maximum Acceleration (m/s2)')
 subplot(3,2,5)
 grid on; hold on; 
 scatter(dat1.acceptedGapDownsample, dat1.minDistance,'k*')
-ylim([0, 9]);
+ylim([0, 13]);
 xlim([0, 9]);
 xlabel('Pedestrian Accepted Gap (s)')
 ylabel('Minimum Distance (m)')

### 2019-01-17 conference/iv2019/manuscript.tex
done with experimental section

@@
 
+Table \ref{tb:speeds} shows a summary of the average speed compared to the baseline speed of 4.4 m/s.
+
+\begin{table}[h]
+\footnotesize
+\begin{center}
+\caption{Average Vehicle Speed}\label{tb:speeds}
+\begin{tabular}{cccc}
+Pedestrian Entry & Vehicle Lane & Average Speed (m/s) & Pct. of Baseline \\\hline\hline
+
+Right Side & Right-Most & 2.90 & 66\% \\
+Right Side & Second-to-Right & 2.93 & 67\% \\
+Left Side  & Right-Most & 4.4 & 100\% \\
+Left Side  & Second-to-Right & 2.80 & 64\% \\\hline
+
+\end{tabular}
+\end{center}
+\end{table}
+
 \subsubsection{Smoothness}
+The requirement of keeping a low acceleration where possible is validated through the middle rows of Fig.\ref{fig:velPlot1} and Fig.\ref{fig:velPlot2}.The vehicle is able to maintain a peak acceleration of 2 $\mathrm{m/s^2}$ for the vast majority of simulated trials, particularly when the pedestrian crosses from the opposing side of the road or when the vehicle is in the second to right line.
+
+However, there are a handful of simulated trials in Fig.\ref{fig:velPlot1} when the pedestrian accepts a gap between 1.75-2.5 seconds and the vehicle is in the right-most lane closest to the sidewalk.
+In this case, the vehicle must enter the \textit{Hard Braking} state and decelerate at a magnitude higher than $a_\mathrm{cmf}$.
+This is required in order to meet the collision avoidance constraint, which naturally takes priority.
+Again, note that this situation occurs relatively rarely, as pedestrians typically accept gaps between 3-6 seconds and are unlikely to risk hazardous behaviors unless distracted.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=3.0in]{figures/G80.jpg}
+\caption{Hyundai Genesis G80 experimental testbed.}
+\label{fig:g80}
+\end{figure}
 
@@
 \label{sec:expres}
+
+The proposed hybrid controller is tested experimentally using a Hyundai Genesis G80 experimental testbed, shown in Fig.
+\ref{fig:g80}.
+A mid-block intersection at Richmond Field Station near Berkeley's campus was chosen for the experimental validation.
+In lieu of an actual pedestrian, a pedestrian was simulated virtually using the ROS simulation architecture described earlier, with the position of the pedestrian sent to the vehicle in real time.
+Differential GPS was used to record the distance to the crosswalk, and a steering controller from \cite{kapania} was used to autonomously steer the vehicle in a straight line path.
+Relevant parameters of the experiments are shown below in Tables \ref{tb:expparams}.
+
+\begin{table}[h]
+\footnotesize
+\begin{center}
+\caption{Experimental Parameters}\label{tb:expparams}
+\begin{tabular}{lccc}
+Parameter & Symbol & Value & Units \\\hline\hline
+Number of lanes & $n$ & 2 & - \\
+Speed feedback gain & $k_s$ & 2.0 & 1 / sec\\
+Brake time delay    & $t_\mathrm{delay}$ & 0.5 & sec \\ 
+Speed limit & $v_\mathrm{speedlimit}$ & 7 & m/s \\
+Comfort acceleration & $a_\mathrm{cmf}$ & 2 & $\mathrm{m/s^2}$ \\
+Maximum time advantage & $t_\mathrm{max}$ & 4 & sec \\
+Maximum acceleration & $a_\mathrm{max}$ & 9 & $\mathrm{m/s^2}$ \\\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+Several key differences to highlight between the simulations and the experiments are that the number of lanes is decreased from 4 to 2 and that the speed limit is increased to 7 m/s to match the parameters of the experimental intersection.
+The experiment consisted of six trials intended to show operation of the four discrete modes of the controller.
+A description of the trial parameters are shown in Table \ref{tb:expparams2}.
+
+\begin{table}[h]
+\footnotesize
+\begin{center}
+\caption{Experimental Parameters}\label{tb:expparams2}
+\begin{tabular}{llll}
+Trial Number & Gap Accepted (sec) & Pedestrian Entry & Mode\\\hline\hline
+1 & 4.0 & right &  Yield\\ 
+2 & 1.0 & right &  Speed Up\\ 
+3 & 7.0 & right &  Yield\\ 
+4 & 2.5 & right &  Hard Brake\\ 
+5 & 3.0 & left  &  Yield \\ 
+6 & 1.0 & left  &  Speed Up\\\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+The experimental velocity and acceleration is plotted vs $d$ in 
+Figure \ref{fig:diagram1}.
+In Trial 1, the virtual pedestrian enters when the vehicle is four seconds from intersecting, giving the vehicle enough time to yield at a deceleration of 2 $m/s^2$.
+Note that the vehicle exhibits a large time delay between commanded and actual acceleration.
+Because this is accounted for in Algorithm 1, the vehicle is able to brake early and the vehicle is nearly stopped at $d = $ 1 meter before the pedestrian exits the crosswalk and the vehicle speeds back up.
+This overshoot, while undesirable, is realistically acceptable given the point $d=0$ is set to be several meters before the crosswalk actually begins.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=3.5in]{figures/expPlot.eps}
+\caption{Experimental velocity and acceleration plotted vs distance to crosswalk. Note that the sign of $d$ is opposite that of Fig.\ref{fig:diagram1}}
+\label{fig:g80}
+\end{figure}
+
+In Trial 2, the pedestrian enters from the right with just 1 second before collision.
+This experiment, while unlikely in reality, was undertaken to verify the vehicle would not brake, but accelerate through the intersection.
+Trial 3 represents a situation where the pedestrian crosses with a large (7 second) gap.
+In this case, the pedestrian has enough time to almost cross before the vehicle needs to yield, resulting in the vehicle only needing to slow down slightly.
+
+In Trial 4, the pedestrian exhibits risky behavior, crossing with a 2.5 second gap.
+In this case, the vehicle has enough time to brake, but must decelerate rapidly at 5 $m/s^2$ in the \textit{hard braking} mode.
+Given the brake delay in the experimental vehicle, significant overshoot occurs, and the vehicle stops nearly 2.5 meters ahead of the desired stopping location.
+Unlike in the yielding mode, the brake delay can not be compensated for because the vehicle must brake suddenly.
+
+Trials 5 and 6 represent pedestrian crossings from the opposite side of the road under accepted gaps of 3.0 and 1.0 seconds, respectively.
+In the former case, the vehicle is able to yield in a fashion similarly to Trial 1, and in the latter case, the vehicle is able to speed through the intersection, as in Trial 2.
 

### 2019-01-17 conference/iv2019/plotScripts/expPlot.m

@@ -1,0 +1,125 @@
+clear all; close all; clc;
+
+%% load bags
+t1 = load('/home/nkapania/Desktop/190110NRKbags/trial1.mat');
+t2 = load('/home/nkapania/Desktop/190110NRKbags/trial2.mat');
+t3 = load('/home/nkapania/Desktop/190110NRKbags/trial3.mat');
+t4 = load('/home/nkapania/Desktop/190110NRKbags/trial4.mat');
+t5 = load('/home/nkapania/Desktop/190110NRKbags/trial5.mat');
+t6 = load('/home/nkapania/Desktop/190110NRKbags/trial6.mat');
+
+
+%% create downsampled versions of actual acceleration
+N1 = length(t1.t);
+N2 = length(t2.t);
+N3 = length(t3.t);
+N4 = length(t4.t);
+N5 = length(t5.t);
+N6 = length(t6.t);
+
+t1.axDS = t1.ax( round(linspace(1, length(t1.ax), N1)) );
+t2.axDS = t2.ax( round(linspace(1, length(t2.ax), N2)) );
+t3.axDS = t3.ax( round(linspace(1, length(t3.ax), N3)) );
+t4.axDS = t4.ax( round(linspace(1, length(t4.ax), N4)) );
+t5.axDS = t5.ax( round(linspace(1, length(t5.ax), N5)) );
+t6.axDS = t6.ax( round(linspace(1, length(t6.ax), N6)) );
+
+%% get relevant indices
+ind1 = intersect(find(t1.distance_m > -5), find(t1.distance_m < 30));
+ind2 = intersect(find(t2.distance_m > -5), find(t2.distance_m < 30));
+ind3 = intersect(find(t3.distance_m > -5), find(t3.distance_m < 30));
+ind4 = intersect(find(t4.distance_m > -5), find(t4.distance_m < 30));
+ind5 = intersect(find(t5.distance_m > -5), find(t5.distance_m < 30));
+ind6 = intersect(find(t6.distance_m > -5), find(t6.distance_m < 30));
+
+
+%% Plot
+
+%Generate offset
+
+offset = 5.5;
+
+subplot(6, 2, 1)
[…]

### 2019-01-22 conference/iv2019/manuscript.tex
making edits from Vijay

@@
 
-\author{Nitin R. Kapania$^{1}$, Francesco Borrelli, J Christian Gerdes% <-this % stops a space
+\author{Nitin R. Kapania$^{1}$, Vijay Govindarajan$^{2}$, Francesco Borrelli$^{2}$, J Christian Gerdes$^{1}$% <-this % stops a space
 \thanks{*This work was supported by the Hyundai Center of Excellence}% <-this % stops a space
-\thanks{$^{1}$Nitin Kapania is with the Dept. of Mechanical Engineering, Stanford University, and is a visiting scholar at UC Berkeley}%
+\thanks{$^{1}$Dept. of Mechanical Engineering, Stanford University. Nitin Kapania is also a visiting scholar at UC Berkeley.}%
+\thanks{$^{2}$Dept. of Mechanical Engineering, UC Berkeley.}%
 }
+
+% \thanks{$^{2}$Dept. of Mechanical Engineering, UC Berkeley. 
+% }
 
@@
 over when the pedestrian should cross and when and how the vehicle should yield.
-By modeling pedestrian behavior through the concept of gap acceptance, the authors show that a hybrid controller with just four distinct modes allows an autonomous vehicle to successfully interact with a pedestrian across a continuous spectrum of possible crosswalk-entry behaviors.
-The controller is validated in simulation through comparison with a previously published control policy obtained through solution of a POMDP, and experimental results are provided on a Hyundai Genesis vehicle for a virtual pedestrian.
+By modeling pedestrian behavior through the concept of gap acceptance, the authors show that a hybrid controller with just four distinct modes allows an autonomous vehicle to successfully interact with a pedestrian across a continuous spectrum of possible crosswalk entry behaviors.
+The controller is validated through extensive simulation, and experimental results are provided on a Hyundai Genesis vehicle for a virtual pedestrian.
 
@@
 
-While autonomous vehicles have the potential to save thousands of lives every year and create tremendous societal benefits \cite{Fagnant2015}, widespread adoption is unlikely until AVs gain the broad trust of society.
+While autonomous vehicles have the potential to save thousands of lives every year and create significant societal benefits \cite{Fagnant2015}, widespread adoption is unlikely until AVs gain the broad trust of society.
 Given that every human is a pedestrian at some point during the day, one of the central ways that autonomous vehicles will be evaluated is through their interactions with pedestrians.
-In fact, recent events with industry leader Waymo highlight that AV-pedestrian interaction has still not been perfected\cite{MattRichtelandConorDougherty2015}.
+In fact, recent events with industry leader Waymo highlight that AV-pedestrian interaction has still not been perfected \cite{MattRichtelandConorDougherty2015}.
 In general, interactions with pedestrians are complex, even for experienced human drivers.
@@
 The need to further understand pedestrian behavior for autonomous driving has created a growing body of interdisciplinary research.
-One branch of research has focused around modeling pedestrian behavior given various sensor inputs.
+One branch of research has focused on modeling pedestrian behavior given various sensor inputs.
 Keller et al \cite{Keller2014} presented a study on pedestrian path prediction and action classification (e.g.
@@
 
-Another branch of literature is focused on developing predictive models for pedestrian crossing.
+Another branch of literature focuses on pedestrian behavior analysis using empirical data, typically in the form of video analysis.
 While typically studied for purposes of road design, this body of literature holds promising insights for AV designers.
@@
 Yannis et al.
-\cite{Yannis2013} and \cite{Sun2002} also found that gap acceptance was influenced by the size of the oncoming vehicle and the presence of other pedestrians.
+\cite{Yannis2013} and Sun et al.\cite{Sun2002} also found that gap acceptance was influenced by the size of the oncoming vehicle and the presence of other pedestrians.
 Lee and Aty \cite{Lee2005} studied interactions in the form of crashes, and found crashes were linked with higher daily traffic.
@@
 Finally, a small but rapidly growing body of literature specifically studies the interaction between pedestrians and autonomous vehicles at crosswalks.
-An excellent review of these studies was conducted by \cite{Rasouli}.
+An excellent review of these studies was conducted by Rasouli et al.
+\cite{Rasouli}.
 As an example, Rothenbucher et al.
 \cite{Rothenbucher2016} studied the interaction between pedestrians and driverless vehicles by constructing a car seat costume to disguise a driver.
-The authors noted that pedestrians overall managed interactions at crosswalks effectively, but later mentioned increased uncertainty about the autonomous vehicles behavior.
-To improve the issue of trust, several researchers have developed external interfaces to more clearly broadcast the intent of the autonomous vehicle \cite{Matthews}, \cite{Lagstrom2015}.
+To improve the issue of trust, several researchers have also developed external communication interfaces to more clearly broadcast the intent of the autonomous vehicle (\cite{Matthews}, \cite{Lagstrom2015}).
 
-Several real-world studies of AV-pedestrian interaction note that once a local population learned a vehicle was programmed to be perfectly safe, pedestrians would regularly take advantage of the AV and walk in front of it.
-\cite{Madigan}.
-This illustrates an issue with automated driving in that an overly conservative crossing algorithm will often be taken advantage of or cause confusion among pedestrians.
-Camara et al \cite{Camara2018} attempted to model the natural negotation for priority between a pedestrian and an AV at an intersection using the framework of ``sequential chicken" adopted from game theory.
+One issue with automated driving for crosswalk scenarios is that an overly conservative crossing algorithm will often be taken advantage of or cause confusion among pedestrians\cite{Camara2018}.
+Camara et al.
+\cite{Camara2018} attempted to model the natural negotation for priority between a pedestrian and an AV at an intersection using the framework of game theory.
 Chen et al.
 \cite{Chen} notes the need for a tradeoff between passive and aggressive driving behavior, and developed a stochastic model of pedestrian behavior to evaluate proposed AV control policies.
-Control polices for pedestrian interaction are also developed by \cite{Bandyopadhyay}, who proposes a Mixed-observable Markov Decision Process (MOMDP) to incorporate the intention uncertainty of a pedestrian.
-A POMDP formulation is also proposed by Thornton \cite{Thornton2018}, who proposes navigation of the pedestrian uncertainty through value-sensitive design.
+Control polices for pedestrian interaction are also developed in \cite{Bandyopadhyay}, who proposes a Mixed-observable Markov Decision Process (MOMDP) to incorporate the intention uncertainty of a pedestrian.
+A Partially-observable Markov Decision Process (POMDP) formulation is also proposed by Thornton \cite{Thornton2018}, who proposes a value-sensitive design framework to assist with the development of engineering specifications for the control algorithm.
 
@@
 An unsignalized crosswalk is chosen as it is the trickiest crosswalk for a pedestrian and vehicle to navigate, as there is no explicit declaration of whose turn it is cross.
-Simulations show that the state machine controller is able to handle a continuous spectrum of pedestrian \textit{gap acceptance} behavior, tolerating a range of highly conservative to highly aggressive pedestrians.
-Moreover, this is possible with just four distinct states.
-Simulations are also used to compare the closed-loop controller with the controller developed by \cite{Thornton2018}.
-Finally, experimental results are shown on a real vehicle to demonstrate the feasibility of the proposed state machine controller.
+Simulations show that the hybrid controller is able to handle a continuous spectrum of pedestrian \textit{gap acceptance} behavior, tolerating a range of highly conservative to highly aggressive pedestrians.
+Moreover, this is possible with just four distinct controller modes.
+Finally, experimental results are shown on a real vehicle to demonstrate the feasibility of the proposed controller in real-time.
 
@@
 In general, pedestrian crosswalks may be controlled or uncontrolled.
-In the case of the former, control devices such as stop signs or walk signals guide the interaction of vehicles and pedestrians explicitly.
-Fig.~\ref{fig:schematic} shows the latter example, in which a pedestrian must select a gap in traffic flow and cross.
+In the case of the former, control devices such as stop lights and walk signals guide the interaction of vehicles and pedestrians explicitly.
+Fig.~\ref{fig:schematic} shows the latter example, in which there is no such control device, and a pedestrian must select a gap in traffic flow and cross.
 
@@
 \centering
-\includegraphics[width=3.5in]{figures/example.eps}
-\caption{Schematic of pedestrian approaching a stream of traffic at a mid-block intersection.}
+\includegraphics[width=3.5in]{figures/example2.eps}
+\caption{Top: Description of relevant problem variables.
+Bottom: Schematic of pedestrian approaching a stream of traffic at a mid-block intersection.
+Ego vehicle shown in green.
+We develop a hybrid control framework for the green vehicle, which considers the distance ($d$) to a stopping point $\Delta$ ahead of the crosswalk and the pedestrian's position in the crosswalk ($x_p$).}
 \label{fig:schematic}
@@
 In general, right-of-way for uncontrolled intersections is complex.
-For example, nine states and the District of Columbia require motorists to stop when approaching a pedestrian in an uncontrolled crosswalk.
-Six states require a motorist to stop when a pedestrian is upon the same half of the roadway or within one lane of the lane that the motorist is traveling upon.
-Another nineteen states require a motorist to yield when a pedestrian is upon any portion of the roadway, and another 20 states mandate that motorists yield when a pedestrian is upon the same half of the roadway or approaching closely from the opposite side of the roadway \cite{NCSL}.
+For example, nine states and the District of Columbia require motorists to \textit{stop} when approaching a pedestrian in an uncontrolled crosswalk.
+Six states require a motorist to stop when a pedestrian is upon the same half of the roadway or within one lane of the lane that the motorist is traveling in.
+Another nineteen states require a motorist to \textit{yield} when a pedestrian is upon any portion of the roadway, and another 20 states mandate that motorists yield when a pedestrian is upon the same half of the roadway or approaching closely from the opposite side of the roadway \cite{Legislatures2018}.
 
-\subsection{Pedestrian Gap Acceptance}
-
-One of the major factors that determines pedestrian crossing behavior at uncontrolled intersections is \textit{gap acceptance}, defined in this paper as:
-
-\begin{equation}
-\mathrm{gap} = \frac{\mathrm{distance\hspace{1mm}to\hspace{1mm}crosswalk}}{\mathrm{vehicle\hspace{1mm}speed}}
-\end{equation}
-
-Similar to time to collision (TTC), the safety gap is a measure of how much time there is before the vehicle would enter the crosswalk if it kept it's current speed constant.
-When faced with a stream of traffic at a crosswalk, a pedestrian inherently decides how much of a gap to accept before crossing.
-The average gap acceptance is reported in the literature to be between 3-7 seconds, meaning pedestrians usually do not cross if the vehicle would enter the crosswalk in under three seconds \cite{DiPietroCharlesMandKing1970}, and are very likely to cross when they have more than seven seconds \cite{Schmidt2009}.
-
-\subsection{Problem Formalization and System Requirements}
+\subsection{Problem Formalization}
 \label{sec:probform}
 
-A diagram of the relevant state variables is shown below in Fig.~\ref{fig:diagram1}.
+A diagram of the relevant state variables is shown in Fig.~\ref{fig:schematic}.
 We assume that knowledge of distance $d$ to a fixed stopping point 3-5 meters ahead of the crosswalk can be estimated by a perception system, and that the velocity of the vehicle $\dot{d}$ is known as well.
-Furthermore, we also assume the perception system is able to determine the position of the pedestrian $x_p$ within the crosswalk, and is able to determine a velocity estimate $\dot{x_p}$ for the pedestrian as well.
+Furthermore, we also assume the perception system is able to determine the position of the pedestrian $x_p$ within the crosswalk, and is able to determine a velocity estimate $\dot{x}_p$ for the pedestrian as well.
 
-\begin{figure}
-\centering
-\includegraphics[width=1in]{figures/diagram.eps}
-\caption{Description of relevant problem variables.}
-\label{fig:diagram1}
-\end{figure}
+% \begin{figure}
+% \centering
+% \includegraphics[width=1in]{figures/diagram.eps}
+% \caption{}
+% \label{fig:diagram1}
+% \end{figure}
 
@@
 In general, this control problem is difficult as there are multiple stakeholders and therefore multiple competing objectives that must be considered.
-For example, relevant stakeholders for this problem are the autonomous vehicle and its passengers, the pedestrian(s), and the other vehicles in the traffic stream.
+For example, relevant stakeholders for this problem are the autonomous vehicle occupants, the pedestrian(s), and the other vehicles in the traffic stream.
 
 As a result, an algorithm that simply yields every time a pedestrian is near the crosswalk is likely to be too conservative, as there is a likelihood the pedestrian is waiting for a larger gap in the traffic stream to cross or is walking through the sidewalk without intending to cross in the first place.
-To formalize the system requirements for this multi-stake holder problem, we propose the following \textit{design constraints}, extended from the engineering specifications proposed by Thornton \cite{Thornton2018}.
+
+\subsection{System Requirements}
+
+To formalize the system requirements for this multi-stake holder problem, we propose the following \textit{design requirements}, extended from the engineering specifications proposed by Thornton \cite{Thornton2018}.
 
@@
  \textbf{Engineering} & \textbf{Design} \\
- \textbf{Specification} & \textbf{Constraint}\\\hline 
+ \textbf{Specification\cite{Thornton2018}} & \textbf{Constraint}\\\hline 
  1. Safety and Legality & a. Avoid collisions for all\\
@@
                & \\
- 3. Smoothness & a. Low accel: Nominal brake / \\ 
- 			   & accel limits of 2 m/s/s \\
- 			   & b. Stop within 3-5 meters\\
- 			   & of crosswalk
+ 3. Smoothness & a. Low acceleration: Nominal brake \\ 
+ 			   & and acceleration limits of 2 $\mathrm{m/s^2}$ \\
 
@@
 
+\subsection{Pedestrian Gap Acceptance}
+
+One of the major factors that determines pedestrian crossing behavior at uncontrolled intersections is \textit{gap acceptance}, defined in this paper as:
+
+\begin{equation}
+\mathrm{gap} = \frac{\mathrm{distance\hspace{1mm}to\hspace{1mm}crosswalk}}{\mathrm{vehicle\hspace{1mm}speed}}
+\end{equation}
+
+Similar to time to collision (TTC), the safety gap is a measure of how much time there is before the vehicle would enter the crosswalk if it kept its current speed constant.
+When faced with a stream of traffic at a crosswalk, a pedestrian inherently decides how much of a gap to accept before crossing.
+The average gap acceptance is reported in the literature to be between 3-7 seconds, meaning pedestrians usually do not cross if the vehicle would enter the crosswalk in under three seconds \cite{DiPietroCharlesMandKing1970}, and are very likely to cross when they have more than seven seconds \cite{Schmidt2009}.
+
 \section{Proposed Control Architecture}
 
-The proposed control architecture is shown below in Fig.~\ref{fig:hybridController}, and the algorithm for operation is shown in Algorithm \ref{alg:hybridController}.
-At heart, the control algorithm uses a feedback-feedforward methodology to compute desired acceleration commands $\ddot{d}$, but computes these commands differently depending on which of four discrete states the controller is in.
-The next sections will describe the four states.
+The proposed control architecture is shown in Fig.~\ref{fig:hybridController}, and the algorithm for operation is shown in Algorithm \ref{alg:hybridController}.
+At heart, the control algorithm uses a feedback-feedforward methodology to compute desired acceleration commands $\ddot{d}$, but computes these commands differently depending on which of four discrete modes the controller is in.
+The next section will describe the four modes.
 
@@
 \centering
-\includegraphics[width=3.5in]{figures/stateDiagram.eps}
+\includegraphics[width=3.5in]{figures/stateDiagram2.eps}
 \caption{Schematic of hybrid control structure.}
@@
 \end{figure}
+
+\begin{table}[h]
+\begin{center}
+\caption{Parameters for Algorithm 1}\label{tb:params}
+\begin{tabular}{lc}
+Symbol & Definition \\\hline\hline
+$d_\mathrm{max}$ & $\frac{\dot{d}^2}{2a_\mathrm{max}}$ \\
+$d_\mathrm{cmf}$ & $\frac{\dot{d}^2}{2a_\mathrm{cmf}}$ \\\hline
+\end{tabular}
+\end{center}
+\end{table}
 
@@
 
-		\ElsIf{$d > \frac{\dot{d}^2}{2a_\mathrm{cmfrt}}$}
+		\ElsIf{$d > d_\mathrm{cmf}$}
 		\State $state \leftarrow YIELDING$
 
-	\ElsIf{$d < \frac{\dot{d}^2}{2a_\mathrm{cmfrt}} \hspace{1mm} \mathbf{and} \hspace{1mm} d > \frac{\dot{d}^2}{2a_\mathrm{max}}$}
+	% \ElsIf{$d < \frac{\dot{d}^2}{2a_\mathrm{cmf}} \hspace{1mm} \mathbf{and} \hspace{1mm} d > \frac{\dot{d}^2}{2a_\mathrm{max}}$}
+
+  \ElsIf{$d_\mathrm{max} < d < d_\mathrm{cmf} $}
+
 	\State $state \leftarrow HARD\hspace{1mm}BRAKING$
@@
 \If {$state = YIELDING$}
-\If {$d > \frac{\dot{d}^2}{2a_\mathrm{cmfrt}} + t_{delay}\dot{d}$}
+\If {$d > d_\mathrm{cmf} + t_{delay}\dot{d}$}
 	\State $\ddot{d}_{des} = 0$
@@
 	\State $\ddot{d} = a_\mathrm{cmf}$ 	
-	\If {$\textbf{not} \hspace{1mm} \mathrm{inCrosswalk}(x_p)$} \State $state \leftarrow DRIVING$
+	\If {$\textbf{not} \hspace{1mm} \mathrm{inCrosswalk}(x_p) \vee d < 0$} \State $state \leftarrow DRIVING$
 
@@
 In the nominal state, the vehicle attempts to drive through the crosswalk at the speed limit $v_\mathrm{speedlimit}$.
-A simple proportional speed control is applied to keep the vehicle at the speed limit, achieving design constraint 2.a.:
+A simple proportional speed control is applied to keep the vehicle at the speed limit, achieving design constraint 2.a:
 
@@
 
-The controller remains in the driving mode unless the pedestrian begins to enter the crosswalk while $d > 0$.
-To allow more time for the controller to respond, we allow the $\mathrm{inCrosswalk}(x_p, \dot{x}_p)$ indicator function to be triggered the moment the pedestrian approaches the crosswalk from the sidewalk - e.g.
-once the pedestrian velocity $\dot{x}_p$ becomes non-zero.
+The controller remains in the driving mode unless the pedestrian begins to enter the crosswalk before the vehicle has passed.
+In this case, we mathematically define ``beginning to enter" as the instant the pedestrian begins moving towards the crosswalk, even if the pedestrian is currently on the sidewalk:
 
- Note that according to design constraint 1.b, the definition of entering the crosswalk will vary according to which of the fifty US states the vehicle is in.
-For example, in California, the pedestrian entering any portion of the crosswalk should trigger the control mode to change - while in Louisiana, the pedestrian entering the same half of the roadway as the vehicle should trigger the change \cite{NCSL2018}.
+\begin{equation}
+\label{eq:crossing}
+\mathrm{inCrosswalk}(x_p, 
+\dot{x}_p) = \left\{
+  \begin{array}{ll}
+    1&: \dot{x}_p \neq 0 \vee \hspace{1 mm} 0 \leq x_p \leq x_F \\
+    0&: \mathrm{otherwise} 
+  \end{array}
+\right.
+\end{equation}
+
+ Where $x_F$ is the end of the crosswalk area of interest.
+Note that according to design constraint 1.b, the definition of $x_F$ will vary according to which US state the vehicle is in.
+For example, in California, $x_F$ would be at the end of the crosswalk, while in Louisiana, $x_F$ would just be the same half of the crosswalk as the vehicle \cite{Legislatures2018}.
 
@@
 
-Frequently he vehicle will pass through the intersection well before the pedestrian will reach the vehicle's lane.
+Frequently the vehicle will pass through the intersection well before the pedestrian will reach the vehicle's lane.
 Assuming the vehicle is operating in a US state where stopping is not explicitly required, design constraint 2.a suggests it is desirable for the vehicle to continue through the intersection.
 
-This can be formalized by calculating the \textit{time advantage} $t_adv$, also known as the \textit{time to collision} as follows:
+This can be formalized by calculating the \textit{time advantage} $t_\mathrm{adv}$, also known as the \textit{time to collision} as follows:
 
 \begin{equation}
-t_{adv} = \frac{d}{\dot{d}} - \frac{x_v - x_p}{\dot{x_p}}
+t_\mathrm{adv} = \frac{d}{\dot{d}} - \frac{x_v - x_p}{\dot{x_p}}
 \end{equation}
@@
 Where $x_v$ is the $x$ position of the vehicle in the crosswalk.
-The control algorithm in this work will explicitly allow the vehicle to continue driving through the intersection as long as $t_{adv}$ exceeds a specfiied threshold $t_{max}$.
+Algorithm 1 explicitly allows the vehicle to continue driving through the intersection as long as $t_\mathrm{adv}$ exceeds a specified threshold $t_\mathrm{max}$.
 
@@
 If the time advantage is not sufficient for the vehicle to pass through, the autonomous vehicle must slow down.
-The preferred option to meet design constraint 3.a is to brake at a low, comfortable deceleration $-a_{cmf}$ for the passenger.
-Note that design constraint 3.b is satisfied by having $d=0$ be defined to be a point several meters ahead of the crosswalk.
+The preferred option to meet design constraint 3.a is to brake at a low, comfortable deceleration $-a_\mathrm{cmf}$ for the passenger.
 
-Comfortable braking at a deceleration $-a_{cmf}$ is possible if sufficient braking distance exists. The braking distance is derived kinematically as
+Comfortable braking at a deceleration $-a_\mathrm{cmf}$ is possible if sufficient braking distance exists. The braking distance is derived kinematically as
 
 \begin{equation}
-d_{brake} = \frac{\dot{d}^2}{2a_{cmf}}
+d_{brake} = \frac{\dot{d}^2}{2a_\mathrm{cmf}}
 \end{equation}
 
-If this holds true when the pedestrian starts to enter the crosswalk, the car enters a $Yielding$ state.
+If this holds true when the pedestrian starts to enter the crosswalk, the car enters a \textit{Yielding} mode.
 
@@
 In the unlikely event a pedestrian begins to cross when the time gap is low (e.g.
-under 3 seconds), the vehicle will need to prioritize design constraint 1.a and come to a stop at a deceleration higher than $-a_{cmf}$.
+under 3 seconds), the vehicle will need to prioritize design constraint 1.a and come to a stop at a deceleration higher than $-a_\mathrm{cmf}$.
 This occurs when the following conditions holds:
@@
 \begin{equation}
-\frac{\dot{d}}{2a_{max}} < d < \frac{\dot{d}^2}{2a_{cmf}}
+\frac{\dot{d}^2}{2a_{max}} < d < \frac{\dot{d}^2}{2a_\mathrm{cmf}}
 \end{equation}
@@
 
-\subsection{Speeding Up}
+\subsubsection{Speeding Up}
 
-As a final condition, consider the case where the pedestrian enters the crosswalk just as the vehicle is crossing $d = 0$ - in this case, it makes little sense to decelerate, as the vehicle has insufficient space to stop before the crosswalk and will risk being rear-ended (Design Constraint 2.a) or stopping in the crosswalk.
+As a final condition, consider the case where the pedestrian enters the crosswalk just as the vehicle is crossing $d = 0$.
+In this case, it makes little sense to decelerate, as the vehicle has insufficient space to stop before the crosswalk and will risk being rear-ended or stopping in the crosswalk.
 It makes more sense in this condition for the vehicle to speed up and exit the crosswalk quickly.
@@
 In the yielding state, the vehicle follows a different version of the feedback-feedforward dynamics.
-The vehicle first follows the speed limit until it reaches the critical value of $\frac{\dot{d}^2}{2a_{cmf}}$.
-In practice, an additional term $t_delay\dot{d}$ is added to compensate for the delay in brake communication.
+The vehicle first follows the speed limit until it reaches the critical value of $\frac{\dot{d}^2}{2a_\mathrm{cmf}}$.
+In practice, an additional term $t_\mathrm{delay}\dot{d}$ is added to compensate for the delay in brake communication.
 
-At the critical istance, the vehicle decelerates at the uniform yielding deceleration of $-a_{cmf}$, with an additional feedback term $k_s(\dot{d} - \dot{d}_des$), where $\dot{d}_des$ is the desired velocity.
+At the critical distance, the vehicle decelerates at the uniform yielding deceleration of $-a_\mathrm{cmf}$, with an additional feedback term $k_s(\dot{d} - \dot{d}_\mathrm{des}$), where $\dot{d}_\mathrm{des}$ is the desired velocity.
 The feedback term helps bring the vehicle speed to 0 at the desired stopping point.
 
-The desired velocity profile $\dot{d}(d)$ for a vehicle decelerating at constant acceleration is given via integration:
+The desired velocity profile $\dot{d}_\mathrm{des}(d)$ for a vehicle decelerating at constant acceleration is given by:
 
 \begin{equation}
-\dot{d}(d) = \sqrt{2a_{cmf}(d-d_o) + \dot{d}_o^2}
+\dot{d}_\mathrm{des}(d) = \sqrt{2a_\mathrm{cmf}(d-d_o) + \dot{d}_o^2}
 \end{equation}
 
-Where $d_o$ and $\dot{d}_o$ is the value of $d$ and $\dot{d}$ when the controller first enters the yielding state.
-Simple inspection shows that the desired velocity is $\dot{d}_o$ when $d = d_o$ and is 0 when $d = 0$ (recall that we enter the yielding state when $d_o = \frac{\dot{d_o}^2}{2a_{cmf}}$.
+Where $d_o$ and $\dot{d}_o$ are the values of $d$ and $\dot{d}$ when the controller first enters the yielding state.
+Inspection shows that the desired velocity is $\dot{d}_o$ when $d = d_o$ and is 0 when $d = 0$ (recall that we enter the yielding state when $d_o = \frac{\dot{d_o}^2}{2a_\mathrm{cmf}}$).
 
-The controller exits the yielding mode and returns to the driving mode once the pedestrian clears the crosswalk.
+The controller exits the yielding mode and returns to the driving mode once the pedestrian clears the crosswalk (e.g. once $x > x_F$ in (\ref{eq:crossing})).
 
@@
 
-In the hard braking state, the vehicle follows a similar feedback-feedforward algorithm, but the desired feedforward deceleration in this case is given by $\ddot{d} = -\frac{d^2}{2d}$.
-Integrating to find the desired speed profile $\dot{d}(d)$ yields the following deceleration:
+In the hard braking state, the vehicle follows a similar feedback-feedforward algorithm, but the desired feedforward deceleration in this case is given by $\ddot{d} = -\frac{\dot{d}^2}{2d}$.
+Integrating to find the desired speed profile $\dot{d}_\mathrm{des}(d)$ yields the following deceleration:
 
 \begin{equation}
-\dot{d}(d) = \frac{\dot{d}_o}{\sqrt{d_o}}\sqrt{d}
+\dot{d}_\mathrm{des}(d) = \frac{\dot{d}_o}{\sqrt{d_o}}\sqrt{d}
 \end{equation}
 
-Where $d_o$ and $\dot{d}_o$ is the value of $d$ and $\dot{d}$ when the controller first enters the braking state.
+Where $d_o$ and $\dot{d}_o$ are the values of $d$ and $\dot{d}$ when the controller first enters the braking state.
 Again, the controller exits the braking mode and returns to the driving mode once the pedestrian clears the crosswalk.
@@
 In the speed up state, there is no need to follow an exact speed profile as the vehicle is merely trying to exit the crosswalk area slightly faster.
-In this case, the commanded acceleration $\ddot{d} = a_{cmf}$ until the pedestrian clears the crosswalk.
+In this case, the commanded acceleration is $\ddot{d} = a_\mathrm{cmf}$ until the pedestrian clears the crosswalk or once the vehicle passes the stopping location (i.e.
+$d< 0$).
 
@@
 In order to check performance of the algorithm, both simulation and experimental studies were undertaken.
-The simulation results focused on testing Algorithm 1 against the design requirements from Table \ref{tb:specs} over many pedestrian crossings.
+The simulation results focused on testing Algorithm 1 against the specified design requirements over many pedestrian crossings.
 
@@
 \ref{fig:simFramework}) developed in Python.
-The simulation environment is built on the ROS \cite{ROS} robotics interface to enable easy porting of the control architecture to the experimental vehicle in Section \ref{sec:expres}.
-The simulation includes features such as multiple lanes, the pedestrian crossing on the right or the left, and animation capabilities.
+The simulation environment is built on the ROS robotics interface to enable easy porting of the control architecture to the experimental vehicle in Section \ref{sec:expres}.
+The simulation includes features such as customizable lanes, the pedestrian crossing on the right or the left, and animation capabilities.
 
@@
 \includegraphics[width=2.5in]{figures/simFramework.png}
-\caption{Crosswalk simulation environment for algorithm validation. Ego vehicle is shown in black, following vehicle in traffic stream shown in blue, and pedestrian shown in red.}
+\caption{Crosswalk simulation environment for algorithm validation. Ego vehicle is shown in black, leading vehicle in traffic stream shown in blue, and pedestrian shown in red.}
 \label{fig:simFramework}
@@
 
-To simulate random pedestrian crossing events, a pedestrian is spawned in an area near the crosswalk.
+To simulate random pedestrian crossing events, a pedestrian is spawned in an area of influence on the sidewalk.
 The pedestrian accepts a randomly sized time gap in the traffic stream to cross the road at constant speed $\dot{x}_p$.
 The size of the time gap is drawn from a normal distribution with mean $\mu_{gap}$ and standard deviation $\sigma_{gap}$.
-Statistics for $\mu_{gap}$ and $\sigma_{gap}$ are obtained from a video graphic analysis by Feliciani et al.
+Statistics for $\mu_{gap}$ and $\sigma_{gap}$ are obtained from results of the video graphic analysis by Feliciani et al.
 \cite{Feliciani2017}.
@@
 Number of lanes & $n$ & 4 & - \\
-Gap variance & $\sigma_{gap}$ & 2.5 & sec\\
+Gap variance & $\sigma^2_{gap}$ & 2.5 & sec\\
+Pedestrian velocity & $\dot{x}_p$ & 1.2 & $\mathrm{m/s}$ \\
 Mean accepted gap & $\mu_{gap}$ & 4.0 & sec\\\hline
@@
 Figure \ref{fig:velPlot1} shows simulation results for the case where the pedestrian enters the crosswalk on the same side of the road as the vehicle.
-The left side of the figure is for the case where the vehicle is on the right-most lane of a four way road, while the right side of the figure is for the case where the vehicle is in the second to right lane.
-Every marker in the figure represents a simulated vehicle crossing.
+The left column of the figure is for the case where the vehicle is on the right-most lane of a four way road (marked ``Lane A"), while the right column of the figure is for the case where the vehicle is in the second to right lane (Lane B).
+Every marker in the figure represents one of 750 simulated vehicle crossings.
 
@@
 
-To meet the collision avoidance design requirement, we generated enough simulations to ensure errant pedestrian crossing behavior at "risky" accepted gaps (< 3 seconds).
-The minimum distance between the vehicle and the pedestrian for each trial is shown in the bottom rows of Fig.
-\ref{fig:velPlot1} and Fig.\ref{fig:velPlot2}.
-In general, the vehicle is able to maintain a safe distance of at least four meters from the pedestrian for the case where the vehicle is in the second to right lane.
+To meet the collision avoidance design requirement, we generated enough simulations to ensure errant pedestrian crossing behavior at ``risky" accepted gaps ($<$~3 seconds).
+The minimum distance between the vehicle and the pedestrian for each trial is shown in (a) and (b) of Figures ~\ref{fig:velPlot1} and \ref{fig:velPlot2}.
+The vehicle is able to maintain a safe distance of at least four meters from the pedestrian for the case where the vehicle is Lane B.
 This is because the algorithm has enough time to react to a pedestrian crossing no matter what size gap is accepted.
 
-The riskiest case occurs when the vehicle is in the right-most lane and the pedestrian crosses from the right side, accepting a gap between 1.25 and 1.75 seconds.
-Note that this is in general unlikely given gap acceptance statistics reported in the literature \cite{Feliciani2017}\cite{Rasouli}, but would represent a safety risk given the vehicle has limited time to react.
+The riskiest case occurs when the vehicle is in Lane A and the pedestrian crosses from the right side, accepting a gap between 1.25 and 1.75 seconds.
+Note that this is in general unlikely given gap acceptance statistics reported in the literature \cite{Rasouli}\cite{Feliciani2017}, but would represent a safety risk given the vehicle has limited time to react.
 However, since the algorithm recognizes that braking will not clear the crosswalk, the vehicle is able to speed up and maintain at least 2 meters of lateral distance from the pedestrian at all times.
-Note that a better way to confirm robustness of the controller to all reasonable pedestrian behaviors would be through a verification analysis - this will be the focus of future work.
+Note that a better way to confirm robustness of the controller to all reasonable pedestrian behaviors would be through a theoretical verification analysis - this will be the focus of future work.
 
@@
 
-The requirement to minimize deviation from speed of traffic is validated in the top row of Fig.\ref{fig:velPlot1} and Fig.\ref{fig:velPlot2}, which shows the average velocity of the vehicle for each simulated crossing.
+The requirement to minimize deviation from speed of traffic is validated in (c) and (d) of Figures ~\ref{fig:velPlot1} and ~\ref{fig:velPlot2}, which shows the average velocity of the vehicle for each simulated crossing.
 For high accepted gaps, the pedestrian simply waits for the ego vehicle to exit the crosswalk before beginning to walk, and the algorithm does not need to slow down from the speed of traffic.
-For low accepted gaps, the vehicle cannot stop, and must speed up or continue driving through the intersection, also keeping the speed of traffic high.
+For low accepted gaps, the vehicle cannot stop, and must speed up or continue driving through the intersection, also keeping the vehicle speed high.
 
-The most disruption to traffic flow occurs when the vehicle must come to a complete stop, which occurs for accepted gaps between 2-6 seconds.
+The most disruption to traffic flow occurs when the vehicle must slow down significantly or stop, which occurs for accepted gaps between 2-6 seconds.
 In this case, the average velocity increases as the accepted gap increases.
-This is because as the pedestrian enters the crosswalk earlier, they are closer to exiting when the vehicle must yield, resulting in less time stopped and higher average velocity.
+Larger accepted gaps give the pedestrian more time to walk before the vehicle approaches, meaning they are closer to exiting the crosswalk when the vehicle must yield.
+This means the vehicle does not need to stop for as much time, resulting in higher average velocity.
 
-Note that a special case occurs in Fig.\ref{fig:velPlot2} where the vehicle is in the right-most lane of the four lane road and the pedestrian crosses from the other side.
+Note that a special case occurs in Fig.~\ref{fig:velPlot2} where the vehicle is in Lane A and the pedestrian crosses from the opposite side of the road.
 Because the vehicle is at a far distance from the pedestrian, the time advantage remains sufficiently high for the vehicle to comfortably continue through the crosswalk without needing to slow down.
 
-Table \ref{tb:speeds} shows a summary of the average speed compared to the baseline speed of 4.4 m/s.
+Table \ref{tb:speeds} shows a summary of the average speed compared to the baseline speed of 4.4 m/s for each of the four cases.
 
@@
 \subsubsection{Smoothness}
-The requirement of keeping a low acceleration where possible is validated through the middle rows of Fig.\ref{fig:velPlot1} and Fig.\ref{fig:velPlot2}.The vehicle is able to maintain a peak acceleration of 2 $\mathrm{m/s^2}$ for the vast majority of simulated trials, particularly when the pedestrian crosses from the opposing side of the road or when the vehicle is in the second to right line.
+The requirement of keeping a low acceleration where possible is validated through (e) and (f) of Figures~\ref{fig:velPlot1} and \ref{fig:velPlot2}.
+The vehicle is able to maintain a peak acceleration of 2 $\mathrm{m/s^2}$ for the vast majority of simulated trials, particularly when the pedestrian crosses from the opposing side of the road or when the vehicle is in the second to right line.
 
-However, there are a handful of simulated trials in Fig.\ref{fig:velPlot1} when the pedestrian accepts a gap between 1.75-2.5 seconds and the vehicle is in the right-most lane closest to the sidewalk.
+However, there are a handful of simulated trials in Fig.~\ref{fig:velPlot1} when the pedestrian accepts a gap between 1.75-2.5 seconds and the vehicle is in Lane A.
 In this case, the vehicle must enter the \textit{Hard Braking} state and decelerate at a magnitude higher than $a_\mathrm{cmf}$.
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[width=2.0in]{figures/crosswalkPic.png}
+\caption{Two lane crosswalk used for experimental testing.}
+\label{fig:crosswalkpic}
+\end{figure}
+
 \section{Experimental Results}
@@
 \ref{fig:g80}.
-A mid-block intersection at Richmond Field Station near Berkeley's campus was chosen for the experimental validation.
-In lieu of an actual pedestrian, a pedestrian was simulated virtually using the ROS simulation architecture described earlier, with the position of the pedestrian sent to the vehicle in real time.
-Differential GPS was used to record the distance to the crosswalk, and a steering controller from \cite{kapania} was used to autonomously steer the vehicle in a straight line path.
-Relevant parameters of the experiments are shown below in Tables \ref{tb:expparams}.
+A two lane unsignalized intersection at Richmond Field Station near Berkeley's campus was chosen for the experimental validation (Fig.~\ref{fig:crosswalkpic}).
+In lieu of an actual pedestrian, a pedestrian was simulated virtually using the ROS simulation architecture, with the position of the pedestrian passed to the vehicle controller in real time.
+Differential GPS was used to record the distance to the crosswalk, and a steering controller from \cite{Kapania2015} was used to autonomously steer the vehicle down the road.
+Relevant parameters of the experiments are shown below in Table \ref{tb:expparams}.
 
@@
 Number of lanes & $n$ & 2 & - \\
-Speed feedback gain & $k_s$ & 2.0 & 1 / sec\\
+Speed feedback gain & $k_s$ & 0.0 & 1 / sec\\
 Brake time delay    & $t_\mathrm{delay}$ & 0.5 & sec \\ 
@@
 Several key differences to highlight between the simulations and the experiments are that the number of lanes is decreased from 4 to 2 and that the speed limit is increased to 7 m/s to match the parameters of the experimental intersection.
+Also note that the speed tracking feedback is disabled as it triggered fault conditions in Hyundai's low-level controller.
 The experiment consisted of six trials intended to show operation of the four discrete modes of the controller.
@@
 The experimental velocity and acceleration is plotted vs $d$ in 
-Figure \ref{fig:diagram1}.
-In Trial 1, the virtual pedestrian enters when the vehicle is four seconds from intersecting, giving the vehicle enough time to yield at a deceleration of 2 $m/s^2$.
-Note that the vehicle exhibits a large time delay between commanded and actual acceleration.
-Because this is accounted for in Algorithm 1, the vehicle is able to brake early and the vehicle is nearly stopped at $d = $ 1 meter before the pedestrian exits the crosswalk and the vehicle speeds back up.
-This overshoot, while undesirable, is realistically acceptable given the point $d=0$ is set to be several meters before the crosswalk actually begins.
+Figure \ref{fig:expPlot}.
+In Trial 1, the virtual pedestrian enters when the vehicle is four seconds from intersecting, giving the vehicle enough time to yield at a deceleration of 2 $\mathrm{m/s^2}$.
+Note that the low level acceleration controller exhibits a large time delay between commanded and actual acceleration.
+Because this is accounted for in Algorithm 1, the vehicle is able to brake early and the vehicle is nearly stopped 1 meter ahead of the desired stop point before the pedestrian exits the crosswalk and the vehicle speeds back up.
+This 1 meter overshoot, while undesirable, is realistically acceptable given the desired stopping point $d=0$ is set to be five meters before the crosswalk actually begins.
 
@@
 \includegraphics[width=3.5in]{figures/expPlot.eps}
-\caption{Experimental velocity and acceleration plotted vs distance to crosswalk. Note that the sign of $d$ is opposite that of Fig.\ref{fig:diagram1}}
-\label{fig:g80}
+\caption{Experimental velocity and acceleration plotted vs distance to crosswalk.
+Note that the sign of $d$ is opposite that of Fig.~\ref{fig:diagram1}.
+Green circles mark the point where the pedestrian first enters crosswalk.
+Red circles mark the point where the pedestrian leaves the crosswalk.
+Note that for Trial 3, the green circle is off the page at -40 meters.}
+\label{fig:expPlot}
 \end{figure}
@@
 In Trial 4, the pedestrian exhibits risky behavior, crossing with a 2.5 second gap.
-In this case, the vehicle has enough time to brake, but must decelerate rapidly at 5 $m/s^2$ in the \textit{hard braking} mode.
+In this case, the vehicle has enough time to brake, but must decelerate rapidly at 5 $\mathrm{m/s^2}$ in the \textit{hard braking} mode.
 Given the brake delay in the experimental vehicle, significant overshoot occurs, and the vehicle stops nearly 2.5 meters ahead of the desired stopping location.
@@
 \section{CONCLUSIONS}
+This paper proposes a novel hybrid control architecture for determing closed loop vehicle control at an unsignalized intersection.
+Through simulation, we show that the system ensures safety across a continuous spectrum of pedestrian gap acceptance behaviors, while balancing the competing demands of smoothness for the vehicle occupants and limited traffic slowdown for other vehicles in the traffic stream.
+Remarkably, with just four distinct modes, the controller is able to handle multiple lanes and pedestrians crossing from either side of the road.
+Further work will focus heavily on integrating perception to accurately determine when to trigger switches in the controller modes, as determining pedestrian intent to cross is a major input required for the controller.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{APPENDIX}
-
-Appendixes should appear before the acknowledgment.
-
-\section*{ACKNOWLEDGMENT}
 

### 2019-01-22 conference/iv2019/plotScripts/expPlot.m

@@ -25,12 +25,12 @@ t5.axDS = t5.ax( round(linspace(1, length(t5.ax), N5)) );
 t6.axDS = t6.ax( round(linspace(1, length(t6.ax), N6)) );
 
 %% get relevant indices
-ind1 = intersect(find(t1.distance_m > -5), find(t1.distance_m < 30));
-ind2 = intersect(find(t2.distance_m > -5), find(t2.distance_m < 30));
-ind3 = intersect(find(t3.distance_m > -5), find(t3.distance_m < 30));
-ind4 = intersect(find(t4.distance_m > -5), find(t4.distance_m < 30));
-ind5 = intersect(find(t5.distance_m > -5), find(t5.distance_m < 30));
-ind6 = intersect(find(t6.distance_m > -5), find(t6.distance_m < 30));
+ind1 = intersect(find(t1.distance_m > -5), find(t1.distance_m < 35));
+ind2 = intersect(find(t2.distance_m > -5), find(t2.distance_m < 35));
+ind3 = intersect(find(t3.distance_m > -5), find(t3.distance_m < 35));
+ind4 = intersect(find(t4.distance_m > -5), find(t4.distance_m < 35));
+ind5 = intersect(find(t5.distance_m > -5), find(t5.distance_m < 35));
+ind6 = intersect(find(t6.distance_m > -5), find(t6.distance_m < 35));
 
 
 %% Plot
@@ -40,80 +40,80 @@ ind6 = intersect(find(t6.distance_m > -5), find(t6.distance_m < 30));
 offset = 5.5;
 
 subplot(6, 2, 1)
-plot(-t1.distance_m(ind1)+offset, t1.velocity_mps(ind1),'LineWidth',2)
+plot(-t1.distance_m(ind1)+offset, t1.velocity_mps(ind1),'LineWidth',2,'FontSize',12)
 title('Velocity m/s')
 ylabel('Trial 1')
 ylim([0,8])
-xlim([-20 10])
+xlim([-30 10])
 
 subplot(6, 2, 2)
-plot(-t1.distance_m(ind1)+offset, t1.ddxV(ind1),'LineWidth',2)
[…]

### 2019-01-22 conference/iv2019/plotScripts/velocityPlot.m

@@ -75,52 +75,50 @@ end
 
 %%
 subplot(3,2,1)
-scatter(dat1.acceptedGapDownsample, dat1.avgVelocity,'k*')
 grid on; hold on; 
-title('Rightmost Lane')
-ylim([0, 5]);
+title('Rightmost Lane','FontSize',18)
+scatter(dat1.acceptedGapDownsample, dat1.minDistance,'k*')
+ylim([0, 13]);
 xlim([0, 9]);
-ylabel('Average Velocity (m/s)')
+ylabel('Minimum Distance to Pedestrian(m)')
+
 
 subplot(3,2,2)
-scatter(dat2.acceptedGapDownsample, dat2.avgVelocity,'ro')
 grid on; hold on; 
-title('Second to Right Lane')
-ylim([0, 5]);
+title('Second to Right Lane','FontSize',18)
+scatter(dat2.acceptedGapDownsample, dat2.minDistance,'ro')
+ylim([0, 9]);
 xlim([0, 9]);
-ylabel('Average Velocity (m/s)')
 
 subplot(3,2,3)
+scatter(dat1.acceptedGapDownsample, dat1.avgVelocity,'k*')
+grid on; hold on; 
+ylim([0, 5]);
+xlim([0, 9]);
+ylabel('Average Velocity (m/s)','FontSize',18)
+
+subplot(3,2,4)
+scatter(dat2.acceptedGapDownsample, dat2.avgVelocity,'ro')
+grid on; hold on; 
+
+ylim([0, 5]);
+xlim([0, 9]);
+
+subplot(3,2,5)
 grid on; hold on; 
 scatter(dat1.acceptedGapDownsample, dat1.maxAccel,'k*')
 ylim([0, 9]);
 xlim([0, 9]);
+xlabel('Pedestrian Accepted Gap (s)')
 ylabel('Maximum Acceleration (m/s2)')
 
-subplot(3,2,4)
+subplot(3,2,6)
 grid on; hold on; 
 scatter(dat2.acceptedGapDownsample, dat2.maxAccel,'ro')
 ylim([0, 9]);
 xlim([0, 9]);
 ylabel('Maximum Acceleration (m/s2)')
-
-
-subplot(3,2,5)
-grid on; hold on; 
-scatter(dat1.acceptedGapDownsample, dat1.minDistance,'k*')
-ylim([0, 13]);
-xlim([0, 9]);
[…]

### 2019-01-28 conference/iv2019/manuscript.tex
final commit before adding in POMPD. This is a good safety point if i run out of time.

@@
 
-To simulate random pedestrian crossing events, a pedestrian is spawned in an area of influence on the sidewalk.
+To simulate random pedestrian crossing events, a pedestrian is spawned  on the sidewalk near the entrance to the crosswalk.
 The pedestrian accepts a randomly sized time gap in the traffic stream to cross the road at constant speed $\dot{x}_p$.
+While pedestrian speeds vary both according to demographic factors and within an individual crossing attempt \cite{Feliciani2017}, the primary focus of this study is varying the gap acceptance, and a constant velocity for all trials is therefore assumed.
+
 The size of the time gap is drawn from a normal distribution with mean $\mu_{gap}$ and standard deviation $\sigma_{gap}$.
@@
 \begin{table}[h]
-\footnotesize
 \begin{center}
@@
 Mean accepted gap & $\mu_{gap}$ & 4.0 & sec\\\hline
+Safety offset       & $\Delta$ & 5.0 & m \\
 Speed feedback gain & $k_s$ & 2.0 & 1 / sec\\
@@
 \includegraphics[width=3.5in]{figures/velPlot.eps}
-\caption{Simulation results for case where pedestrian enters the crosswalk on the same side of the road as the vehicle.}
+\caption{Simulation results for case where pedestrian enters the crosswalk on the same side of the road as the vehicle.
+Plots (a) and (b) show the closest distance between the vehicle and pedestrian for each crossing trial, (c) and (d) plots the average vehicle velocity for each simulated crossing, and (e) and (f) plots the peak acceleration / deceleration magnitude for each trial.}
 \label{fig:velPlot1}
@@
 \includegraphics[width=3.5in]{figures/velPlot2.eps}
-\caption{Simulation results for case where pedestrian enters the crosswalk on the opposite side of the road as the vehicle.}
+\caption{Simulation results for case where pedestrian enters the crosswalk on the opposite side of the road as the vehicle.
+Plots (a) and (b) show the closest distance between the vehicle and pedestrian for each crossing trial, (c) and (d) plots the average vehicle velocity for each simulated crossing, and (e) and (f) plots the peak acceleration / deceleration magnitude for each trial.}
 \label{fig:velPlot2}
@@
 \subsubsection{Smoothness}
-The requirement of keeping a low acceleration where possible is validated through (e) and (f) of Figures~\ref{fig:velPlot1} and \ref{fig:velPlot2}.
+The requirement of keeping a low acceleration magntude where possible is validated through (e) and (f) of Figures~\ref{fig:velPlot1} and \ref{fig:velPlot2}, which plots the peak acceleration / deceleration magnitude for each simulated crossing.
 The vehicle is able to maintain a peak acceleration of 2 $\mathrm{m/s^2}$ for the vast majority of simulated trials, particularly when the pedestrian crosses from the opposing side of the road or when the vehicle is in the second to right line.
@@
 Number of lanes & $n$ & 2 & - \\
-Speed feedback gain & $k_s$ & 0.0 & 1 / sec\\
+Safety offset   & $\Delta$ & 5 & m \\
+Speed feedback gain & $k_s$ & 1.0 & 1 / sec\\
 Brake time delay    & $t_\mathrm{delay}$ & 0.5 & sec \\ 
@@
 Several key differences to highlight between the simulations and the experiments are that the number of lanes is decreased from 4 to 2 and that the speed limit is increased to 7 m/s to match the parameters of the experimental intersection.
-Also note that the speed tracking feedback is disabled as it triggered fault conditions in Hyundai's low-level controller.
 The experiment consisted of six trials intended to show operation of the four discrete modes of the controller.
@@
 Figure \ref{fig:expPlot}.
-In Trial 1, the virtual pedestrian enters when the vehicle is four seconds from intersecting, giving the vehicle enough time to yield at a deceleration of 2 $\mathrm{m/s^2}$.
+In Trial 1, the virtual pedestrian accepts a gap of four seconds, giving the vehicle enough time to yield at a deceleration of 2 $\mathrm{m/s^2}$.
 Note that the low level acceleration controller exhibits a large time delay between commanded and actual acceleration.
-Because this is accounted for in Algorithm 1, the vehicle is able to brake early and the vehicle is nearly stopped 1 meter ahead of the desired stop point before the pedestrian exits the crosswalk and the vehicle speeds back up.
-This 1 meter overshoot, while undesirable, is realistically acceptable given the desired stopping point $d=0$ is set to be five meters before the crosswalk actually begins.
+Because this is accounted for in Algorithm 1, the vehicle is able to brake early and the vehicle is stopped right at $d=0$ before the pedestrian exits the crosswalk and the vehicle speeds back up.
 
@@
 \caption{Experimental velocity and acceleration plotted vs distance to crosswalk.
-Note that the sign of $d$ is opposite that of Fig.~\ref{fig:diagram1}.
-Green circles mark the point where the pedestrian first enters crosswalk.
-Red circles mark the point where the pedestrian leaves the crosswalk.
-Note that for Trial 3, the green circle is off the page at -40 meters.}
+Green lines mark the point where the pedestrian first enters crosswalk.
+Black lines mark the point where the pedestrian leaves the crosswalk.
+Note that for Trial 3, the green line is off the page at 40 meters.}
 \label{fig:expPlot}
@@
 Given the brake delay in the experimental vehicle, significant overshoot occurs, and the vehicle stops nearly 2.5 meters ahead of the desired stopping location.
-Unlike in the yielding mode, the brake delay can not be compensated for because the vehicle must brake suddenly.
+However, given that the desired stop location is 5 meters ahead of the crosswalk, this is sufficient to avoid a collision.
+Note that unlike in the yielding mode, the brake delay can not be compensated for because the pedestrian enters unexpectedly.
 

### 2019-01-28 conference/iv2019/plotScripts/expPlot.m

@@ -1,13 +1,19 @@
 clear all; close all; clc;
 
 %% load bags
-t1 = load('/home/nkapania/Desktop/190110NRKbags/trial1.mat');
-t2 = load('/home/nkapania/Desktop/190110NRKbags/trial2.mat');
-t3 = load('/home/nkapania/Desktop/190110NRKbags/trial3.mat');
-t4 = load('/home/nkapania/Desktop/190110NRKbags/trial4.mat');
-t5 = load('/home/nkapania/Desktop/190110NRKbags/trial5.mat');
-t6 = load('/home/nkapania/Desktop/190110NRKbags/trial6.mat');
+% t1 = load('/home/nkapania/Desktop/190110NRKbags/trial1.mat');
+% t2 = load('/home/nkapania/Desktop/190110NRKbags/trial2.mat');
+% t3 = load('/home/nkapania/Desktop/190110NRKbags/trial3.mat');
+% t4 = load('/home/nkapania/Desktop/190110NRKbags/trial4.mat');
+% t5 = load('/home/nkapania/Desktop/190110NRKbags/trial5.mat');
+% t6 = load('/home/nkapania/Desktop/190110NRKbags/trial6.mat');
 
+t1 = load('/home/nkapania/Desktop/190128NRKbags/out1.mat');
+t2 = load('/home/nkapania/Desktop/190110NRKbags/trial2.mat');
+t3 = load('/home/nkapania/Desktop/190128NRKbags/out3.mat');
+t4 = load('/home/nkapania/Desktop/190128NRKbags/out4.mat');
+t5 = load('/home/nkapania/Desktop/190128NRKbags/out5.mat');
+t6 = load('/home/nkapania/Desktop/190128NRKbags/out6.mat');
 
 %% create downsampled versions of actual acceleration
 N1 = length(t1.t);
@@ -39,87 +45,156 @@ ind6 = intersect(find(t6.distance_m > -5), find(t6.distance_m < 35));
 
 offset = 5.5;
 
+
 subplot(6, 2, 1)
-plot(-t1.distance_m(ind1)+offset, t1.velocity_mps(ind1),'LineWidth',2,'FontSize',12)
-title('Velocity m/s')
-ylabel('Trial 1')
[…]

### 2019-01-30 conference/iv2019/manuscript.tex
now ready for read through

@@
 By modeling pedestrian behavior through the concept of gap acceptance, the authors show that a hybrid controller with just four distinct modes allows an autonomous vehicle to successfully interact with a pedestrian across a continuous spectrum of possible crosswalk entry behaviors.
-The controller is validated through extensive simulation, and experimental results are provided on a Hyundai Genesis vehicle for a virtual pedestrian.
+The controller is validated through extensive simulation and compared to an alternate POMDP solution, and experimental results are provided on a Hyundai Genesis vehicle for a virtual pedestrian.
 
@@
 What is also missing is analysis of how this controller should behave across multiple traffic scenarios - for example, the navigation problem is different for the pedestrian crossing on the opposing side of traffic, and also depends on which particular lane the vehicle is in.
+To the authors knowledge, there are also no studies that explicitly compare different approaches in order to provide a comparision of alternative solution methods.
 
-This paper aims to address this gap by developing a hybrid control architecture that accounts for several distinct pedestrian modes of behavior at an unsignalized crosswalk.
+This paper aims to address these gap by developing a hybrid control architecture that accounts for several distinct pedestrian modes of behavior at an unsignalized crosswalk.
 An unsignalized crosswalk is chosen as it is the trickiest crosswalk for a pedestrian and vehicle to navigate, as there is no explicit declaration of whose turn it is cross.
@@
 Moreover, this is possible with just four distinct controller modes.
+Additionally, this paper provides a simulated comparison between the hybrid controller and the solution method proposed by Thornton \cite{Thornton2018}, citing advantages and disadvantages of each method.
 Finally, experimental results are shown on a real vehicle to demonstrate the feasibility of the proposed controller in real-time.
@@
 
-\section{Evaluation Methodology and Simulation Results}
+\section{Evaluation Methodology and Comparison with POMDP}
 
@@
 \ref{fig:simFramework}) developed in Python.
-The simulation environment is built on the ROS robotics interface to enable easy porting of the control architecture to the experimental vehicle in Section \ref{sec:expres}.
+The simulation environment is built on the ROS robotics interface to enable easy porting of the control architecture to the experimental vehicle in Section \ref{sec:expres}, and for easy comparison to the method provided by Thornton \cite{Thornton2018}.
 The simulation includes features such as customizable lanes, the pedestrian crossing on the right or the left, and animation capabilities.
@@
 
-\subsection{Simulation Results}
+\subsection{Alternate Approach - POMDP}
+The method proposed by Thornton \cite{Thornton2018} models the unsignalized crosswalk problem as a Partially 
+Observable Markov Decision Process (POMDP) \cite{Kochenderfer}. The relevant discretized state variable $x$ is given by
+
+\begin{equation}
+x_t = [\dot{d}_t \hspace{2mm} c_t \hspace{2mm} d_t \hspace{2mm} p_t \hspace{2mm} \ddot{d}_{t-1}]^T
+\end{equation}
+
+Where $c_t$ is a Boolean variable that is true when the pedestrian is in the crosswalk, $\ddot{d}_{t-1}$ is the previous acceleration command, and $p_t$ is the pedestrian posture, which can be \textit{distracted}, \textit{walking}, or \textit{stopped}.
+Using a point mass model for the vehicle state transition dynamics and empirical data for the $c_t$dynamics, the authors compute a closed loop policy $\ddot{d}_t = \pi(x_t)$ using the QMDP solver \cite{QMDP}.
+The reward function that is optimized is given by:
+
+\begin{equation}
+r(x_t, \ddot{d}_t) = g_\mathrm{legality} + g_\mathrm{safety} + g_\mathrm{efficient} + g_\mathrm{smooth}
+\end{equation}
+
+Where $g_\mathrm{legality}$ and $g_\mathrm{safety}$ are terms encouraging the vehicle to stop when the pedestrian is in the crosswalk, $g_\mathrm{efficient}$ is a term encouraging high velocity when there is no pedestrian, and $g_\mathrm{smooth}$ encourages the vehicle to avoid abrupt changes in acceleration.
+Note that for our implementation of the method in \cite{Thornton2018}, we assumed $p_t$ was set to \textit{stopped}.
+
+\section{Simulation Results}
 
@@
 The left column of the figure is for the case where the vehicle is on the right-most lane of a four way road (marked ``Lane A"), while the right column of the figure is for the case where the vehicle is in the second to right lane (Lane B).
-Every marker in the figure represents one of 750 simulated vehicle crossings.
+Every marker in the figure represents one of 1500 simulated vehicle crossings, 750 for each method.
 
@@
 \centering
-\includegraphics[width=3.5in]{figures/velPlot.eps}
+\includegraphics[width=3.5in]{figures/velPlotBoth.eps}
 \caption{Simulation results for case where pedestrian enters the crosswalk on the same side of the road as the vehicle.
@@
 \centering
-\includegraphics[width=3.5in]{figures/velPlot2.eps}
+\includegraphics[width=3.5in]{figures/velPlot2Both.eps}
 \caption{Simulation results for case where pedestrian enters the crosswalk on the opposite side of the road as the vehicle.
@@
 
-\subsubsection{Safety and Legality}
+\subsection{Safety and Legality}
 
@@
 
-\subsubsection{Efficiency}
+\subsubsection{Comparison with POMDP}
+
+In terms of safety and legality, both the POMDP and hybrid controller maintain safe distances from the pedestrian across all trials.
+For the hybrid controller, this is because the need to stop at $d=0$ is explicitly encoded in the design.
+For the POMDP, this is due to the reward function terms $g_\mathrm{legality}$ and $g_\mathrm{safety}$
+
+\subsection{Efficiency}
 
@@
 \begin{tabular}{cccc}
-Pedestrian Entry & Vehicle Lane & Average Speed (m/s) & Pct. of Baseline \\\hline\hline
+Pedestrian Entry & Vehicle Lane & Hybrid Controller Average Speed (m/s) & POMDP Average Speed \\\hline\hline
 
@@
 
-\subsubsection{Smoothness}
+\subsubsection{Comparison with POMDP}
+The policy from the POMDP displays interesting behavior at low and high pedestrian accepted gaps.
+At high accepted gaps, even though the pedestrian is waiting for the car to pass before crossing, the POMDP still slows the car down, resulting in a slight drop in average velocity.
+This is because the POMDP is accounting for the chance that the pedestrian might actually cross at a very low accepted gap.
+The closed loop POMDP policy does not appear to have knowledge that the vehicle can speed up at the last second.
+Meanwhile, the hybrid control policy has this case built in, and does not need to slow down if the pedestrian has not entered the crosswalk, resulting in an overall efficiency gain.
+
+Another note is that the POMDP only models the pedestrian as being in the crosswalk or out of the crosswalk - this is necessary to limit the size of the discretized state space for the QMDP solver.
+This results in the POMDP having the same control policy across all four simulated cases.
+Because the proposed hybrid controller is continuous, there is no computational burden on accounting for the position and velocity of the pedestrian in the crosswalk.
+This allows for the controller to consider factors such as time advantage, and continue through the crosswalk in Fig.~\ref{fig:velPlot2}.
+
+\subsection{Smoothness}
 The requirement of keeping a low acceleration magntude where possible is validated through (e) and (f) of Figures~\ref{fig:velPlot1} and \ref{fig:velPlot2}, which plots the peak acceleration / deceleration magnitude for each simulated crossing.
@@
 Again, note that this situation occurs relatively rarely, as pedestrians typically accept gaps between 3-6 seconds and are unlikely to risk hazardous behaviors unless distracted.
+
+\subsubsection{Comparison with POMDP}
+Both the POMDP and the hybrid controller must decelerate at a high magnitude when a pedestrian accepts a gap between 1.75-2.5 seconds.
+One relative advantage of the POMDP is that it is able to decelerate at rates below $a_\mathrm{cmf}$ - at pedestrian accepted gaps of 5-6 seconds, the vehicle decelerates early and at a very low rate in order to avoid excessive jerk.
+However, by using the same policy for all four simulated scenarios and by not considering the ability to speed through the intersection, the POMDP also tends to decelerate when driving through the intersection would have sufficed.
 

### 2019-01-30 conference/iv2019/plotScripts/velocityPlot.m

@@ -1,8 +1,8 @@
 clear all; close all; clc;
-dat1 =load('/home/nkapania/catkin_ws/src/harpagornis/bags/out4.mat');
+dat1 =load('/home/nkapania/catkin_ws/src/harpagornis/bags/out1.mat');
 dat1.lane = 3;
 
-dat2 =load('/home/nkapania/catkin_ws/src/harpagornis/bags/out3.mat');
+dat2 =load('/home/nkapania/catkin_ws/src/harpagornis/bags/out2.mat');
 dat2.lane = 2;
 
 laneWidth = 3.7;

### 2019-01-30 conference/iv2019/manuscript.tex
making borrelli edits

@@
 \title{\LARGE \bf
-A Hybrid Control Framework for Autonomous Vehicles at Uncontrolled Intersections}
+A Hybrid Control Design for Autonomous Vehicles at Uncontrolled Intersections}
 
-\author{Nitin R. Kapania$^{1}$, Vijay Govindarajan$^{2}$, Francesco Borrelli$^{2}$, J Christian Gerdes$^{1}$% <-this % stops a space
+\author{Nitin R. Kapania$^{1}$, Vijay Govindarajan$^{2}$, Francesco Borrelli$^{2}$% <-this % stops a space
 \thanks{*This work was supported by the Hyundai Center of Excellence}% <-this % stops a space
@@
 By modeling pedestrian behavior through the concept of gap acceptance, the authors show that a hybrid controller with just four distinct modes allows an autonomous vehicle to successfully interact with a pedestrian across a continuous spectrum of possible crosswalk entry behaviors.
-The controller is validated through extensive simulation and compared to an alternate POMDP solution, and experimental results are provided on a Hyundai Genesis vehicle for a virtual pedestrian.
+The controller is validated through extensive simulation and compared to an alternate POMDP solution and experimental results are provided on a research vehicle for a virtual pedestrian.
 
@@
 In general, interactions with pedestrians are complex, even for experienced human drivers.
-Issues such as lack of visibility, improper communication, poorly marked roads, and distraction on both the driver or pedestrian side can lead to accidents and fatalities.
 From 2015-2016, pedestrian fatalaties increased by 9\% to 5987, representing the highest number since 1990, and also representing 16\% of all automotive fatalities \cite{HighwayTrafficSafetyAdministration2016}.
@@
 
-The need to further understand pedestrian behavior for autonomous driving has created a growing body of interdisciplinary research.
-One branch of research has focused on modeling pedestrian behavior given various sensor inputs.
+The need to further understand pedestrian behavior for autonomous driving has created a signficant body of research focused on modeling pedestrian behavior given various sensor inputs.
 Keller et al \cite{Keller2014} presented a study on pedestrian path prediction and action classification (e.g.
@@
 waiting) using Gaussian process dynamical models and trajectory matching from optical data.
-Several other techniques for pedestrian trajectory prediction have been proposed, including LQR \cite{Batkovic}, set-based reachability analysis \cite{Koschi2018}, and Markov processes \cite{Karasev2016}
-
-Another branch of literature focuses on pedestrian behavior analysis using empirical data, typically in the form of video analysis.
-While typically studied for purposes of road design, this body of literature holds promising insights for AV designers.
-For example, Schroeder and Rouphail \cite{Schroeder2011} explored factors associated with driver yielding behavior at unsignalized pedestrian crossings.
+Several other techniques for pedestrian trajectory prediction have been proposed, including LQR \cite{Batkovic}, set-based reachability analysis \cite{Koschi2018}, and Markov processes \cite{Karasev2016} Another branch of literature focuses on modeling pedestrian crossing behavior in the form of  video analysis.
+Schroeder and Rouphail \cite{Schroeder2011} explored factors associated with driver yielding behavior at unsignalized pedestrian crossings.
 Using logistic regression, the authors found that drivers are more likely to yield to assertive pedestrians who walk briskly in their approach to a crosswalk.
-Kadali and Perumal \cite{RaghuramKadali2012} studied the ``gap acceptance" behavior of pedestrians at mid-block crosswalks through a video graphic survey, and found that the gap accepted for crossing was explained by factors such as crossing direction, vehicle speed, and pedestrian age.
+Kadali and Perumal \cite{RaghuramKadali2012} studied the ``gap acceptance" behavior of pedestrians at mid-block crosswalks, and found that the gap accepted for crossing was explained by factors such as crossing direction, vehicle speed, and pedestrian age.
 Yannis et al.
 \cite{Yannis2013} and Sun et al.\cite{Sun2002} also found that gap acceptance was influenced by the size of the oncoming vehicle and the presence of other pedestrians.
-Lee and Aty \cite{Lee2005} studied interactions in the form of crashes, and found crashes were linked with higher daily traffic.
 
-Finally, a small but rapidly growing body of literature specifically studies the interaction between pedestrians and autonomous vehicles at crosswalks.
+A small but rapidly growing body of literature specifically studies the interaction between pedestrians and autonomous vehicles at crosswalks.
 An excellent review of these studies was conducted by Rasouli et al.
@@
 Camara et al.
-\cite{Camara2018} attempted to model the natural negotation for priority between a pedestrian and an AV at an intersection using the framework of game theory.
+\cite{Camara2018} attempted to model the natural negotiation for priority between a pedestrian and an AV at an intersection using the framework of game theory.
 Chen et al.
@@
 Significant research has been conducted to understand the likelihood of pedestrian crossing given certain traffic conditions and pedestrian demographics \cite{Schroeder2011} -\cite{Lee2005}, and a small but growing body of literature develops control strategies for pedestrian avoidance \cite{Bandyopadhyay}-\cite{Thornton2018}.
-However, what is missing is a contribution that explicitly  tests whether a proposed control strategy is robust to the variety of pedestrian behaviors that have been observed from experimental studies on real roads.
-What is also missing is analysis of how this controller should behave across multiple traffic scenarios - for example, the navigation problem is different for the pedestrian crossing on the opposing side of traffic, and also depends on which particular lane the vehicle is in.
+However, the literature lacks a contribution that combines the two and explicitly  tests whether a proposed control strategy is robust to the variety of pedestrian behaviors that have been observed from experimental studies on real roads.
+What is also lacking is analysis of how this controller should behave across multiple traffic scenarios - for example, the navigation problem is different for the pedestrian crossing on the opposing side of traffic.
 To the authors knowledge, there are also no studies that explicitly compare different approaches in order to provide a comparision of alternative solution methods.
 
-This paper aims to address these gap by developing a hybrid control architecture that accounts for several distinct pedestrian modes of behavior at an unsignalized crosswalk.
-An unsignalized crosswalk is chosen as it is the trickiest crosswalk for a pedestrian and vehicle to navigate, as there is no explicit declaration of whose turn it is cross.
-Simulations show that the hybrid controller is able to handle a continuous spectrum of pedestrian \textit{gap acceptance} behavior, tolerating a range of highly conservative to highly aggressive pedestrians.
-Moreover, this is possible with just four distinct controller modes.
+This paper aims to address the above issues by developing a hybrid control architecture that accounts for several distinct pedestrian modes of behavior at an unsignalized crosswalk.
+An unsignalized crosswalk is chosen as it is the trickiest crosswalk for a pedestrian and vehicle to navigate, as there is no explicit declaration of whose turn it is to cross.
+Simulations show that the proposed hybrid controller is able to handle a continuous spectrum of pedestrian \textit{gap acceptance} behavior, tolerating a range of highly conservative to highly aggressive pedestrians.
 Additionally, this paper provides a simulated comparison between the hybrid controller and the solution method proposed by Thornton \cite{Thornton2018}, citing advantages and disadvantages of each method.
@@
 In general, right-of-way for uncontrolled intersections is complex.
-For example, nine states and the District of Columbia require motorists to \textit{stop} when approaching a pedestrian in an uncontrolled crosswalk.
-Six states require a motorist to stop when a pedestrian is upon the same half of the roadway or within one lane of the lane that the motorist is traveling in.
-Another nineteen states require a motorist to \textit{yield} when a pedestrian is upon any portion of the roadway, and another 20 states mandate that motorists yield when a pedestrian is upon the same half of the roadway or approaching closely from the opposite side of the roadway \cite{Legislatures2018}.
+For example, nine states and the District of Columbia require motorists to \textit{stop} when approaching a pedestrian in an uncontrolled crosswalk, while another nineteen states only require a motorist to \textit{yield} when a pedestrian is upon any portion of the roadway.
+A full guideline for right-of-way for all fifty states is provided in \cite{Legislatures2018}.
 
@@
 A diagram of the relevant state variables is shown in Fig.~\ref{fig:schematic}.
-We assume that knowledge of distance $d$ to a fixed stopping point 3-5 meters ahead of the crosswalk can be estimated by a perception system, and that the velocity of the vehicle $\dot{d}$ is known as well.
-Furthermore, we also assume the perception system is able to determine the position of the pedestrian $x_p$ within the crosswalk, and is able to determine a velocity estimate $\dot{x}_p$ for the pedestrian as well.
+We assume that knowledge of distance $d(t)$ to a fixed stopping point 3-5 meters ahead of the crosswalk can be estimated by a perception system, and that the vehicle velocity $v(t)$ is known as well.
+Furthermore, we also assume the perception system is able to determine the position of the pedestrian $x_p$ within the crosswalk, and is able to determine a velocity estimate $\dot{x}_p$ for the pedestrian.
 
@@
 
-The relevant control problem is therefore to determine an appropriate longitudinal acceleration command for the vehicle, $u = \ddot{d}$, as a closed-loop function of the vehicle and pedestrian states:
+The relevant control problem is therefore to determine an appropriate longitudinal acceleration command for the vehicle, $u(t) = a(t)$, as a closed-loop function of the vehicle and pedestrian states:
 
 \begin{equation}
-\ddot{d} = f(d, \dot{d}, x_p, \dot{x_p})
+a = f(d, v, x_p, \dot{x_p})
 \end{equation}
+
+% As a result, an algorithm that simply yields every time a pedestrian is near the crosswalk is likely to be too conservative, as there is a likelihood the pedestrian is waiting for a larger gap in the traffic stream to cross or is walking through the sidewalk without intending to cross in the first place.
+
+\subsection{System Requirements}
 
@@
 For example, relevant stakeholders for this problem are the autonomous vehicle occupants, the pedestrian(s), and the other vehicles in the traffic stream.
-
-As a result, an algorithm that simply yields every time a pedestrian is near the crosswalk is likely to be too conservative, as there is a likelihood the pedestrian is waiting for a larger gap in the traffic stream to cross or is walking through the sidewalk without intending to cross in the first place.
-
-\subsection{System Requirements}
 
@@
 \begin{equation}
-\mathrm{gap} = \frac{\mathrm{distance\hspace{1mm}to\hspace{1mm}crosswalk}}{\mathrm{vehicle\hspace{1mm}speed}}
+\mathrm{gap} = \frac{\mathrm{distance\hspace{1mm}to\hspace{1mm}crosswalk}}{\mathrm{vehicle\hspace{1mm}speed}} = \frac{d}{v}
 \end{equation}
@@
 Similar to time to collision (TTC), the safety gap is a measure of how much time there is before the vehicle would enter the crosswalk if it kept its current speed constant.
-When faced with a stream of traffic at a crosswalk, a pedestrian inherently decides how much of a gap to accept before crossing.
+When faced with a stream of traffic at an unsignalized crosswalk, a pedestrian inherently decides how much of a time gap to accept before crossing.
 The average gap acceptance is reported in the literature to be between 3-7 seconds, meaning pedestrians usually do not cross if the vehicle would enter the crosswalk in under three seconds \cite{DiPietroCharlesMandKing1970}, and are very likely to cross when they have more than seven seconds \cite{Schmidt2009}.
@@
 The proposed control architecture is shown in Fig.~\ref{fig:hybridController}, and the algorithm for operation is shown in Algorithm \ref{alg:hybridController}.
-At heart, the control algorithm uses a feedback-feedforward methodology to compute desired acceleration commands $\ddot{d}$, but computes these commands differently depending on which of four discrete modes the controller is in.
+At heart, the control algorithm uses a feedback-feedforward methodology to compute desired acceleration commands, but computes these commands differently depending on which of four discrete modes the controller is in.
 The next section will describe the four modes.
@@
 Symbol & Definition \\\hline\hline
-$d_\mathrm{max}$ & $\frac{\dot{d}^2}{2a_\mathrm{max}}$ \\
-$d_\mathrm{cmf}$ & $\frac{\dot{d}^2}{2a_\mathrm{cmf}}$ \\\hline
+$d_\mathrm{max}$ & $\frac{v^2}{2a_\mathrm{max}}$ \\
+$d_\mathrm{cmf}$ & $\frac{v^2}{2a_\mathrm{cmf}}$ \\\hline
 \end{tabular}
@@
 \begin{algorithm} % enter the algorithm environment
-\caption{Calculate $\ddot{d} = f(d, \dot{d}, x_p, \dot{x}_p)$} % give the algorithm a caption
+\caption{Calculate $a = f(d, v, x_p, \dot{x}_p)$} % give the algorithm a caption
 \label{alg:hybridController} % and a label for \ref{} commands later in the document
@@
 \If {$state = DRIVING$}
-	\State $\ddot{d} \gets k_s(\dot{d}-v_\mathrm{speed limit})$ 
+	\State $a \gets k_s(v-v_\mathrm{speed limit})$ 
 	\If{$d > 0 \hspace{1mm} \mathbf{and} \hspace{1mm} \mathrm{inCrosswalk}(x_p)$}
-		\If{$\mathrm{timeAdvantage}(x_p, \dot{x_p}, d, \dot{d}) > t_{max}$} 
+		\If{$\mathrm{timeAdvantage}(x_p, \dot{x_p}, d, v) > t_{max}$} 
 		\State $state \leftarrow DRIVING$
@@
 
-	% \ElsIf{$d < \frac{\dot{d}^2}{2a_\mathrm{cmf}} \hspace{1mm} \mathbf{and} \hspace{1mm} d > \frac{\dot{d}^2}{2a_\mathrm{max}}$}
+	% \ElsIf{$d < \frac{v^2}{2a_\mathrm{cmf}} \hspace{1mm} \mathbf{and} \hspace{1mm} d > \frac{v^2}{2a_\mathrm{max}}$}
 
@@
 \If {$state = YIELDING$}
-\If {$d > d_\mathrm{cmf} + t_{delay}\dot{d}$}
-	\State $\ddot{d}_{des} = 0$
-	\State $\dot{d}_{des} = v_\mathrm{speed limit}$
+\If {$d > d_\mathrm{cmf} + t_{delay}v$}
+	\State $a_{des} = 0$
+	\State $v_{des} = v_\mathrm{speed limit}$
 \Else 
-	\State $\ddot{d}_{des} = -a_\mathrm{cmf}$
-	\State $\dot{d}_{des} = \mathrm{getDesiredSpeed}(d, \dot{d})$
+	\State $a_{des} = -a_\mathrm{cmf}$
+	\State $v_{des} = \mathrm{getDesiredSpeed}(d, v)$
 \EndIf 
-	\State $\ddot{d} = \ddot{d}_{des} + k_s(\dot{d} - \dot{d}_{des})$
+	\State $a = a_{des} + k_s(v - v_{des})$
 
@@
 \If {$state = HARD\hspace{1 mm}BRAKING$}
-	\State $\dot{d}_{des} = \mathrm{getDesiredSpeed}(d, \dot{d})$
-	\State $\ddot{d} = -\frac{\dot{d}^2}{2d} + k_s(\dot{d} - \dot{d}_{des})$ 	
+	\State $v_{des} = \mathrm{getDesiredSpeed}(d, v)$
+	\State $a = -\frac{v^2}{2d} + k_s(v - v_{des})$ 	
 	\If {$\textbf{not} \hspace{1 mm} \mathrm{inCrosswalk}(x_p)$} \State $state \leftarrow DRIVING$
@@
 \If {$state = SPEED\hspace{1mm}UP$}
-	\State $\ddot{d} = a_\mathrm{cmf}$ 	
+	\State $a = a_\mathrm{cmf}$ 	
 	\If {$\textbf{not} \hspace{1mm} \mathrm{inCrosswalk}(x_p) \vee d < 0$} \State $state \leftarrow DRIVING$
@@
 \begin{equation}
-	\ddot{d} = k_s(\dot{d} - v_\mathrm{speedlimit})
+	a = k_s(v - v_\mathrm{speedlimit})
 \end{equation}
@@
 \begin{equation}
-t_\mathrm{adv} = \frac{d}{\dot{d}} - \frac{x_v - x_p}{\dot{x_p}}
+t_\mathrm{adv} = \frac{x_v - x_p}{\dot{x_p}} - \frac{d}{v}
 \end{equation}
@@
 \begin{equation}
-d_{brake} = \frac{\dot{d}^2}{2a_\mathrm{cmf}}
+d_{brake} = \frac{v^2}{2a_\mathrm{cmf}}
 \end{equation}
@@
 \begin{equation}
-\frac{\dot{d}^2}{2a_{max}} < d < \frac{\dot{d}^2}{2a_\mathrm{cmf}}
+\frac{v^2}{2a_\mathrm{max}} < d < \frac{v^2}{2a_\mathrm{cmf}}
 \end{equation}
 
-Where $a_{max}$ is the largest deceleration magnitude allowed by the tire-road friction.
+Where $a_\mathrm{max}$ is the largest deceleration magnitude allowed by the tire-road friction.
 
@@
 It makes more sense in this condition for the vehicle to speed up and exit the crosswalk quickly.
-As a result, if $d < \frac{\dot{d}}{2a_{max}}$, the vehicle speeds up.
+As a result, if $d < \frac{v}{2a_\mathrm{max}}$, the vehicle speeds up.
 
@@
 In the yielding state, the vehicle follows a different version of the feedback-feedforward dynamics.
-The vehicle first follows the speed limit until it reaches the critical value of $\frac{\dot{d}^2}{2a_\mathrm{cmf}}$.
-In practice, an additional term $t_\mathrm{delay}\dot{d}$ is added to compensate for the delay in brake communication.
+The vehicle first follows the speed limit until it reaches the critical value of $\frac{v^2}{2a_\mathrm{cmf}}$.
+In practice, an additional term $t_\mathrm{delay}v$ is added to compensate for the delay in brake communication.
 
-At the critical distance, the vehicle decelerates at the uniform yielding deceleration of $-a_\mathrm{cmf}$, with an additional feedback term $k_s(\dot{d} - \dot{d}_\mathrm{des}$), where $\dot{d}_\mathrm{des}$ is the desired velocity.
+At the critical distance, the vehicle decelerates at the uniform yielding deceleration of $-a_\mathrm{cmf}$, with an additional feedback term $k_s(v - v_\mathrm{des}$), where $v_\mathrm{des}$ is the desired velocity.
 The feedback term helps bring the vehicle speed to 0 at the desired stopping point.
 
-The desired velocity profile $\dot{d}_\mathrm{des}(d)$ for a vehicle decelerating at constant acceleration is given by:
+The desired velocity profile $v_\mathrm{des}(d)$ for a vehicle decelerating at constant acceleration is given by:
 
 \begin{equation}
-\dot{d}_\mathrm{des}(d) = \sqrt{2a_\mathrm{cmf}(d-d_o) + \dot{d}_o^2}
+v_\mathrm{des}(d) = \sqrt{2a_\mathrm{cmf}(d-d_o) + v_o^2}
 \end{equation}
 
-Where $d_o$ and $\dot{d}_o$ are the values of $d$ and $\dot{d}$ when the controller first enters the yielding state.
-Inspection shows that the desired velocity is $\dot{d}_o$ when $d = d_o$ and is 0 when $d = 0$ (recall that we enter the yielding state when $d_o = \frac{\dot{d_o}^2}{2a_\mathrm{cmf}}$).
+Where $d_o$ and $v_o$ are the values of $d$ and $v$ when the controller first enters the yielding state.
+Inspection shows that the desired velocity is $v_o$ when $d = d_o$ and is 0 when $d = 0$ (recall that we enter the yielding state when $d_o = \frac{\dot{d_o}^2}{2a_\mathrm{cmf}}$).
 
@@
 
-In the hard braking state, the vehicle follows a similar feedback-feedforward algorithm, but the desired feedforward deceleration in this case is given by $\ddot{d} = -\frac{\dot{d}^2}{2d}$.
-Integrating to find the desired speed profile $\dot{d}_\mathrm{des}(d)$ yields the following deceleration:
+In the hard braking state, the vehicle follows a similar feedback-feedforward algorithm, but the desired feedforward deceleration in this case is given by $a = -\frac{v^2}{2d}$.
+Integrating to find the desired speed profile $v_\mathrm{des}(d)$ yields the following velocity profile:
 
 \begin{equation}
-\dot{d}_\mathrm{des}(d) = \frac{\dot{d}_o}{\sqrt{d_o}}\sqrt{d}
+v_\mathrm{des}(d) = \frac{v_o}{\sqrt{d_o}}\sqrt{d}
 \end{equation}
 
-Where $d_o$ and $\dot{d}_o$ are the values of $d$ and $\dot{d}$ when the controller first enters the braking state.
+Where $d_o$ and $v_o$ are the values of $d$ and $v$ when the controller first enters the braking state.
 Again, the controller exits the braking mode and returns to the driving mode once the pedestrian clears the crosswalk.
@@
 
-In the speed up state, there is no need to follow an exact speed profile as the vehicle is merely trying to exit the crosswalk area slightly faster.
-In this case, the commanded acceleration is $\ddot{d} = a_\mathrm{cmf}$ until the pedestrian clears the crosswalk or once the vehicle passes the stopping location (i.e.
+In the speed up state, there is no need to follow an exact speed profile as the vehicle is trying to exit the crosswalk area slightly faster.
+In this case, the commanded acceleration is $a = a_\mathrm{cmf}$ until the pedestrian clears the crosswalk or once the vehicle passes the stopping location (i.e.
 $d< 0$).
@@
 \ref{fig:simFramework}) developed in Python.
-The simulation environment is built on the ROS robotics interface to enable easy porting of the control architecture to the experimental vehicle in Section \ref{sec:expres}, and for easy comparison to the method provided by Thornton \cite{Thornton2018}.
-The simulation includes features such as customizable lanes, the pedestrian crossing on the right or the left, and animation capabilities.
+The simulation environment is built on the ROS robotics interface to enable easy porting of the control architecture to the experimental vehicle in Section \ref{sec:expres}, and for ease of comparison to the method provided by Thornton \cite{Thornton2018}.
 
@@
 \subsection{Alternate Approach - POMDP}
-The method proposed by Thornton \cite{Thornton2018} models the unsignalized crosswalk problem as a Partially 
-Observable Markov Decision Process (POMDP) \cite{Kochenderfer}. The relevant discretized state variable $x$ is given by
+The method proposed by Thornton \cite{Thornton2018} offers a point of comparison to the proposed algorithm, and models the unsignalized crosswalk problem as a Partially 
+Observable Markov Decision Process (POMDP) \cite{Kochenderfer2015}. The relevant discretized state variable $x$ is given by
 
 \begin{equation}
-x_t = [\dot{d}_t \hspace{2mm} c_t \hspace{2mm} d_t \hspace{2mm} p_t \hspace{2mm} \ddot{d}_{t-1}]^T
+x_t = [v_t \hspace{2mm} c_t \hspace{2mm} d_t \hspace{2mm} p_t \hspace{2mm} a_{t-1}]^T
 \end{equation}
 
-Where $c_t$ is a Boolean variable that is true when the pedestrian is in the crosswalk, $\ddot{d}_{t-1}$ is the previous acceleration command, and $p_t$ is the pedestrian posture, which can be \textit{distracted}, \textit{walking}, or \textit{stopped}.
-Using a point mass model for the vehicle state transition dynamics and empirical data for the $c_t$dynamics, the authors compute a closed loop policy $\ddot{d}_t = \pi(x_t)$ using the QMDP solver \cite{QMDP}.
+Where $c_t$ is a Boolean variable that is true when the pedestrian is in the crosswalk, $a_{t-1}$ is the previous acceleration command, and $p_t$ is the pedestrian posture, which can be \textit{distracted}, \textit{walking}, or \textit{stopped}.
+Using a point mass model for the vehicle state transition dynamics and empirical data for the $c_t$ dynamics, the authors compute a closed loop policy $a_t = \pi(x_t)$ using the QMDP solver \cite{Kochenderfer2015}.
 The reward function that is optimized is given by:
@@
 \begin{equation}
-r(x_t, \ddot{d}_t) = g_\mathrm{legality} + g_\mathrm{safety} + g_\mathrm{efficient} + g_\mathrm{smooth}
+r(x_t, a_t) = g_\mathrm{legality} + g_\mathrm{safety} + g_\mathrm{efficient} + g_\mathrm{smooth}
 \end{equation}
@@
 Figure \ref{fig:velPlot2} shows the same simulation results, but for the case where the pedestrian enters the crosswalk on the opposite side of the road as the vehicle.
-Again, the total number of pedestrian crossing trials conducted was 750, with 375 trials for each side.
+Again, the total number of pedestrian crossing trials conducted was 1500, with 750 trials for each method.
 
@@
 This is required in order to meet the collision avoidance constraint, which naturally takes priority.
-Again, note that this situation occurs relatively rarely, as pedestrians typically accept gaps between 3-6 seconds and are unlikely to risk hazardous behaviors unless distracted.
 
@@
 In Trial 2, the pedestrian enters from the right with just 1 second before collision.
-This experiment, while unlikely in reality, was undertaken to verify the vehicle would not brake, but accelerate through the intersection.
+This experiment was undertaken to verify the vehicle would not brake, but accelerate through the intersection.
 Trial 3 represents a situation where the pedestrian crosses with a large (7 second) gap.

### 2019-01-30 conference/iv2019/manuscript.tex
ready to submit!

@@
 high importance, as every human is a pedestrian at some point of the day.
-This paper considers the interaction of a pedestrian and an autonomous vehicle at a mid-block, uncontrolled intersection where there is ambiguity
+This paper considers the interaction of a pedestrian and an autonomous vehicle at a mid-block, unsignalized intersection where there is ambiguity
 over when the pedestrian should cross and when and how the vehicle should yield.
@@
 Given that every human is a pedestrian at some point during the day, one of the central ways that autonomous vehicles will be evaluated is through their interactions with pedestrians.
-In fact, recent events with industry leader Waymo highlight that AV-pedestrian interaction has still not been perfected \cite{MattRichtelandConorDougherty2015}.
-In general, interactions with pedestrians are complex, even for experienced human drivers.
+Interactions with pedestrians can be complex, even for experienced human drivers.
 From 2015-2016, pedestrian fatalaties increased by 9\% to 5987, representing the highest number since 1990, and also representing 16\% of all automotive fatalities \cite{HighwayTrafficSafetyAdministration2016}.
@@
 The need to further understand pedestrian behavior for autonomous driving has created a signficant body of research focused on modeling pedestrian behavior given various sensor inputs.
-Keller et al \cite{Keller2014} presented a study on pedestrian path prediction and action classification (e.g.
+Keller et al.
+\cite{Keller2014} presented a study on pedestrian path prediction and action classification (e.g.
 crossing vs.
 waiting) using Gaussian process dynamical models and trajectory matching from optical data.
-Several other techniques for pedestrian trajectory prediction have been proposed, including LQR \cite{Batkovic}, set-based reachability analysis \cite{Koschi2018}, and Markov processes \cite{Karasev2016} Another branch of literature focuses on modeling pedestrian crossing behavior in the form of  video analysis.
+Several other techniques for pedestrian trajectory prediction have been proposed, including LQR \cite{Batkovic}, set-based reachability analysis \cite{Koschi2018}, and Markov processes \cite{Karasev2016}.
+Another branch of literature focuses on modeling pedestrian crossing behavior in the form of video analysis.
 Schroeder and Rouphail \cite{Schroeder2011} explored factors associated with driver yielding behavior at unsignalized pedestrian crossings.
@@
 This paper aims to address the above issues by developing a hybrid control architecture that accounts for several distinct pedestrian modes of behavior at an unsignalized crosswalk.
-An unsignalized crosswalk is chosen as it is the trickiest crosswalk for a pedestrian and vehicle to navigate, as there is no explicit declaration of whose turn it is to cross.
 Simulations show that the proposed hybrid controller is able to handle a continuous spectrum of pedestrian \textit{gap acceptance} behavior, tolerating a range of highly conservative to highly aggressive pedestrians.
 Additionally, this paper provides a simulated comparison between the hybrid controller and the solution method proposed by Thornton \cite{Thornton2018}, citing advantages and disadvantages of each method.
-Finally, experimental results are shown on a real vehicle to demonstrate the feasibility of the proposed controller in real-time.
+Finally, experimental results are shown on a real vehicle to demonstrate the feasibility of the proposed controller.
 
@@
 \centering
-\includegraphics[width=3.5in]{figures/example2.eps}
+\includegraphics[width=2.5in]{figures/example2.eps}
 \caption{Top: Description of relevant problem variables.
@@
 A diagram of the relevant state variables is shown in Fig.~\ref{fig:schematic}.
-We assume that knowledge of distance $d(t)$ to a fixed stopping point 3-5 meters ahead of the crosswalk can be estimated by a perception system, and that the vehicle velocity $v(t)$ is known as well.
-Furthermore, we also assume the perception system is able to determine the position of the pedestrian $x_p$ within the crosswalk, and is able to determine a velocity estimate $\dot{x}_p$ for the pedestrian.
+We assume that knowledge of distance $d(t)$ to a fixed stopping point $\Delta$ 3-5 meters ahead of the crosswalk can be estimated by a perception system, and that the vehicle velocity $v(t)$ is known as well.
+We also assume the perception system is able to determine the position of the pedestrian $x_p$ within the crosswalk, and is able to determine a velocity estimate $\dot{x}_p$ for the pedestrian.
+Note that $x_p$ can be negative for the case where the pedestrian is on the sidewalk.
 
@@
 \begin{equation}
-a = f(d, v, x_p, \dot{x_p})
+a = k(d, v, x_p, \dot{x_p})
 \end{equation}
@@
 
-In general, this control problem is difficult as there are multiple stakeholders and therefore multiple competing objectives that must be considered.
-For example, relevant stakeholders for this problem are the autonomous vehicle occupants, the pedestrian(s), and the other vehicles in the traffic stream.
+A difficulty with this control problem is that there are multiple stakeholders and therefore multiple competing objectives that must be considered.
+Relevant stakeholders for this problem are the autonomous vehicle occupants, the pedestrian(s), and the other vehicles in the traffic stream.
 
@@
 
-Similar to time to collision (TTC), the safety gap is a measure of how much time there is before the vehicle would enter the crosswalk if it kept its current speed constant.
+The accepted gap is a measure of how much time there is before the vehicle would enter the crosswalk if it kept its current speed constant.
 When faced with a stream of traffic at an unsignalized crosswalk, a pedestrian inherently decides how much of a time gap to accept before crossing.
@@
 
-The proposed control architecture is shown in Fig.~\ref{fig:hybridController}, and the algorithm for operation is shown in Algorithm \ref{alg:hybridController}.
-At heart, the control algorithm uses a feedback-feedforward methodology to compute desired acceleration commands, but computes these commands differently depending on which of four discrete modes the controller is in.
+The proposed hybrid control architecture is shown in Fig.~\ref{fig:hybridController}, and the algorithm for operation is shown in Algorithm \ref{alg:hybridController}.
+At its core, the control algorithm uses a feedback-feedforward methodology to compute desired acceleration commands, but computes these commands differently depending on which of four discrete modes the controller is in.
 The next section will describe the four modes.
@@
 \begin{algorithm} % enter the algorithm environment
-\caption{Calculate $a = f(d, v, x_p, \dot{x}_p)$} % give the algorithm a caption
+\caption{Calculate $a = k(d, v, x_p, \dot{x}_p)$} % give the algorithm a caption
 \label{alg:hybridController} % and a label for \ref{} commands later in the document
@@
 	\If{$d > 0 \hspace{1mm} \mathbf{and} \hspace{1mm} \mathrm{inCrosswalk}(x_p)$}
-		\If{$\mathrm{timeAdvantage}(x_p, \dot{x_p}, d, v) > t_{max}$} 
+		\If{$\mathrm{timeAdvantage}(x_p, \dot{x_p}, d, v) > \tau_\mathrm{max}$} 
 		\State $state \leftarrow DRIVING$
@@
 In the nominal state, the vehicle attempts to drive through the crosswalk at the speed limit $v_\mathrm{speedlimit}$.
-A simple proportional speed control is applied to keep the vehicle at the speed limit, achieving design constraint 2.a:
+A proportional speed control is applied to keep the vehicle at the speed limit, achieving design constraint 2.a:
 
@@
 The controller remains in the driving mode unless the pedestrian begins to enter the crosswalk before the vehicle has passed.
-In this case, we mathematically define ``beginning to enter" as the instant the pedestrian begins moving towards the crosswalk, even if the pedestrian is currently on the sidewalk:
+We mathematically define ``beginning to enter" as the instant the pedestrian begins moving towards the crosswalk, even if the pedestrian is currently on the sidewalk:
 
@@
  Where $x_F$ is the end of the crosswalk area of interest.
-Note that according to design constraint 1.b, the definition of $x_F$ will vary according to which US state the vehicle is in.
+According to design constraint 1.b, the definition of $x_F$ will vary according to which US state the vehicle is in.
 For example, in California, $x_F$ would be at the end of the crosswalk, while in Louisiana, $x_F$ would just be the same half of the crosswalk as the vehicle \cite{Legislatures2018}.
-
 Once the pedestrian enters the crosswalk, the algorithm decides which mode to enter next.
@@
 
-Frequently the vehicle will pass through the intersection well before the pedestrian will reach the vehicle's lane.
+Frequently the vehicle will pass through the crosswalk well before the pedestrian will reach the vehicle's lane.
 Assuming the vehicle is operating in a US state where stopping is not explicitly required, design constraint 2.a suggests it is desirable for the vehicle to continue through the intersection.
 
-This can be formalized by calculating the \textit{time advantage} $t_\mathrm{adv}$, also known as the \textit{time to collision} as follows:
+This can be formalized by calculating the \textit{time advantage} $t_\mathrm{adv}$ \cite{Chen} as follows:
 
@@
 Where $x_v$ is the $x$ position of the vehicle in the crosswalk.
-Algorithm 1 explicitly allows the vehicle to continue driving through the intersection as long as $t_\mathrm{adv}$ exceeds a specified threshold $t_\mathrm{max}$.
+Algorithm 1 explicitly allows the vehicle to continue driving through the intersection as long as $t_\mathrm{adv}$ exceeds a specified threshold $\tau_\mathrm{max}$.
 
@@
 
-In the unlikely event a pedestrian begins to cross when the time gap is low (e.g.
-under 3 seconds), the vehicle will need to prioritize design constraint 1.a and come to a stop at a deceleration higher than $-a_\mathrm{cmf}$.
+In the unlikely event the pedestrian crosses when the time gap is low, the vehicle will need to prioritize design constraint 1.a and come to a stop at a deceleration higher than $-a_\mathrm{cmf}$.
 This occurs when the following conditions holds:
@@
 As a final condition, consider the case where the pedestrian enters the crosswalk just as the vehicle is crossing $d = 0$.
-In this case, it makes little sense to decelerate, as the vehicle has insufficient space to stop before the crosswalk and will risk being rear-ended or stopping in the crosswalk.
+In this case, the vehicle has insufficient space to stop before the crosswalk and will risk being rear-ended or stopping in the crosswalk.
 It makes more sense in this condition for the vehicle to speed up and exit the crosswalk quickly.
@@
 
-Where $d_o$ and $v_o$ are the values of $d$ and $v$ when the controller first enters the yielding state.
-Inspection shows that the desired velocity is $v_o$ when $d = d_o$ and is 0 when $d = 0$ (recall that we enter the yielding state when $d_o = \frac{\dot{d_o}^2}{2a_\mathrm{cmf}}$).
+Where $d_o$ and $v_o$ are the values of $d$ and $v$ when the controller first enters the yielding state. Inspection shows that the desired velocity is $v_o$ when $d = d_o$ and is 0 when $d = 0$. 
+% (recall that we enter the yielding state when $d_o = \frac{\dot{d_o}^2}{2a_\mathrm{cmf}}$).
 
-The controller exits the yielding mode and returns to the driving mode once the pedestrian clears the crosswalk (e.g. once $x > x_F$ in (\ref{eq:crossing})).
+The controller exits the yielding mode and returns to the driving mode once the pedestrian clears the crosswalk.
 
@@
 
-In order to check performance of the algorithm, both simulation and experimental studies were undertaken.
+To validate the algorithm, both simulation and experimental studies were undertaken.
 The simulation results focused on testing Algorithm 1 against the specified design requirements over many pedestrian crossings.
 
-The algorithm is validated in simulation via a built-from-scratch crosswalk simulation environment (Fig.
-\ref{fig:simFramework}) developed in Python.
+The algorithm is validated in simulation via a built-from-scratch crosswalk simulation environment developed in Python (Fig.~\ref{fig:simFramework}).
 The simulation environment is built on the ROS robotics interface to enable easy porting of the control architecture to the experimental vehicle in Section \ref{sec:expres}, and for ease of comparison to the method provided by Thornton \cite{Thornton2018}.
@@
 \centering
-\includegraphics[width=2.5in]{figures/simFramework.png}
+\includegraphics[width=2.0in]{figures/simFramework.png}
 \caption{Crosswalk simulation environment for algorithm validation. Ego vehicle is shown in black, leading vehicle in traffic stream shown in blue, and pedestrian shown in red.}
@@
 The pedestrian accepts a randomly sized time gap in the traffic stream to cross the road at constant speed $\dot{x}_p$.
-While pedestrian speeds vary both according to demographic factors and within an individual crossing attempt \cite{Feliciani2017}, the primary focus of this study is varying the gap acceptance, and a constant velocity for all trials is therefore assumed.
-
 The size of the time gap is drawn from a normal distribution with mean $\mu_{gap}$ and standard deviation $\sigma_{gap}$.
@@
 \cite{Feliciani2017}.
-
 The ego vehicle's response to the pedestrian is simulated over Algorithm 1 for the parameters shown in Table \ref{tb:params}.
@@
 Comfort acceleration & $a_\mathrm{cmf}$ & 2 & $\mathrm{m/s^2}$ \\
-Maximum time advantage & $t_\mathrm{max}$ & 4 & sec \\
+Maximum time advantage & $\tau_\mathrm{max}$ & 4 & sec \\
 Maximum acceleration & $a_\mathrm{max}$ & 9 & $\mathrm{m/s^2}$ \\\hline
@@
 \subsection{Alternate Approach - POMDP}
-The method proposed by Thornton \cite{Thornton2018} offers a point of comparison to the proposed algorithm, and models the unsignalized crosswalk problem as a Partially 
+The method proposed by Thornton \cite{Thornton2018} offers a comparison to the proposed algorithm and models the unsignalized crosswalk problem as a Partially 
 Observable Markov Decision Process (POMDP) \cite{Kochenderfer2015}. The relevant discretized state variable $x$ is given by
@@
 Where $g_\mathrm{legality}$ and $g_\mathrm{safety}$ are terms encouraging the vehicle to stop when the pedestrian is in the crosswalk, $g_\mathrm{efficient}$ is a term encouraging high velocity when there is no pedestrian, and $g_\mathrm{smooth}$ encourages the vehicle to avoid abrupt changes in acceleration.
-Note that for our implementation of the method in \cite{Thornton2018}, we assumed $p_t$ was set to \textit{stopped}.
+For our implementation of \cite{Thornton2018}, we assumed $p_t$ was set to \textit{stopped}.
 
@@
 Figure \ref{fig:velPlot1} shows simulation results for the case where the pedestrian enters the crosswalk on the same side of the road as the vehicle.
-The left column of the figure is for the case where the vehicle is on the right-most lane of a four way road (marked ``Lane A"), while the right column of the figure is for the case where the vehicle is in the second to right lane (Lane B).
+The vehicle is simulated in both Lane A and Lane B in Fig.~\ref{fig:simFramework}.
 Every marker in the figure represents one of 1500 simulated vehicle crossings, 750 for each method.
+Figure \ref{fig:velPlot2} shows the same simulation results, but for the case where the pedestrian enters the crosswalk on the opposite side of the road as the vehicle.
 
@@
 \end{figure}
-
-Figure \ref{fig:velPlot2} shows the same simulation results, but for the case where the pedestrian enters the crosswalk on the opposite side of the road as the vehicle.
-Again, the total number of pedestrian crossing trials conducted was 1500, with 750 trials for each method.
 
@@
 The riskiest case occurs when the vehicle is in Lane A and the pedestrian crosses from the right side, accepting a gap between 1.25 and 1.75 seconds.
-Note that this is in general unlikely given gap acceptance statistics reported in the literature \cite{Rasouli}\cite{Feliciani2017}, but would represent a safety risk given the vehicle has limited time to react.
+This is unlikely given gap acceptance statistics reported in the literature \cite{Rasouli}\cite{Feliciani2017}, but would represent a safety risk given the vehicle has limited time to react.
 However, since the algorithm recognizes that braking will not clear the crosswalk, the vehicle is able to speed up and maintain at least 2 meters of lateral distance from the pedestrian at all times.
-Note that a better way to confirm robustness of the controller to all reasonable pedestrian behaviors would be through a theoretical verification analysis - this will be the focus of future work.
+Future work will use formal verification methods in order to show safety for all pedestrian accepted gaps.
 
-In terms of the fifty state legality requirement, we conducted our simulation for the twenty states where the vehicle is required to yield, not stop, to a pedestrian in the same side of the crosswalk or approaching from the opposite side.
+% In terms of the fifty state legality requirement, we conducted our simulation for the twenty states where the vehicle is required to yield, not stop, to a pedestrian in the same side of the crosswalk or approaching from the opposite side.
 For the US states where the vehicle is explicitly required to stop, the change to the controller is trivial - the controller may not stay in the driving mode if sufficient time advantage exits (see Algorithm 1).
@@
 
-The requirement to minimize deviation from speed of traffic is validated in (c) and (d) of Figures ~\ref{fig:velPlot1} and ~\ref{fig:velPlot2}, which shows the average velocity of the vehicle for each simulated crossing.
-For high accepted gaps, the pedestrian simply waits for the ego vehicle to exit the crosswalk before beginning to walk, and the algorithm does not need to slow down from the speed of traffic.
+ Figures ~\ref{fig:velPlot1} and ~\ref{fig:velPlot2} show the average velocity of the vehicle for each simulated crossing in (c) and (d).
+For high accepted gaps, the pedestrian waits for the ego vehicle to exit the crosswalk before beginning to walk, and the algorithm does not need to slow down from the speed of traffic.
 For low accepted gaps, the vehicle cannot stop, and must speed up or continue driving through the intersection, also keeping the vehicle speed high.
@@
 The most disruption to traffic flow occurs when the vehicle must slow down significantly or stop, which occurs for accepted gaps between 2-6 seconds.
-In this case, the average velocity increases as the accepted gap increases.
-Larger accepted gaps give the pedestrian more time to walk before the vehicle approaches, meaning they are closer to exiting the crosswalk when the vehicle must yield.
-This means the vehicle does not need to stop for as much time, resulting in higher average velocity.
-
-Note that a special case occurs in Fig.~\ref{fig:velPlot2} where the vehicle is in Lane A and the pedestrian crosses from the opposite side of the road.
+A special case occurs in Fig.~\ref{fig:velPlot2} where the vehicle is in Lane A and the pedestrian crosses from the opposite side of the road.
 Because the vehicle is at a far distance from the pedestrian, the time advantage remains sufficiently high for the vehicle to comfortably continue through the crosswalk without needing to slow down.
-
-Table \ref{tb:speeds} shows a summary of the average speed compared to the baseline speed of 4.4 m/s for each of the four cases.
-
-\begin{table}[h]
-\begin{center}
-\caption{Average Vehicle Speed}\label{tb:speeds}
-\begin{tabular}{cccc}
-Pedestrian Entry & Vehicle Lane & Hybrid Controller Average Speed (m/s) & POMDP Average Speed \\\hline\hline
-
-Right Side & Right-Most & 2.90 & 66\% \\
-Right Side & Second-to-Right & 2.93 & 67\% \\
-Left Side  & Right-Most & 4.4 & 100\% \\
-Left Side  & Second-to-Right & 2.80 & 64\% \\\hline
-
-\end{tabular}
-\end{center}
-\end{table}
 
@@
 The policy from the POMDP displays interesting behavior at low and high pedestrian accepted gaps.
-At high accepted gaps, even though the pedestrian is waiting for the car to pass before crossing, the POMDP still slows the car down, resulting in a slight drop in average velocity.
-This is because the POMDP is accounting for the chance that the pedestrian might actually cross at a very low accepted gap.
-The closed loop POMDP policy does not appear to have knowledge that the vehicle can speed up at the last second.
-Meanwhile, the hybrid control policy has this case built in, and does not need to slow down if the pedestrian has not entered the crosswalk, resulting in an overall efficiency gain.
+At high accepted gaps, even though the pedestrian is waiting for the car to pass before crossing, the POMDP still slows the car down.
+The POMDP, which uses a probabilistic model of pedestrian crossing, is accounting for the chance the pedestrian might cross at a very low accepted gap.
+The POMDP policy does not account for the possibility of the vehicle speeding up at the last second.
+The hybrid control policy has this case built in, and does not need to slow down if the pedestrian has not entered the crosswalk, resulting in an overall efficiency gain.
 
-Another note is that the POMDP only models the pedestrian as being in the crosswalk or out of the crosswalk - this is necessary to limit the size of the discretized state space for the QMDP solver.
+Also note the POMDP models the pedestrian as being in the crosswalk or out of the crosswalk - this is necessary to limit the size of the discretized state space for the QMDP solver.
 This results in the POMDP having the same control policy across all four simulated cases.
 Because the proposed hybrid controller is continuous, there is no computational burden on accounting for the position and velocity of the pedestrian in the crosswalk.
-This allows for the controller to consider factors such as time advantage, and continue through the crosswalk in Fig.~\ref{fig:velPlot2}.
+This allows for the controller to consider factors such as time advantage, and continue through the crosswalk in many cases.
+
+% Table \ref{tb:speeds} shows a summary of the average speed compared to the baseline speed of 4.4 m/s for each of the four cases.
+
+% \begin{table}[h]
+% \begin{center}
+% \caption{Average Vehicle Speed}\label{tb:speeds}
+% \begin{tabular}{cccc}
+% Pedestrian Entry & Vehicle Lane & Hybrid Controller Average Speed (m/s) & POMDP Average Speed \\\hline\hline
+
+% Right Side & Right-Most & 2.90 & 66\% \\
+% Right Side & Second-to-Right & 2.93 & 67\% \\
+% Left Side & Right-Most & 4.4 & 100\% \\
+% Left Side & Second-to-Right & 2.80 & 64\% \\\hline
+
+% \end{tabular}
+% \end{center}
+% \end{table}
 
 \subsection{Smoothness}
-The requirement of keeping a low acceleration magntude where possible is validated through (e) and (f) of Figures~\ref{fig:velPlot1} and \ref{fig:velPlot2}, which plots the peak acceleration / deceleration magnitude for each simulated crossing.
-The vehicle is able to maintain a peak acceleration of 2 $\mathrm{m/s^2}$ for the vast majority of simulated trials, particularly when the pedestrian crosses from the opposing side of the road or when the vehicle is in the second to right line.
+Plots (e) and (f) of Figures~\ref{fig:velPlot1} and \ref{fig:velPlot2} show the peak acceleration / deceleration magnitude for each simulated crossing.
+The hybrid controller maintains a peak acceleration of 2 $\mathrm{m/s^2}$ for the vast majority of simulated trials, particularly when the pedestrian crosses from the opposing side of the road or when the vehicle is in Lane B.
 
@@
 \subsubsection{Comparison with POMDP}
-Both the POMDP and the hybrid controller must decelerate at a high magnitude when a pedestrian accepts a gap between 1.75-2.5 seconds.
-One relative advantage of the POMDP is that it is able to decelerate at rates below $a_\mathrm{cmf}$ - at pedestrian accepted gaps of 5-6 seconds, the vehicle decelerates early and at a very low rate in order to avoid excessive jerk.
-However, by using the same policy for all four simulated scenarios and by not considering the ability to speed through the intersection, the POMDP also tends to decelerate when driving through the intersection would have sufficed.
+ One relative advantage of the POMDP is that it is able to decelerate at rates below $a_\mathrm{cmf}$.
+At pedestrian accepted gaps of 5-6 seconds, the vehicle decelerates early and at a very low rate in order to avoid excessive jerk.
+However, by using the same policy for all four simulated scenarios and by not considering the ability to drive through the intersection, the POMDP also tends to decelerate when driving through the intersection would have sufficed.
 
@@
 \ref{fig:g80}.
-A two lane unsignalized intersection at Richmond Field Station near Berkeley's campus was chosen for the experimental validation (Fig.~\ref{fig:crosswalkpic}).
-In lieu of an actual pedestrian, a pedestrian was simulated virtually using the ROS simulation architecture, with the position of the pedestrian passed to the vehicle controller in real time.
+A two lane unsignalized intersection near the Berkeley campus was chosen for the experimental validation.
+In lieu of an actual pedestrian, a pedestrian was simulated virtually using the ROS simulation architecture, with the position and velocity of the pedestrian passed to the controller.
 Differential GPS was used to record the distance to the crosswalk, and a steering controller from \cite{Kapania2015} was used to autonomously steer the vehicle down the road.
 Relevant parameters of the experiments are shown below in Table \ref{tb:expparams}.
+The experiment consisted of six trials intended to show operation of the four discrete modes of the controller.
+A description of the trial parameters are shown in Table \ref{tb:expparams2}.
 
@@
 \end{table}
-
-Several key differences to highlight between the simulations and the experiments are that the number of lanes is decreased from 4 to 2 and that the speed limit is increased to 7 m/s to match the parameters of the experimental intersection.
-The experiment consisted of six trials intended to show operation of the four discrete modes of the controller.
-A description of the trial parameters are shown in Table \ref{tb:expparams2}.
 
@@
 Figure \ref{fig:expPlot}.
-In Trial 1, the virtual pedestrian accepts a gap of four seconds, giving the vehicle enough time to yield at a deceleration of 2 $\mathrm{m/s^2}$.
-Note that the low level acceleration controller exhibits a large time delay between commanded and actual acceleration.
+In Trial 1, the pedestrian accepts a gap of four seconds, giving the vehicle enough time to yield at a deceleration of 2 $\mathrm{m/s^2}$.
+Note that the low level longitudinal controller exhibits a large time delay between commanded and actual acceleration.
 Because this is accounted for in Algorithm 1, the vehicle is able to brake early and the vehicle is stopped right at $d=0$ before the pedestrian exits the crosswalk and the vehicle speeds back up.
@@
 In Trial 4, the pedestrian exhibits risky behavior, crossing with a 2.5 second gap.
-In this case, the vehicle has enough time to brake, but must decelerate rapidly at 5 $\mathrm{m/s^2}$ in the \textit{hard braking} mode.
+The vehicle has enough time to brake, but must decelerate rapidly at 5 $\mathrm{m/s^2}$ in the \textit{hard braking} mode.
 Given the brake delay in the experimental vehicle, significant overshoot occurs, and the vehicle stops nearly 2.5 meters ahead of the desired stopping location.
 However, given that the desired stop location is 5 meters ahead of the crosswalk, this is sufficient to avoid a collision.
-Note that unlike in the yielding mode, the brake delay can not be compensated for because the pedestrian enters unexpectedly.
 

### 2019-02-13 conference/acc2015/acc2015.tex
converting last two chapters of thesis into journal paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%2345678901234567890123456789012345678901234567890123456789012345678901234567890
%        1         2         3         4         5         6         7         8

\documentclass[letterpaper, 10 pt, conference]{ieeeconf}  % Comment this line out if you need a4paper
%\documentclass[letterpaper, 11 pt, conference, draftcls, onecolumn]{ieeeconf}

%\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4 paper

\IEEEoverridecommandlockouts                              % This command is only needed if 
                                                          % you want to use the \thanks command

\overrideIEEEmargins                                      % Needed to meet printer requirements.

% See the \addtolength command later in the file to balance the column lengths
% on the last page of the document

% The following packages can be found on http:\\www.ctan.org
\usepackage{graphicx} % for pdf, bitmapped graphics files
%\usepackage{epsfig} % for postscript graphics files
%\usepackage{mathptmx} % assumes new font selection scheme installed
%\usepackage{times} % assumes new font selection scheme installed
\usepackage{amsmath} % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed

\bibliographystyle{IEEEtran}
\def\mywidth{8.5cm}

\title{\LARGE \bf
Multiple-Lap Path Tracking for an Autonomous Race Vehicle via Iterative Learning Control }

\author{Nitin R. Kapania $^{1,2}$ and J. Christian Gerdes$^{1}$% <-this % stops a space
\thanks{$^{1}$Department of Mechanical Engineering, Stanford University, Stanford, CA 94305, USA}
\thanks{$^{2}$Corresponding author: {\tt\small nkapania@stanford.edu}}
}

\begin{document}

\maketitle
\thispagestyle{empty}
\pagestyle{empty}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}

Iterative learning control has been successfully used for several decades to improve the performance of control 
systems that perform a single repeated task. Using information from prior control executions, learning controllers gradually 
determine open-loop control inputs whose reference tracking performance can exceed that of traditional feedback-feedforward control algorithms. This paper considers
iterative learning control for a previously unexplored field: autonomous racing. Racecars are driven multiple laps around
the same sequence of turns while operating near the physical limits of tire-road friction, where steering dynamics become highly nonlinear and transient, making
accurate path tracking difficult. However, because the vehicle trajectory is identical for each lap in the case of single-car racing, the nonlinear vehicle dynamics
 and unmodelled road conditions are repeatable and can be accounted for using iterative learning control, provided the tire force limits have not been exceeded.
 This paper describes the design and application of proportional-derivative (PD) and quadratically
optimal (Q-ILC) learning algorithms for multiple-lap path tracking of an autonomous race vehicle. Simulation results are used to tune controller gains and test convergence, and 
experimental results are presented on an Audi TTS race vehicle driving several laps around Thunderhill Raceway in Willows, CA at lateral accelerations of up to 8 $\mathrm{m/s^2}$. Both
control algorithms are able to correct transient path tracking errors and improve the performance provided by a reference feedforward controller.

\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

Iterative learning control (ILC) solves the control problem of tracking a specific reference command that is repeated many times. 
By using tracking error information from prior attempts, ILC techniques are used to gradually determine the open-loop
control inputs that cause the system output to track a desired reference with minimal tracking error.

ILC algorithms work best when the exogenous system signals (i.e. disturbances and reference inputs)
are constant from iteration-to-iteration. As a result, iterative learning controllers have frequently been applied in relatively structured robotics and automation environments, with recent publications 
considering piezolectric positioning \cite{huang}, robotic arm tracking \cite{freeman},  and microdeposition \cite{hoelzle}. However, 
iterative learning control has recently expanded into applications
outside of the traditional automation and process control setting. Chen and Moore \cite{chen} proposed a simple iterative learning scheme in 2006 to improve path-following of a ground vehicle with omni-directional wheels, where
 double integration of the previous feedback input was used to improve the feedforward signal. In 2013, Sun et al. proposed an iterative learning controller for overspeed proctection of high-speed
trains \cite{sun}. Purwin and Andrea synthesized an iterative controller using least-squares methods to aggressively maneuver a quadrotor unmanned aerial vehicle from
one state to another \cite{purwin}. Iterative learning control has also extended into the neuromuscular and biological domains, with Rogers et al. \cite{rogers} developing an ILC algorithm 
for robotically assisted stroke rehabilitation.

This paper presents autonomous race car piloting as a previously unexplored application area for iterative learning control. Race car drivers must drive multiple laps around
the same sequence of turns on a closed track, while operating near the physical limits of tire-road friction to minimize lap times.
 At the limits of tire friction, the steering dynamics associated with lateral  tracking of the desired path become highly nonlinear,
and difficult-to-measure disturbances such as bank, grade and local friction variation of the road surface have a large effect on the transient dynamics of the vehicle. 
These factors make development of a suitable feedforward steering controller challenging. However, because operation of the race vehicle 
occurs over multiple laps, with the reference road curvature unchanging from lap-to-lap, the unknown transient disturbances and
 vehicle dynamics tend be constant from lap to lap and can therefore be accounted for via iterative learning control. A notable exception occurs when the vehicle has significantly
 understeered due to lack of front tire force availability. In this case, additional steering will have no impact on path tracking.

 This paper is further divided as follows. Section II introduces a linear model for the planar vehicle dynamics of a race car following a fixed reference path. Because the transfer function between
the steering wheel input and the vehicle's path deviation is open-loop unstable, a stabilizing lanekeeping controller is added to the steering system and the closed loop dynamics
are represented in the commonly used ``lifted domain". Section III presents a PD-type iterative learning controller with a low-pass filter used to speed
up convergence. Gain tuning and stability at low lateral accelerations are shown using lifted domain techniques, while nonlinear simulations are used to predict a desirable tracking response in the presence of
high lateral acceleration. Section IV presents a quadratically optimal (Q-ILC) iterative learning controller, which has the benefit of explicitly
accounting for changes in vehicle speed along the race track. Section V presents experimental data of both controllers implemented on an Audi TTS race vehicle at combined 
lateral/longitudinal accelerations of up to 8 $\mathrm{m/s^2}$.

\section{Vehicle Dynamics and Problem Overview}
\label{sec:problemDescription}

For this paper, the objective of piloting an autonomous vehicle along a fixed race track in minimum time is divided into separate lateral and longitudinal vehicle control problems. The lateral 
controller uses the the steering wheel input to track a desired ``racing-line", shown in 
Fig.~\ref{problemInfo}a. The racing line is frequently represented by a path curvature function $\kappa(s)$ parametrized by distance along the track (Fig. \ref{problemInfo})b. 
The longitudinal controller tracks a desired speed profile that keeps the vehicle at a specified lateral-longitudinal acceleration magnitude, typically near the limits of tire-road friction (Fig.~\ref{problemInfo})c. 
This paper focuses on learning the desired steer angle command $\delta_L$ over multiple laps in order to accurately track the reference path at high speeds, and we therefore assume minimum time 
velocity and curvature profiles have been computed using methods published in \cite{theodosis}, and that tracking of the velocity profile is handled by a separate controller \cite{mickthesis}.

\begin{figure}[h]
\centering
\includegraphics[width=3in]{figures/problemInfo.png}
\caption{(a) Overhead plot of racing line for Thunderhill Raceway, Willows CA, USA. (b) Racing line represented by curvature along distance traveled. (c) Velocity profile
to keep vehicle at combined lateral/longitudinal acceleration of 8 $\mathrm{m/s^2}$.}
\label{problemInfo}
\end{figure}

\subsection{Lateral Vehicle Dynamics}

Fig.~\ref{fig:bikemodel} shows a schematic of a vehicle following a path
with curvature profile $\kappa(s)$. The lateral deviation of the vehicle from the desired path ($e$) is the measured output for the ILC algorithms.
Since iterative learning controllers determine appropriate \textit{feedforward} control inputs over several iterations, it is necessary for the open-loop dynamics between the control input and
control output to be asymptotally stable. For the case of active steering control, the transfer function between the vehicle steer angle ($\delta$) and the lateral path error $e$ is
characterized by two poles at the origin, requiring the addition of a stabilizing feedback controller.

A lookahead controller provides the stabilizing feedback command for this paper. The ``lookahead error" is defined by

\begin{equation}
	e_\mathrm{LA}=e+x_\mathrm{LA}\Delta\Psi
\end{equation}

Where $\Delta\Psi$ is the vehicle heading error and $x_\mathrm{LA}$ is the lookahead distance, typically 5-20 meters for autonomous driving.  The resulting feedback control law is

\begin{equation}
	\delta_\mathrm{FB} = -k_\mathrm{P}e_\mathrm{LA}
	\label{eqn:lookahead}
\end{equation}

with proportional gain $k_\mathrm{P}$. The control law (\ref{eqn:lookahead}) is a natural extension of potential field lanekeeping, as described by Rossetter et al. in \cite{rossetter2002}, which also provides heuristics for 
selecting $k_\mathrm{P}$ and $x_\mathrm{LA}$. Desirable stability properties over significant tire saturation levels are demonstrated in \cite{talvala}.

With the feedback controller added, closed loop dynamics of the lateral path deviation are dependent on three other states: vehicle sideslip $\beta$, yaw rate $r$ and heading error $\Delta\Psi$. For controller development
and testing, these dynamics are given by the planar bicycle model:

\begin{figure}
\centering
\includegraphics[width=3.5in]{figures/BikeModelSchematic.png}
\caption{Left: Schematic of bicycle model. Right: Diagram showing tracking error $e$, heading error $\Delta\Psi$, lookahead distance $x_\mathrm{LA}$, and lookahead error $e_\mathrm{LA}$.}
\label{fig:bikemodel}
\end{figure}

\begin{subequations}
\label{eq:bm}
\begin{align}
	\dot{\beta} &= \frac{F_\mathrm{yf}+F_\mathrm{yr}}{mU_x} - r \qquad \dot{r} = \frac{aF_\mathrm{yf} - bF_\mathrm{yr}}{I_z} \label{bm1} \\
	\dot{e} &= U_x (\beta + \Delta\Psi) \qquad \Delta\dot{\Psi} = r - U_x\kappa \label{eq:bm2} 
\end{align}
\end{subequations}

Where $U_x$ is the vehicle forward velocity and $F_{yf}$ and $F_{yr}$ the front and rear lateral tire forces. The vehicle mass and yaw inertia are denoted by $m$ and $I_z$, while the geometric
parameters $a$ and $b$ are shown in Fig.~\ref{fig:bikemodel}.

As automotive racing frequently occurs near the limits of tire force saturation, 
lateral tire force is modeled using the nonlinear Fiala brush tire model, assuming a single coefficient of friction and a parabolic force distribution \cite{Pacejka2012}.
 The lateral tire forces are functions of the front and rear tire slip angles $\alpha_\mathrm{F}$ and $\alpha_\mathrm{R}$.

\begin{eqnarray}
\label{eq:fiala}
\small
	F_\mathrm{y}&=&\begin{cases} -C_{\alpha}\tan\alpha + \frac{C_{\alpha}^2}{3\mu F_\mathrm{z}} |\tan\alpha| \tan\alpha \vspace{2mm} \\ \hspace{4mm}- \frac{C_{\alpha}^3}{27\mu^2F_\mathrm{z}^2}\tan^3\alpha,
\hspace{4mm}  |\alpha| < \arctan{\left(\frac{3\mu F_\mathrm{z}}{C_\alpha}\right)} \\ -\mu F_\mathrm{z}\text{sgn} \ \alpha, \hspace{14mm} \mathrm{otherwise} \end{cases} \nonumber\\
\end{eqnarray}

where $\mu$ is the surface coefficient of friction, $F_\mathrm{z}$ is the normal load, and $C_\alpha$ is the tire cornering stiffness. The linearized tire slip angles are given by

\begin{subequations}
\begin{align}
	\alpha_\mathrm{F} &= \beta + \frac{ar}{U_\mathrm{x}} - \delta\\
	\alpha_\mathrm{R} &= \beta - \frac{br}{U_\mathrm{x}}
\end{align}
\end{subequations}

\subsection{Linear Time Varying Model in the Lifted Domain}

While the nonlinear tire model presented in (\ref{eq:fiala}) captures the effect of tire saturation at the limits of handling, established methods for design and analysis
of iterative learning controllers requires a linear system description. With the assumption of low lateral acceleration, a simple linear tire model is given by,

\begin{align}
	F_y = -C_\alpha\alpha
\end{align}

The lateral error dynamics of the vehicle in response to a \textit{feedforward} steer angle input (recall that there is also a feedback steer angle $\delta_\mathrm{FB}$ in the loop)
 can then be represented by the continuous, linear time varying (LTV) state space model,

\begin{subequations}
\label{eq:cts}
\begin{align}
	\dot{x}(t) &= A_c(t) x + B_c(t)\delta_L + d_c(t)\\
	 e &= C_cx(t)\\
	 x &= [e \hspace{2mm} \Delta\Psi \hspace{2mm} r \hspace{2mm} \beta]^T
\end{align}
\end{subequations}
Where $\delta_L$ is the learned steering input. The LTV framework is chosen to account for the variation in vehicle 
velocity $U_x$ along the track (Fig.~\ref{problemInfo}). While the longitudinal dynamics are not explicitly accounted for in (\ref{bm1}), 
allowing for the state matrices to vary with measured values of $U_x$ enables more accurate controllers and analysis. The time-varying matrices $A_c$, $B_c$, and $C_c$ are given by

\begin{multline}
\label{eqn:Amatrix}
A_c(t)  =  \\
\left[\begin{smallmatrix}
  0 & U_\mathrm{x}(t) & 0 & U_\mathrm{x}(t) \\ 
  0 & 0 & 1 & 0 \\ 
  \frac{-ak_\mathrm{P} C_\mathrm{F}}{I_\mathrm{z}}  & \frac{-ak_\mathrm{P}x_{LA}C_\mathrm{F}}{I_\mathrm{z}}  & \frac{-(a^2C_\mathrm{F}+b^2C_\mathrm{R})}{U_\mathrm{x}(t)I_\mathrm{z}} & \frac{bC_\mathrm{R} - aC_\mathrm{F}}{I_\mathrm{z}}  \\
  \frac{-k_\mathrm{P}C_\mathrm{F}}{mU_\mathrm{x}(t)}  & \frac{-k_\mathrm{P}x_{LA}C_\mathrm{F}}{mU_\mathrm{x}(t)}  & \frac{bC_\mathrm{r}-aC_\mathrm{f}}{mU_\mathrm{x}(t)^2}-1 & \frac{-(C_\mathrm{F} + C_\mathrm{R})}{mU_\mathrm{x}(t)}
 \end{smallmatrix}\right]
 \end{multline}

\begin{align}
B_c(t) =[0 \hspace{2 mm} 0 \hspace{3 mm} \frac{a C_\mathrm{F}}{I_\mathrm{z}} \hspace{3 mm}  \frac{C_\mathrm{F}}{mU_\mathrm{x}(t)}]^T
\end {align}
\begin{align}
C_c =[1 \hspace{2 mm} 0 \hspace{2 mm} 0 \hspace{2 mm} 0]
\end{align}

The disturbance $d_c$, is assumed constant from lap to lap and is given by

\begin{align}
d_c(t) = [0 \hspace{2 mm} -\kappa U_x(t) \hspace{3 mm} 0 \hspace{3 mm}  0]^T
\end{align}

The next step is to discretize (\ref{eq:cts}) by the controller sample
time $T_s$, resulting in the discrete time system,

\begin{align}
x_j(k\!\!+\!\!1) &= A(k)x_j(k) + B(k)\delta^L_j(k)\\
e_j(k) &= Cx_j(k)
\label{eq:discreteEOM}
\end{align}

Where $k = 1\ldots N$ is the time sample index, and $j = 1\ldots M$ is the number of iterations (i.e. the number of laps around the track). 
Development and analysis of the iterative learning controllers in the next section will be made easier by representing the system dynamics in the ``lifted-domain",
where the inputs  and outputs  are stacked into arrays and related by matrix multiplication, as follows:

\begin{subequations}
\label{eq:liftedDomain}
\begin{align}
\mathbf{e}_j        &= \mathbf{P}\mathbf{\delta^L_j} + \mathbf{d}_j\\
\bf{\delta}_j^L  &=[\delta^L_j(0) \cdots \delta^L_j(N\!\!-\!\!1)]^T\\
\bf{e}_j        &= [e_j(1) \cdots e_j(N)]^T
\end{align}
\end{subequations}

Where the elements of the $N \times  N$ matrix $\bf{P}$ are given by,

\begin{equation}
p_{lk} = \begin{cases} 0 &\mbox{if } l < k \\ 
CB(k) & \mbox{if } l = k \\
CA(l)A(l-1)\cdots A(k)B(k) &\mbox{if } l > k \end{cases}
\end{equation}

Note that for the case where $U_x$ is constant, we have a linear time invariant (LTI) system given by only
$N$ elements, with $p(k) = CA^{k-1}B$, and the resulting lifted-domain matrix $\bf{P}$ is Toeplitz.

\section{Controller Design}

With the representation of the steering dynamics for a given lap represented by (\ref{eq:liftedDomain}), the next step is to
design algorithms that determine the learned steering input $\bf{\delta}^L_{j\!+\!1}$ for the next lap, given the error response
from the completed lap $\bf{e_{j}}$. A common framework for iterative learning algorithms is to choose $\bf{\delta}^L_{j\!+\!1}$ as follows \cite{bristow},

\begin{align}
\bf{\delta}^L_{j\!+\!1} = Q(\bf{\delta_{L,j}} - L\bf{e_{j}})
\label{eq:ctrlLaw}
\end{align}

Where $\bf{Q}$ is the $N \times N$ filter matrix, and $\bf{L}$ is the $N \times N$ learning matrix. In following sections, the
matrices $\bf{Q}$ and $\bf{L}$ will be obtained by designing a PD type iterative learning controller as well as a quadratically
optimal (Q-ILC) learning controller.

\subsection{Proportional-Derivative Controller}

The proportional-derivative iterative learning controller computes the steering addition $\delta^L_j(k)$ at a given time index $k$ based on the error $e_{j\!-\!1}(k)$ and the change in error
at the same time index from the previous lap,

\begin{equation}
	\delta^L_{{j\!+\!1}}(k) = \delta^L_j(k) - k_pe_{j\!-\!1}(k) - k_d(e_{j\!-\!1}(k) - e_{j\!-\!1}(k-1))
\end{equation}

Where $k_p$ and $k_d$ are the proportional and derivative gains. In the lifted domain representation from (\ref{eq:ctrlLaw}), the resulting learning matrix $\bf{L}$ is given by
\begin{equation}
	\bf{L} = \begin{bmatrix} -(k_p\!+\!k_d) &               &    0 \\ 
						 k_d           &  \ddots  &   \\ 
						 0             &   k_d            &    -(k_p\!+\!k_d)  \\ \end{bmatrix}
\end{equation}

 An important consideration in choosing the gains $k_p$ and $k_d$ is achieving a monotonic decrease in the path tracking error on every lap. 
 This property, known as monotonic convergence, occurs if the following condition is met \cite{bristow}:

\begin{equation}
	\gamma \triangleq \bar{\sigma}(\mathbf{P}\mathbf{Q}(I-\mathbf{L}\mathbf{P})\mathbf{P}^{-1}) < 1
	\label{eq:MS}
\end{equation}

Where $\bar{\sigma}$ is the maximum singular value. In this case, the value of $\gamma$ provides an upper bound on the decrease in the tracking error norm from lap to lap, i.e. 
\begin{equation}
	||e_\infty - e_{j+1}||_2 \leq \gamma ||e_\infty-e_j||_2
\end{equation}
where $e_\infty$ is the converged path tracking error. The monotonic stability condition differs from its weaker counterpart of asymptotic stability in that we are guaranteed that 
 the path tracking error on the first lap (i.e. with no added learning input) is the worst case performance.

Fig. \ref{fig:stabPlot} shows values of $\gamma$ for both an unfiltered PD controller ($\mathbf{Q} = I$), and for a PD controller with a 2 Hz low pass filter. The $\gamma$ values are plotted as a
contour map against the controller gains $k_p$ and $k_d$. Addition of the low-pass filter
assists in achieving controller monotonic stability by removing oscillations that are frequently generated by iterative learning controllers when trying to remove small reference tracking
errors after several iterations. Since the filtering occurs when generating a control signal for the next lap, the filter $\mathbf{Q}$ can be zero-phase.

\begin{figure}
\centering
\includegraphics[width=3.5 in]{figures/monotonicStability.png}
\caption{Values of convergence bound $\gamma$ vs. $k_p$ and $k_d$ for PD iterative learning controller with (left) no filtering and (right) with a 2 Hz low-pass filter. Lower values of $\gamma$ correspond to faster convergence.
Shaded regions correspond to systems with monotonic stability. }
\label{fig:stabPlot}
\end{figure}

However, testing for linear stability is insufficient controller design given that racing frequently occurs near the limits of vehicle handling, when the vehicle dynamics are described 
by the nonlinear equations of motion presented in Section \ref{sec:problemDescription}. To test the PD controller feasibility, the vehicle tracking performance over multiple laps is
simulated using the path curvature and speed profile shown in Fig.~\ref{problemInfo}.
Simulated results for the root-mean-square (RMS) tracking error are shown in Fig.~\ref{fig:PDsimResults} for both the linear state dynamics prescribed by (\ref{eq:discreteEOM}) and 
the nonlinear dynamics model given by (\ref{eq:bm})
and (\ref{eq:fiala}). The results indicate that as the vehicle corners closer to the limits of handling, the tracking performance of the ILC degrades relative to the expected performance
given by the linear model, but can still be expected to converge over relatively few iterations.

\begin{figure}
\centering
\includegraphics[width=3 in]{figures/PDsimResults.png}
\caption{Simulated results for root-mean-square path tracking error for PD iterative learning control at several values of vehicle lateral acceleration. Controller gains are $k_p = .05$, $k_d = .05$. Results from a nonlinear vehicle dynamics simulation are compared to results
from the linear vehicle model. }
\label{fig:PDsimResults}
\end{figure}

\subsection{Quadratically Optimal Controller}\label{sec:controller}

An alternate approach to determining the learned steering input $\bf{\delta}^L_j$ is to minimize a quadratic cost function for the next lap:

\begin{align}
J_{j\!+\!1} = e_{j\!+\!1}^TTe_{j\!+\!1} + \delta^{L\,T}_{j\!+\!1} R \; \delta^L_{j\!+\!1}+\Delta_{j\!+\!1}^TS\Delta_{j\!+\!1}
\label{eq:QILC}
\end{align}

Where $\Delta_{j\!+\!1} = \mathbf{\delta^L_{j\!+\!1}} - \mathbf{\delta^L_{j}}$ and the $N \times N$ matrices $T$, $R$, and $S$ are weighting matrices.
This formulation allows the control designer to weight the competing objectives of minimizing tracking error, control effort, and change in the control signal from lap to lap.
While constraints can be added to the optimization problem, the unconstrained problem in (\ref{eq:QILC}) can be solved analytically \cite{bristow} to obtain desired controller and filter matrices:

\begin{subequations}
\label{eq:analSol}
\begin{align}
	\mathbf{Q} &= (P^TTP + R + S)^{-1}(P^TTP+S)\\
	\mathbf{L} &= (P^TTP + S)^{-1}P^TT
\end{align}
\end{subequations}

An advantage to the quadratically optimal control design over the simple PD controller is that the controller matrices $\mathbf{Q}$ and $\mathbf{L}$ take the linear time-varying dynamics $\mathbf{P}$ into account. This allows the iterative learning algorithm
to take into account changes in the steering dynamics due to changes in vehicle velocity. However, a disadvantage is that determining new $\mathbf{Q}$ and $\mathbf{L}$ matrices every lap
requires resolving (\ref{eq:analSol}), which can be computationally expensive for fast sampling rates.

\begin{figure}
\centering
\includegraphics[width=3 in]{figures/QILCsimResults.png}
\caption{Simulated results for root-mean-square path tracking error for Q-ILC at several values of vehicle lateral acceleration, with $T = R = I$ and $S = \mathrm{100} I$. Results from a nonlinear vehicle dynamics simulation are compared to results
from the linear vehicle model.}
\label{fig:QILCsimResults}
\end{figure}

Fig.~\ref{fig:QILCsimResults} shows simulated results for the quadratically optimal ILC controller. The results are similar to those in Fig.~\ref{fig:PDsimResults} in that the predicted controller performance
is worse when the simulation accounts for nonlinear tire dynamics. However, the simulation still shows a rapid decrease in path tracking error over the first ten laps, with RMS errors on the order
of 8-9 cm at lateral accelerations of .8 $g$.

\section{Experimental Results}

Experimental data for both controllers was collected over multiple laps at Thunderhill Raceway, a 3 mile paved racetrack in Willows, CA, with track boundaries shown in Fig.~\ref{problemInfo}a. The experimental testbed is an autonomous Audi TTS
equipped with an electronic power steering motor, active brake booster, and throttle by wire Fig.~\ref{fig:shelleyPic}. Vehicle
and controller parameters are shown in Table 1.

\begin{figure}
\centering
\includegraphics[width=3.5 in]{figures/shelleyPic.png}
\caption{Autonomous Audi TTS with electronic power steering, brake booster, and throttle by wire.}
\label{fig:shelleyPic}
\end{figure}

An integrated Differential Global Positioning System (DGPS) and Inertial Measurement Unit (IMU) is used to obtain vehicle state information, and a localization algorithm 
determines the lateral path tracking error $e$, heading error $\Delta\Psi$, and distance along the desired racing line. The steering controller updates at 200 Hz, 
and the iterative steering corrections are calculated after each lap from data downsampled to 10 Hz. The corrections are then applied as a function of distance along the track using an
interpolated lookup table. For safety reasons, a steady-state feedforward steering algorithm \cite{kapania} is also applied to keep the tracking error on the first lap below 1 m.

\begin{table}[h]
\small
\begin{center}
\caption{Vehicle Parameters}\label{tb:params}
\begin{tabular}{lccc}
Parameter & Symbol & Value & Units \\\hline
Vehicle mass & $m$ & 1500 & kg \\
Yaw moment of inertia & $I_z$ & 2250 & $\mathrm{kg \cdot m}^2$\\
Front axle to CG & $a$ & 1.04 & m\\
Rear axle to CG & $b$ & 1.42 & m\\
Front cornering stiffness & $\mathrm{C}_\mathrm{F}$ & 160 & $\mathrm{kN \cdot rad}^{-1}$ \\
Rear cornering stiffness & $\mathrm{C}_\mathrm{R}$ & 180 & $\mathrm{kN \cdot rad}^{-1}$ \\
Lookahead Distance       & $x_\mathrm{LA}$          &  15.2 & $\mathrm{m} $ \\
Lanekeeping Gain         & $k_{\mathrm{LK}}$         & .053 & $\mathrm{rad\,m^{-1}}$\\
Lanekeeping Sample Time  & $t_s$                        & .005 & s\\
ILC Sample Time          & $T_s$                        & .1   & s\\
PD Gains                 & $k_p$ and $k_d$           & .02 \& .4 & $\mathrm{rad\,m^{-1}}$\\
Q-ILC Matrix             & $T$ and $R$              &  $I$      & - \\
Q-ILC Matrix             & $S$                       & 100\,$I$  & - \\\hline
\end{tabular}
\end{center}
\end{table}

Fig.~\ref{fig:PDexp} shows the applied iterative learning signals and resulting path tracking error over four laps using the PD learning algorithm. The car is driven
aggressively at peak lateral/longitudinal accelerations of 8 $\mathrm{m/s^2}$. On the first lap, despite the 
incorporation of a feedforward-feedback controller operating at a high sampling rate, several transient spikes in tracking error are visible due to the underdamped tire dynamics near the limits of handling.
However, the iterative learning algorithm is able to significantly attenuate these transient spikes over just two or three laps. Similar qualitative
results occur for the quadratically optimal ILC.

\begin{figure}[h]
\centering
\includegraphics[width=3.5 in]{figures/PDexp.png}
\caption{Experimental results for path tracking error with PD learning controller, at peak lateral accelerations of 8 $m/s^2$.}
\label{fig:PDexp}
\end{figure}

In Fig.~\ref{fig:allExp}, results are shown for versions of the velocity profile in Fig.~\ref{problemInfo} that are scaled to achieve different vehicle accelerations. The results
show that at low vehicle accelerations, when vehicle dynamics are accurately prescribed by a linear tire model, the feedback-feedforward controller is able to keep the vehicle close to the
desired path, leaving little room for improvement through iterative learning control. However, as the speed profile becomes more aggressive, the path tracking degrades in the presence
of highly transient tire dynamics, and iterative learning control can be effectively used to obtain tight path tracking path over two or three laps of racing. 
In practice, the performance of both the PD algorithm and quadratically optimal algorithms are similar, and an important observation is that the RMS tracking error increases slightly from
lap-to-lap at the end of some tests. While not predicted in simulation, this behavior is not unreasonable given unmodelled sensor noise and disturbances that vary from lap to lap. 
More refined tuning of the gain matrices may be able to prevent this RMS error increase, or the ILC algorithm can be stopped after several iterations once the tracking performance is acceptable.

\begin{figure}
\centering
\includegraphics[width=3 in]{figures/allExp.png}
\caption{Experimental results for both controller types over a variety of lateral accelerations.}
\label{fig:allExp}
\end{figure}

%\addtolength{\textheight}{-2cm}   % This command serves to balance the column lengths
                                  % on the last page of the document manually. It shortens
                                  % the textheight of the last page by a suitable amount.
                                  % This command does not take effect until the next page
                                  % so it should come on the page before the last. Make
                                  % sure that you do not shorten the textheight too much.

\section{Conclusion}
This paper demonstrates the application of iterative learning control (ILC) methods to achieve accurate steering control
of an autonomous race car over multiple laps. Two different algorithms, proportional-derivative (PD) and 
quadratically optimal (Q-ILC) learning control are tested in simulation and then used to experimentally 
eliminate path tracking errors caused by the highly transient nature of the vehicle lateral dynamics near the limits of tire-road friction. 
Both learning algorithms provide comparable lap-to-lap tracking performance, although the PD method is computationally fast enough to run in real time. Two clear limitations of the presented iterative learning controllers present avenues for future work. While representing the vehicle dynamics
with a linear, time-varying model allows for the quadratically optimal ILC algorithm to account for varying longitudinal speed, a better approach is to linearize the vehicle
dynamics at each point on the track and create an affine time-varying model whose path tracking error can be minimized. Furthermore, applying a steering wheel input to eliminate
lateral errors will work only if the vehicle is near the limits of handling, but has not fully saturated the available tire force and entered a limit understeer condition.
In this case, a separate controller must be developed to modify the racing line and velocity profile for future laps in order to reduce the vehicle cornering forces.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{ACKNOWLEDGMENT}
This research is supported by the National Science Foundation Graduate Research Fellowship Program (GRFP). 
The authors would like to thank the members of the Dynamic Design Lab at Stanford University and the Audi Electronics Research Lab.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliography{ACC2015}

\end{document}

### 2019-02-13 conference/iv2019/manuscript.tex
converting last two chapters of thesis into journal paper

@@
 
-\author{Nitin R. Kapania$^{1}$, Vijay Govindarajan$^{2}$, Francesco Borrelli$^{2}$% <-this % stops a space
+\author{Nitin R. Kapania$^{1}$, Vijay Govindarajan$^{2}$, Francesco Borrelli$^{2}$, J. Christian Gerdes$^{1}$% <-this % stops a space
 \thanks{*This work was supported by the Hyundai Center of Excellence}% <-this % stops a space
@@
         & b. Wait for pedestrian crossing \\
-        & event to materialize before yielding. \\
+        & event to materialize before \\
+        & yielding. \\
 
         & \\
- 3. Smoothness & a. Low acceleration: Nominal brake \\ 
- 			  & and acceleration limits of 2 $\mathrm{m/s^2}$ \\
+ 3. Smoothness & a. Low acceleration: Nominal \\ 
+ 			  & brake and acceleration limits \\
+ 			  & of 2 $\mathrm{m/s^2}$ \\
 
@@
 
-\If {$state = HARD\hspace{1 mm}BRAKING$}
+\If {$state = HARD\hspace{1mm}BRAKING$}
 	\State $v_{des} = \mathrm{getDesiredSpeed}(d, v)$
 	\State $a = -\frac{v^2}{2d} + k_s(v - v_{des})$ 	
-	\If {$\textbf{not} \hspace{1 mm} \mathrm{inCrosswalk}(x_p)$} \State $state \leftarrow DRIVING$
+	\If {$\textbf{not} \hspace{1mm} \mathrm{inCrosswalk}(x_p)$} \State $state \leftarrow DRIVING$
 
@@
 
-\EndIf
+% \EndIf
 \If {$state = SPEED\hspace{1mm}UP$}
@@
 
+\begin{table}[h]
+\begin{center}
+\caption{Experimental Parameters}\label{tb:expparams2}
+\begin{tabular}{llll}
+Trial Number & Gap Accepted (sec) & Pedestrian Entry & Mode\\\hline\hline
+1 & 4.0 & right & Yield\\ 
+2 & 1.0 & right & Speed Up\\ 
+3 & 7.0 & right & Yield\\ 
+4 & 2.5 & right & Hard Brake\\ 
+5 & 3.0 & left & Yield \\ 
+6 & 1.0 & left & Speed Up\\\hline
+\end{tabular}
+\end{center}
+\end{table}
+
 \section{Experimental Results}
@@
 
-\begin{table}[h]
-\begin{center}
-\caption{Experimental Parameters}\label{tb:expparams2}
-\begin{tabular}{llll}
-Trial Number & Gap Accepted (sec) & Pedestrian Entry & Mode\\\hline\hline
-1 & 4.0 & right & Yield\\ 
-2 & 1.0 & right & Speed Up\\ 
-3 & 7.0 & right & Yield\\ 
-4 & 2.5 & right & Hard Brake\\ 
-5 & 3.0 & left & Yield \\ 
-6 & 1.0 & left & Speed Up\\\hline
-\end{tabular}
-\end{center}
-\end{table}
-
-The experimental velocity and acceleration is plotted vs $d$ in 
-Figure \ref{fig:expPlot}.
+The experimental velocity and acceleration is plotted vs $d$ in Figure \ref{fig:expPlot}.
 In Trial 1, the pedestrian accepts a gap of four seconds, giving the vehicle enough time to yield at a deceleration of 2 $\mathrm{m/s^2}$.

### 2019-02-13 journal/TCST2019/manuscript.tex
converting last two chapters of thesis into journal paper

\documentclass[9pt,shortpaper,twoside,web]{ieeecolor}
\usepackage{generic}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{dsfont}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{epsfig}
\usepackage{tikz}
\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
            \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\markboth{\journalname, VOL. XX, NO. XX, XXXX 2017}
{Author \MakeLowercase{\textit{et al.}}: Preparation of Brief Papers for IEEE TRANSACTIONS and JOURNALS (February 2017)}
\begin{document}
\title{Learning at the Racetrack: Data-driven Methods to Improve Racing Performance over Multiple Laps}
\author{Nitin R. Kapania and J Christian Gerdes
\thanks{This work was supported in part by the National Science Foundation Graduate Research Fellowship Program and the Stanford Graduate Fellowship. }
\thanks{Nitin Kapania is a postdoctoral fellow at Stanford University and a visiting scholar at UC Berkeley (e-mail: nkapania@stanford.edu).}
% \thanks{S. B. Author, Jr., was with Rice University, Houston, TX 77005 USA. He is 
% now with the Department of Physics, Colorado State University, Fort Collins, 
% CO 80523 USA (e-mail: author@lamar.colostate.edu).}
\thanks{J. C. Gerdes is a Professor 
at Stanford University (e-mail: gerdes@stanford.edu).}}

\maketitle

\begin{abstract}
Autonomous vehicles will generate tremendous data from the variety of sensors they employ to track the surrounding environment. This data is inherently valuable, as it gives algorithm designers the potential to leverage prior experience in order to improve driving performance over time. This paper uses the lens of autonomous racing to provide an example of how data from previous iterations of driving can be used to improve quantitative metrics of performance. Two complementary algorithms are demonstrated in this paper. The first type uses iterative learning control (ILC) to simultaneously improve lateral and longitudinal tracking of the desired racing trajectory from lap-to-lap, while the second type of algorithm is focused on altering the trajectory itself using a search method. When driven experimentally at the limits of handling, the result is a reduction in lap time of nearly 1.4 seconds, a major improvement.

\end{abstract}

\begin{IEEEkeywords}
 Autonomous driving, iterative learning control, A* search
\end{IEEEkeywords}

\section{Introduction}
\label{sec:introduction}

\subsection{Problem Description and Proposed Solution}
Several methods exist in the literature to generate autonomous racing trajectories (~\cite{casanova},\cite{kelly},\cite{theodosis},\cite{kapaniadscc}), which typically comprise of a longitudinal speed profile $U_x(s)$ and a desired curvature profile $\kappa(s)$ as a function of distance along the track. These trajectories can then be tracked using one of several path following controllers, such as (\cite{mickcop},\cite{fricStudy},\cite{mickgeneral},\cite{shladover},\cite{nagai}).

Using the trajectory generated by \cite{kapaniadscc} and the trajectory following controller presented by \cite{kapania}, the
authors generated a test run of an autonomous race vehicle (see Fig.\ref{fig:carpic}(a)) at the limits of handling on the Thunderhill Raceway course in Willows, CA (Fig.~\ref{fig:carpic}(b)). The results are shown in Fig. \ref{fig:expErrors}.

\begin{figure}[h]
\centering
\includegraphics[width=3.5in]{figures/pic1.png}
\caption{(a) ``Shelley" - Stanford's autonomous self-driving race vehicle. (b) Top-down view of Thunderhill Raceway course. }
\label{fig:carpic}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=3.3in]{figures/expErrors.eps}
\caption{Controller tracking performance and tire slip norm on a test run at the limits of handling ($\mu =$ 0.94).(a) Desired
vs actual speed of the vehicle. (b) Lateral tracking error and (c) tire slip norm as a function of distance along the path.}
\label{fig:expErrors}
\end{figure}

While the vehicle is able to generate lap times that are comparable to an amateur expert, the resulting lap times are 2-3 seconds below that of a professional human race car driver. There are several ways to analyze why this difference arises, but a simple insight that makes the case for lap-to-lap learning comes from viewing the trajectory tracking performance and friction utilization of the controller, shown in Fig.~\ref{fig:expErrors}.

One of the issues shown in Fig.~\ref{fig:expErrors} is the relatively poor controller tracking on a few labeled sections of the race track. This includes errors in the lateral path tracking 
(sections \circled{2}, \circled{4} - \circled{7}) and failing to drive at the desired speed (sections \circled{1}, \circled{2}, \circled{7}), resulting directly in a loss 
of lap time. The speed and lateral path tracking will be improved in this paper with the addition of iterative learning control
(ILC) algorithms.

The second issue shown in Fig.~\ref{fig:expErrors} is the inconsistent usage of the tire friction capacity, as judged by the tire slip norm metric.
The tire slip norm, formalized in \cite{mickthesis}, is given by $\zeta$:

\begin{equation}
\label{eq:zeta}
\zeta = \sqrt{\left(\dfrac{\alpha}{\alpha_{p}}\right)^2 + \left(\dfrac{\sigma}{\sigma_p}\right)^2}
\end{equation}

Where $\sigma$ and $\alpha$ are the longitudinal and lateral tire slip for a given tire, and $\sigma_p$ and $\alpha_p$ are empirically
determined \textit{peak slip} values resulting in maximum longitudinal and lateral tire force generation. As a result, $\zeta < 1 $ corresponds to the tires having excess force generation capacity, while $\zeta > 1$ corresponds to tire saturation.

Fig.~\ref{fig:expErrors} shows inconsistent usage of tire friction across many turns. On some turns (sections \circled{2} and \circled{8}), the vehicle significantly exceeds the limits of handling, and the car's stability control systems kick in to regain control, slowing the car down in the process. On other turns (sections \circled{3}, \circled{5}, and \circled{7}),
the vehicle uses only a portion of the available tire force, indicating the vehicle can actually drive with higher acceleration on the next lap. Learning from prior runs to find the optimal acceleration (or $\mu$ parameter) for each part of the track will also be accomplished in this paper via an A* search algorithm.

\subsection{Relation to Prior Work}
\subsubsection{Iterative Learning Control}

Iterative learning control (ILC) is based on the notion that the performance of a system that executes the \textit{same task} multiple times can be improved by learning from previous executions (trials, iterations, or in our case, laps of racing) \cite{bristow}.
On every iteration, a control signal is applied to a system in order to follow an ideal, unchanging ``reference trajectory". 
The tracking error for that iteration is recorded, and a learning algorithm is applied to improve the control signal and achieve
more accurate system performance on the next iteration. There are a variety of learning algorithms used, but most attempt to correct the tracking error by using a model of the system to determine the augmentation to apply to the prior control signal. This process is repeated until the reference tracking performance becomes satisfactory.

Because iterative learning control works best when learning to follow the same reference trajectory under nearly constant ambient conditions, the most common applications of ILC are in the field of automated manufacturing. Notable examples include CNC machining \cite{kimdi}, industrial robotics \cite{freeman}\cite{hladowski}, piezolectric stage positioning \cite{huang}, motor control \cite{mohammad}, and microdeposition \cite{hoelzle}. However, the rise of automated systems outside factory environments has led to important applications of ILC for ground and air robotics. Chen and Moore \cite{chen} proposed a simple iterative learning scheme to improve path-following of a ground vehicle with omni-directional wheels. Purwin and Andrea synthesized an iterative controller using least-squares methods to aggressively maneuver a quadrotor unmanned aerial vehicle (UAV) from one state to another \cite{purwin}. Sun et al. \cite{sun} proposed an iterative learning controller for speed regulation of high-speed trains. Other novel applications where ILC has been applied include Homogeneous Charge Compression Ignition (HCCI) \cite{slepicka2016iterative} and wing vibration suppression \cite{xu2015iterative}.

\subsubsection{Trajectory Modification Algorithms}

Iterative learning control (ILC) algorithms can help an autonomous vehicle follow a desired trajectory more precisely over 
several laps of driving. However, ILC algorithms do not alter the desired trajectory itself, only the input signals that attempt
to track the trajectory. This will not be sufficient at the limits of handling. Consider Fig.~{\ref{fig:expErrors}, where the desired speed profile was generated assuming a tire-road friction value of $\mu = 0.94$. While this is a reasonable assumption overall, there are several parts on the track where the vehicle exceeds the available friction and begins to understeer. Region \circled{2} is one example. The tire slip norm $\zeta$ climbs above one, and as a result, the vehicle begins
to slide off the track, resulting in the large negative tracking error spike in Fig.~{\ref{fig:expErrors}(b). The vehicle's stability algorithms must kick in and slow the vehicle down, which results in the speed tracking error seen in Fig.~{\ref{fig:expErrors}(a). In this situation, ILC will be unable to achieve better speed tracking and path tracking performance. The tires are saturated and the car is slowly beginning to careen off the track. Simply steering more
on the next lap will not achieve better tracking performance due to the saturation of the steering actuator. In order to recover, the vehicle must deviate from the planned trajectory, either by slowing down or by taking a wider radius turn.

Some trajectory modification algorithms previously investigated in
the literature have focused on modifying the \textit{lateral} trajectory by altering the curvature profile as the vehicle understeers or oversteers. For example, Theodosis and Gerdes presented an algorithm to gradually widen the radius of a turn in response to a detected understeer \cite{paulthesis}. The algorithm was validated
experimentally, but assumed sufficient availability of road width. Klomp and Gordon also developed a strategy for recovering from
vehicle understeer by solving for an optimal emergency braking profile to minimize deviation from the desired path \cite{klomp}. On the other hand, Brown et al.~\cite{brown2017safe} presented a model predictive control (MPC) approach that generally aimed to follow a planned vehicle trajectory at the limits. However, if the vehicle was at risk of understeering or oversteering, the MPC algorithm would deviate laterally from the planned trajectory in order to maintain stability of the vehicle without driving off the road.

Another approach to trajectory modification is Learning Model Predictive Control (LMPC), developed by Rosolia et al. \cite{rosolia2017autonomous} for the purpose of autonomous racing. In LMPC, data from every lap of racing is used to build invariant sets and approximate the value function for the generally non-convex lap time minimization problem. In practice, this enables the controller to explore different areas of the state space on each successive lap, modifying the desired trajectory on each lap in order to continuously improve the lap time.

\subsection{Order of Contributions}

Due to the repetitive nature of automotive racing, iterative learning control techniques are a promising method to gradually
eliminate \textit{trajectory tracking} errors . For this application of ILC, the repetitive trials are laps of racing, and the reference trajectory is the optimal speed profile and curvature developed in \cite{kapaniadscc}. This paper will present adaptations of two established ILC methods 
(proportional-derivative and quadratically optimal) for use in the Audi TTS racing system. Section \ref{sec:ch4dsm} presents
both coupled and decoupled models of the vehicle lateral and longitudinal dynamics. These models are converted from state-space representations to \textit{lifted domain}
representations required for iterative learning control, and the proportional-derivative and quadratically optimal ILC algorithms are presented in \S \ref{sec:pdcontroller} and \S \ref{sec:qopt}. Section \ref{sec:simres} presents simulated results of the ILC algorithms for a sample vehicle trajectory, and finally,
experimental results showing a gradual reduction of trajectory-following errors is presented in \S \ref{sec:ch4ExpRes}.

For situations where the vehicle underperforms due to excessive oversteer or understeer, an A* search algorithm is presented
that iteratively modifies portions of the \textit{velocity} profile to be more conservative if a stability violation is encountered on a prior lap. This is accomplished by generalizing the speed profile such that each part of the track can be driven with a learned  value of friction $\mu$ and therefore a different maximum
acceleration. The algorithm also modifies the velocity profile to be more aggressive if the tires are not being actuated at the limits. In this way, the algorithm is focused on learning the time-optimal friction profile $\mu^\star(s)$ by searching through datasets obtained over multiple laps.

\section{Iterative Learning Control}

\subsection{Dynamic System Model}
\label{sec:ch4dsm}
The ILC algorithms we consider require a set of closed-loop system dynamics that are (a) stable to any disturbance input, and (b) expressible as an \textit{affine} discrete dynamical system.
In our case, we have two subsystems: the steering controller and the longitudinal speed control. Stability of the steering controller
under lanekeeping feedback was shown in the linear case by \cite{rossetter2002} and in the saturated case by \cite{talvala}. Similar analyses can be considered to show the stability of a simple proportional speed-following controller.

The more difficult task is expressing the dynamics of the two subsystems using an affine model, given the tendency for the vehicle tires to saturate at the limits. This paper will model the dynamics with a bicycle model with single longitudinal input $F_x$ and front steer input $\delta$, shown in Fig.\ref{fig:bikemodel}.

\begin{figure}[h]
\centering
\includegraphics[width=3.5in]{figures/BikeModelSchematic.eps}
\caption{Schematic of bicycle model with steer input $\delta$ and longitudinal input $F_x$.}
\label{fig:bikemodel}
\end{figure}

Nonlinear, coupled equations of motion are provided in (\ref{eq:fullNL})-(\ref{eq:fullNL2}):

\begin{align}
\label{eq:fullNL}
	\dfrac{de}{dt} &= \left(v+U^\mathrm{des}_x(s)\right)(\beta + \Delta\Psi) \\
	\dfrac{dv}{dt} &= \left(v+U^\mathrm{des}_x(s)\right)\beta r + \frac{F_x}{m} - \frac{F_\mathrm{yf}(\alpha_\mathrm{f},F_x)\delta}{m}\\
	\dfrac{d\beta}{dt} &= \frac{F_\mathrm{yf}(\alpha_\mathrm{f}, F_x) + F_\mathrm{yr}(\alpha_\mathrm{r},F_x)}{m\left(v + U^\mathrm{des}_x(s)\right)}-r \\
	\dfrac{dr}{dt}     &= \frac{aF_\mathrm{yf}(\alpha_\mathrm{f}, F_x) - bF_\mathrm{yr}(\alpha_\mathrm{r}, F_x)}{I_z} \\
	\dfrac{d\Delta\Psi}{dt} &= r \label{eq:fullNL2}
\end{align}
The system dynamics presented in (\ref{eq:fullNL}) have five states, yaw rate $r$, vehicle sideslip $\beta$, lateral path tracking error $e$, heading error $\Delta \Psi$, and speed tracking error $v$, defined by:

\begin{equation}
v = U_x - U^\mathrm{des}_x
\end{equation}

The potential for coupling between the subsystems is apparent from  (\ref{eq:fullNL})-(\ref{eq:fullNL2}), not only directly from the state equations but also due to the complex nature of tire force generation at the handling limits. As shown in Fig.~\ref{fig:coupledTires}, as longitudinal
force $F_x$ is distributed across the tires, the available lateral force decreases. At the limits of handling, this
\textit{derating} of the lateral force may become significant. As a result, we model lateral tire force generation $F_y$ as a function
of both lateral tire slip $\alpha$ and the applied longitudinal force $F_x$, using a modified form of the Fiala tire model \cite{rami}. The front and rear tire slip are themselves functions of the vehicle state and are given by:

\begin{subequations}
\begin{align}
	\alpha_\mathrm{f} &= \beta + \frac{ar}{U_x} - \delta\\
	\alpha_\mathrm{r} &= \beta - \frac{br}{U_x}
\end{align}
\end{subequations}

\begin{figure}[tb]
\centering
\includegraphics[width=2.8in]{figures/tirecurves.eps}
\caption{Lateral tire force curve as a function of longitudinal force $F_x$ and lateral tire slip $\alpha$.}
\label{fig:coupledTires}
\end{figure}

The next step is to break up the control inputs into the closed-loop feedback term and the \textit{learned} component that is
modified on every lap by the ILC algorithm:
\begin{align}
\label{eqn:splitTerms}
F_x &= F^{\mathrm{FB}}_x + F^L_x\\
    &= -K_xv + F^L_x\\
\delta &= \delta_{\mathrm{FB}} + \delta_L\\
\delta &= -k_p(e + x_\mathrm{LA}\Delta\Psi) + \delta_L
\end{align}
where $K_x$ is the proportional speed tracking gain and $k_p$ and $x_\mathrm{LA}$ are the lookahead gains presented in \cite{kapania}.

The closed loop system dynamics are now given by: 
\begin{align}
\label{eq:fullNLCL}	
	\frac{de}{dt} &= \left(v+U^\mathrm{des}_x(s)\right)(\beta + \Delta\Psi) \\
	\frac{dv}{dt} &= \left(v+U^\mathrm{des}_x(s)\right)\beta r + \frac{-K_xv + F^L_x}{m} \\
	& - \frac{F_\mathrm{yf}(\alpha_F,F_x)(k_p(e + x_{LA}\Delta\Psi) + \delta_L)}{m}\\
	\frac{d\beta}{dt} &= \frac{F_\mathrm{yf}(\alpha_\mathrm{f}, F_x) + F_\mathrm{yr}(\alpha_\mathrm{r},F_x)}{m\left(v + U^\mathrm{des}_x(s)\right)} - r \\
	\frac{dr}{dt}     &= \frac{aF_\mathrm{yf}(\alpha_\mathrm{f}, F_x) - bF_\mathrm{yr}(\alpha_\mathrm{r}, F_x)}{I_z} \\
	\frac{d\Delta\Psi}{dt} &= r \label{eq:fullNLCL2}	
\end{align}

The nonlinear closed-loop dynamics must be converted into an affine, discrete-time dynamical system to apply conventional iterative learning
control algorithms. In our case, we have two system outputs ($y = [e \hspace{2mm} v]^T$) that are measured and two input signals to learn ($u = [\delta_L \hspace{2mm} F^L_x]^T$). 
Since we run the iterative learning control algorithm after seeing a trial of data, we can approximate
the dynamics in (\ref{eq:fullNLCL})-(\ref{eq:fullNLCL2}) by linearizing about the observed states and inputs $x_o, u_o$ from the first lap. The affine model is therefore given by: 
 \begin{align}
 \label{eq:ssMAT}
 \begin{bmatrix} \dot{e} \\ \dot{\beta} \\ \dot{r} \\ \dot{\Delta\Psi} \\ \dot{v}\end{bmatrix} &= \begin{bmatrix} A(t) \end{bmatrix} \begin{bmatrix} e \\ \beta \\ r \\ \Delta\Psi \\ v \end{bmatrix} + \begin{bmatrix} B(t) \end{bmatrix} \begin{bmatrix} \delta_L \\ F^L_x \end{bmatrix} + d(t) \\
 \begin{bmatrix} \dot{e} \\ \dot{v} \end{bmatrix}  &= \begin{bmatrix} 1 & 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 & 1 \end{bmatrix} x
 \end{align}

 The time-varying state space matrices $A(t)$ and $B(t)$ are given by Jacobian linearizations of the closed loop nonlinear dynamics (\ref{eq:fullNLCL}) about the observed states and inputs $x_o(t), u_o(t)$ from the last trial:
 \begin{align}
 \label{eq:C4nl}
 \begin{bmatrix} A(t) \end{bmatrix} &= \frac{\partial{f}}{\partial{x}}\Big|_{x_o(t)} \\
 \begin{bmatrix} B(t) \end{bmatrix} &= \frac{\partial{f}}{\partial{u}}\Big|_{u_o(t)}
 \end{align}

 While this multiple-input, multiple-output (MIMO) model captures the coupled behavior of the longitudinal and lateral inputs, it is tedious to compute,
 either numerically or analytically. If we make the assumption that the longitudinal and lateral inputs are decoupled and make small angle approximations, we obtain the following linear state equations for the lateral dynamics: 
\begin{subequations}
\label{eq:bmC4}
\begin{align}
	\dot{\beta} &= \frac{F_\mathrm{yf}+F_\mathrm{yr}}{mU_x} - r \qquad \dot{r} = \frac{aF_\mathrm{yf} - bF_\mathrm{yr}}{I_z} \label{bm1} \\
	\dot{e} &= U_x (\beta + \Delta\Psi) \qquad \Delta\dot{\Psi} = r - U_x\kappa \label{eq:bm2} 
\end{align}
\end{subequations} 
and the following first order equation for the longitudinal dynamics, assuming a simple point mass model with proportional
speed tracking feedback:
\begin{equation}
\dot{v} = \frac{-K_xv + F^L_x}{m}
\end{equation}
The resulting state matrices $A(t)$ and $B(t)$ for (\ref{eq:ssMAT}) are then simply block diagonal matrices consisting of the
linearized lateral dynamics and the first order longitudinal dynamics:

\begin{align}
\label{eqn:C4dyn}
 &A(t)  = \\ \nonumber
  &\left[\begin{smallmatrix}
  0 & U_x(t) & 0 & U_x(t) & 0\\ 
  0 & 0 & 1 & 0 & 0 \\ 
  \frac{-ak_\mathrm{p} \tilde{C}_\mathrm{f}(t)}{I_\mathrm{z}}  & \frac{-ak_\mathrm{p}x_\mathrm{LA}\tilde{C}_\mathrm{f}(t)}{I_\mathrm{z}}  & \frac{-a^2\tilde{C}_\mathrm{f}(t)-b^2\tilde{C}_\mathrm{r}(t)}{U_x(t)I_\mathrm{z}} & \frac{b\tilde{C}_\mathrm{r}(t) - a\tilde{C}_\mathrm{f}(t)}{I_\mathrm{z}} & 0 \\
  \frac{-k_\mathrm{p}\tilde{C}_\mathrm{f}(t)}{mU_x(t)}  & \frac{-k_\mathrm{p}x_\mathrm{LA}\tilde{C}_\mathrm{f}(t)}{mU_x(t)}  & \frac{b\tilde{C}_\mathrm{r}(t)-a\tilde{C}_\mathrm{f}(t)}{mU_x(t)^2}-1 & \frac{-\tilde{C}_\mathrm{f}(t) - \tilde{C}_\mathrm{r}(t)}{mU_x(t)} & 0 \\
  0 & 0 & 0 & 0 & -K_x
  \end{smallmatrix}\right] \\
&B(t) =\begin{bmatrix} 0 & 0 \\	 	0 & 0 		\\ 		\frac{a \tilde{C}_\mathrm{f}(t)}{I_\mathrm{z}} & 0 		\\ 		\frac{\tilde{C}_\mathrm{f}(t)}{mU_x(t)} & 0 		\\ 0 & 1 \end{bmatrix} 
\end{align}
The affine term $d(t)$ is given by: 
\begin{align}
d(t) = \left[\begin{matrix} 0 \\
               -\kappa(t) U_x(t) \\ 
			    \frac{a\tilde{C}_\mathrm{f}(t)\tilde{\alpha}_\mathrm{f}(t) - b\tilde{C}_\mathrm{r}(t)\tilde{\alpha}_\mathrm{r}(t) + a\tilde{F}_\mathrm{yf}(t) - b\tilde{F}_\mathrm{yr}(t)}{I_z}\\
				\frac{\tilde{C}_\mathrm{f}(t)\tilde{\alpha}_\mathrm{f}(t) + \tilde{C}_\mathrm{r}(t)\tilde{\alpha}_\mathrm{r}(t) + \tilde{F}_\mathrm{yf}(t) + \tilde{F}_\mathrm{yr}(t)}{mU_x(t)}\\
				0 \\
				0
				\end{matrix}\right]
\end{align}

Where $\tilde{F}_\mathrm{yf,r}(t)$ are the steady-state front and rear tire forces from the first lap, $\tilde{C}_\mathrm{f,r}(t)$ are the effective cornering stiffnesses at those tire forces, and $\tilde{\alpha}_\mathrm{f,r}(t)$ are the corresponding front and rear tire slips. While (\ref{eqn:C4dyn}) is written as a MIMO system for compactness, assuming decoupled lateral and longitudinal 
dynamics provides two single-input, single-output (SISO) systems.

\subsection{Lifted Domain Representation and ILC \newline Problem Statement}
\label{sec:liftedD}
Whether the coupled (\ref{eq:C4nl}) or decoupled (\ref{eqn:C4dyn}) dynamics are assumed, the
final modeling step is to apply standard discretization techniques to obtain dynamics in the following form:
\begin{align}
\label{C4ad}
x_{k+1} &= A_kx_k + B_ku_k + d_k \\
y_{k}   &= Cx_k
\end{align}
For a given lap of racing $j$, sensor measurements provide $N$ observations of both the lateral path deviation $e$ and longitudinal
speed tracking error $v$. These measurements can be stacked into a 2$N \times 1$ array: 
\begin{equation}
\mathbf{e}_j = \begin{bmatrix} e_1 & \hdots & e_N & v_1 & \hdots & v_N \end{bmatrix}^T
\end{equation} 
These measurement errors are related to the learned control inputs $\delta^L$ and $F^L_x$ as follows: 
\begin{align}
\label{eq:liftedDomain}
\mathbf{e}_j &= P\mathbf{u}^L_j + \mathbf{w} \\
\mathbf{u}^L_j &= \begin{bmatrix} \delta^L_1 & \hdots & \delta^L_N & F^L_{x1} & \hdots & F^L_{xN} \end{bmatrix}^T
\end{align}
The system dynamics modeled in the previous section are represented by the \textit{lifted-domain} dynamics matrix $P$,
which is $\mathrm{2}N \times \mathrm{2}N$ and given by:
\begin{equation}
\label{eq:C4bm}
P=\left[
\begin{array}{c|c}
P_{e\delta} & P_{eF} \\ \hline
P_{v\delta} & P_{vF} 
\end{array}\right]
\end{equation}
Where each submatrix in (\ref{eq:C4bm}) is $N \times N$ and represents the lifted-domain dynamics from a given input to a given output. 
Individual terms of the sub-matrices are given by:

\begin{equation}
\label{eq:how2Lift}
p_{lk} = \begin{cases} 0 &\mbox{if } l < k \\ 
C_yB_u(k) & \mbox{if } l = k \\
C_yA(l)A(l-1)\cdots A(k)B_u(k) &\mbox{if } l > k \end{cases}
\end{equation} 
Where $C_y$ is the row of $C$ in (\ref{C4ad}) corresponding to the desired output and $B_u$ the column of $B$ in (\ref{C4ad})
corresponding to the desired input. Note that
for the case of uncoupled lateral and longitudinal dynamics, the off-diagonal sub-matrices of $P$ are $[0]$ since we have
two SISO systems. The term $\mathbf{w}$ in (\ref{eq:liftedDomain}) is the unknown disturbance. Iterative
learning control relies on the assumption that this disturbance is the underlying cause of the observed errors $\mathbf{e}_j$, and that the disturbance, while unknown, is constant from lap to lap.

Given the error signal $\mathbf{e}_j$ for a given lap $j$, the iterative learning problem is to find the inputs $\mathbf{u}_{j+1}$ that will cancel out the tracking error on the next lap. The learned inputs are then applied, the observed error $\mathbf{e}_{j+1}$ is recorded,
and the process is repeated until the tracking error falls to a desired level. There is a wide body of literature on methods to determine $\mathbf{u}_{j+1}$ given $P$ and $\mathbf{e}_j$, but this paper will compute the ILC input for the next lap with the following formulation:

\begin{align}
 \label{eqn:ctrlLaw}
 \mathbf{u}^L_{j\!+\!1} = Q(\mathbf{u}^L_j - L\mathbf{e}_j)
\end{align}
where $Q$ is the $2N \times 2N$ \textit{filter} matrix, and $L$ is the $2N \times 2N$ \textit{learning} matrix. 
In the following two sections, the
matrices $Q$ and $L$ will be obtained by designing a proportional-derivative (PD) iterative learning controller as well as a quadratically
optimal (Q-ILC) learning controller.

\subsection{Proportional-Derivative Controller}
\label{sec:pdcontroller}

The proportional-derivative ILC computes the steering $\delta^L$ and force $F^L$ correction for the current lap $j$ based on the error and 
error derivative at the same time index $k$ from the previous lap:
\begin{align}
	\label{eq:PDlaw}
	\delta^L_{j}(k) &= \delta^L_{j\!-\!1}(k) - k_{p\delta}e_{j\!-\!1}(k) - k_{d\delta}(e_{j\!-\!1}(k) - e_{j\!-\!1}(k-1))\\
	    F^L_{j}(k) &=  F^L_{j-1}(k) - k_{pF}v_{j\!-\!1}(k) - k_{dF}(v_{j\!-\!1}(k) - v_{j\!-\!1}(k-1))
\end{align}
where $k_{p\delta}$ and $k_{pF}$ are proportional gains and $k_{d\delta}$ and $k_{dF}$ are derivative gains. In the lifted domain representation from (\ref{eqn:ctrlLaw}), the resulting learning matrix $L$ is given by
\begin{equation}
	L = \left[\begin{smallmatrix} 		-(k_{p\delta}+k_{d\delta}) &          &  0                            & 0 & \hdots & 0 \\ 
									   k_{d\delta}                     &  \ddots  &                               & \vdots & \ddots & \vdots\\ 
									   0                       &   k_{d\delta}    &    -(k_{p\delta}+k_{d\delta}) & 0 & \hdots & 0 \\
						               0 & \hdots & 0 & -(k_{pF}+k_{dF}) &          &  0\\
									   \vdots & \ddots & \vdots & k_{dF} &  \ddots  &                               &\\
									   0 & \hdots & 0  & 0 & k_{dF} & -(k_{pF} + k_{dF}) \\			
			\end{smallmatrix}\right]
\end{equation}

 The PD equation (\ref{eq:PDlaw}) determines $\delta^L$ only using lateral path deviation $e$ and $F^L_x$ using only the speed tracking error $v$. The filter matrix $Q$ is obtained by taking any filter transfer function and converting into the lifted domain via (\ref{eq:how2Lift}).  An important design consideration in choosing the two $k_p$ and $k_d$ gains is avoiding a poor lap-to-lap ``transient" response, where the path tracking error increases rapidly over the first several laps before
eventually decreasing to a converged error response $e_\infty$. This is a commonly encountered design requirement
for ILC systems, and can be solved by ensuring the following \textit{monotonic convergence} condition is met \cite{bristow}:

\begin{equation}
	\gamma \triangleq \bar{\sigma}(PQ(I-LP)P^{-1}) < 1
	\label{eq:MS}
\end{equation}
where $\bar{\sigma}$ is the maximum singular value. In this case, the value of $\gamma$ provides an upper bound on the change
 in the tracking error norm from lap to lap, i.e.

\begin{equation}
	||\mathbf{e}_\infty - \mathbf{e}_{j+1}||_2 \leq \gamma ||\mathbf{e}_\infty-\mathbf{e}_j||_2
\end{equation}

Fig.~\ref{fig:stabPlot} shows values of $\gamma$ for both an unfiltered PD controller ($Q = I$), and for a PD controller with a 2 Hz, first order low pass filter. The $\gamma$ values are plotted as a
contour map against the controller gains $k_{p\delta}$ and $k_{d\delta}$. Addition of the low-pass filter
assists with monotonic stability by removing oscillations in the control input generated when trying to remove small reference tracking errors after several iterations. Since the filtering occurs when generating a control signal for the next lap, the filter $Q$ can be zero-phase. 
The plot shown in Fig.~\ref{fig:stabPlot} is for the steering ILC design only, but the same analysis is possible for the longitudinal
ILC design as well. Because the $P$ matrix in general can change from iteration to iteration and will also
 change depending on the trajectory, making a general assertion or proof about stability of the iterative learning controller is difficult.

\begin{figure}[tb]
\centering
\includegraphics[width=2.45in]{figures/monotonicStability.png}
\caption[Values of convergence bound $\gamma$ vs. $k_{p\delta}$ and $k_{d\delta}$ for PD iterative learning controller]{Values of convergence bound $\gamma$ vs. $k_{p\delta}$ and $k_{d\delta}$ for PD iterative learning controller with (top) no filtering and (bottom) with a 2 Hz low-pass filter. Lower values of $\gamma$ correspond to faster convergence.
Shaded regions correspond to gains that result in system monotonic stability. The stability analysis in Fig.~\ref{fig:stabPlot} assumes the $P$ matrix is constant for all iterations and is generated assuming decoupled vehicle dynamics
 for a straight-line trajectory at a constant speed of 20 m/s.}
\label{fig:stabPlot}
\end{figure}

\subsection{Quadratically Optimal Controller}\label{sec:controller}
\label{sec:qopt}
An alternate approach to determining the learned steering and longitudinal force input is to minimize a quadratic cost function for the next lap:

\begin{align}
J_{j\!+\!1} = \textbf{e}_{j\!+\!1}^TT\textbf{e}_{j\!+\!1} + \textbf{u}^T_{j\!+\!1} R \textbf{u}^T_{j\!+\!1}+\Delta_{j\!+\!1}^TS\Delta_{j\!+\!1}
\label{eq:QILC}
\end{align}
where $\Delta_{j\!+\!1} = \mathbf{u}^L_{j\!+\!1} - \mathbf{u}^L_j$ and the $2N \times 2N$ matrices $T$, $R$, and $S$ are weighting matrices, each given
by a scalar multiplied by the identity matrix for simplicity.
This formulation allows the control designer to weight the competing objectives of minimizing the tracking errors $e$ and $v$, control effort $|\delta^L|$ and $|F^L_x|$, and change in the control signal from lap to lap.
While constraints can be added to the optimization problem, the unconstrained problem in (\ref{eq:QILC}) can be solved analytically \cite{bristow2008} to obtain desired controller and filter matrices:

\begin{subequations}
\label{eq:analSol}
\begin{align}
	Q &= (P^TTP + R + S)^{-1}(P^TTP + S)\\
	L &= (P^TTP + S)^{-1}P^TTP(T^{1/2}P)^{-1}T^{1/2}
\end{align}
\end{subequations}

An advantage of the quadratically optimal control design over the simple PD controller is that the controller matrices $Q$ and $L$ take the linearized, time-varying dynamics $P$ into account. 
This allows the iterative learning algorithm
to take into account changes in the steering dynamics due to changes in vehicle velocity.
Furthermore, if the fully coupled dynamics (\ref{eq:C4nl}) are used, the iterative learning algorithm also accounts for the
second-order effect of steering on the longitudinal dynamics and longitudinal force application on the lateral dynamics. 
 However, a disadvantage is that computing $\mathbf{\delta}^L$ in (\ref{eqn:ctrlLaw}) requires matrix multiplications with
the typically dense matrices $Q$ and $L$ for every lap, which can be computationally expensive for fast sampling rates.

\subsection{Simulated Results}
\label{sec:simres}

To test the feasibility of the PD and Q-ILC learning algorithms, the vehicle tracking performance over multiple laps is
simulated using the path curvature and speed profiles shown in Fig.~\ref{fig:simSetup}. To test the performance
of the controller at varying accelerations, four speed profiles are tested. Each profile is
generated with a different level of peak combined longitudinal/lateral acceleration, ranging from 5 $\mathrm{m/s^2}$ (below the limits)
up to 9.5 $\mathrm{m/s^2}$ (close to exceeding the limits). For accurate results, simulations were conducted using numerical integration with
 fully nonlinear equations of motion (\ref{eq:fullNL})-(\ref{eq:fullNL2}) and coupled lateral/longitudinal dynamics \cite{rami}.

\begin{figure}[bh]
\centering
\includegraphics[width= 3.35in]{figures/simSetup.eps}
\caption[Curvature profile used for ILC simulation.]{ (a) Curvature profile used for ILC simulation. (b) Velocity profiles generated for four different levels of longitudinal/lateral acceleration. }
\label{fig:simSetup}
\end{figure}

Simulated results for the root-mean-square (RMS) lateral path deviation is shown in Fig.~\ref{fig:simRes1}. The results
show the change in RMS error as the number of ILC iterations increase. Three different ILC controllers are tested. 
The first controller is the simple PD controller
with low-pass filter (\ref{eq:PDlaw}), and the second controller is the quadratically optimal ILC algorithm (\ref{eq:analSol}) assuming fully coupled dynamics (\ref{eq:C4nl}) in the plant
matrix $P$ (i.e. the full MIMO system).
The third controller is also the quadratically optimal ILC algorithm, but the $P$ matrix used in the optimization
assumes decoupled dynamics (\ref{eqn:C4dyn}) and therefore solves the lateral and longitudinal SISO problems separately.

\begin{figure}[h!]
\centering
\includegraphics[width=3.35in]{figures/simRes1.eps}
\caption[Simulated results for root-mean-square path tracking error at several values of vehicle acceleration]{Simulated results for root-mean-square path tracking error at several values of vehicle acceleration, with $T = R = I$ and $S = \mathrm{100} I$.
Results are plotted on a log scale.}
\label{fig:simRes1}
\end{figure}

Fig.~\ref{fig:simRes1} shows that both quadratically optimal ILC algorithms exponentially reduce the lateral
path tracking error as the number of learning iterations is increased. Overall, the RMS lateral tracking performance is
better at lower vehicle accelerations. This is unsurprising for two reasons. First, lateral path deviation in general increases for the lookahead steering feedback at higher accelerations. Second, our estimate of the vehicle dynamics contained in $P$
is based on linearization, and the vehicle dynamics at lower accelerations are mostly linear.

Fig.~\ref{fig:simRes2} shows the same results as Fig.~\ref{fig:simRes1}, but for the speed tracking performance. 
The overall trends are very similar. In both plots, there is very little difference between the coupled MIMO formulation and decoupled SISO 
formulation at low accelerations. This is expected, as the longitudinal and lateral dynamics are independent when the vehicle tires are not saturated.
At higher accelerations, there are small differences, but the overall RMS errors are still quite similar. A reason for this is the
nature of the speed profiles in Fig.~\ref{fig:simSetup}. The vehicle spends the majority of time either fully braking/accelerating or
turning at a constant velocity. There are only a few small transient regions where the vehicle needs significant amounts of both 
lateral and longitudinal acceleration. As a result, the need to account for the coupled dynamics may not be important in practice, especially
given the larger computation time needed when $P$ is dense and not block-diagonal.

\begin{figure}[tb]
\centering
\includegraphics[width=3.5in]{figures/simRes2.eps}
\caption[Simulated results for root-mean-square speed tracking error $v$ at several values of vehicle acceleration.]{Simulated results for root-mean-square speed tracking error $v$ at several values of vehicle acceleration, with $T = I$, $R = 0$, and $S = \mathrm{1e-7} I$.
Results are plotted on a log scale.}
\label{fig:simRes2}
\end{figure}

A final comment is that the proportional-derivative ILC algorithm performs relatively poorly. At low accelerations, the
speed and path tracking performance both improve initially, but fail to improve after the second learning iteration. At high
lateral and longitudinal accelerations, the tracking performance becomes even worse for the steering ILC in Fig.~\ref{fig:simRes1}.  
This is unsurprising given that the linearized plant dynamics $P$ are not explicitly accounted for in the selection of the PD gains. 
While the point mass model for the longitudinal dynamics is a relatively simple first order model, the lateral dynamics 
are fourth order and highly speed dependent. A simple PD approach for learning control is likely insufficient at the limits of handling.

\subsection{Experimental Results}
\label{sec:ch4ExpRes}

Experimental data for iterative learning control was collected over four laps at Thunderhill Raceway with the autonomous Audi TTS
described in Fig.~\ref{fig:carpic}. The controller parameters are shown in Table \ref{tb:ilcprm}.

% \begin{figure}[h]
% \centering
% \includegraphics[width=3.5in]{figures/expSetupC4.eps}
% \caption{Controller setup for experimental testing of iterative learning control. The differential GPS (DGPS) returns global coordinates}
% \label{fig:C4setup}
% \end{figure}

To save computation time, the learned controller inputs $\delta^L$ and $F^L_x$  are calculated at a 10 Hz update rate at the end of every race lap and stored as lookup tables in the controller. 
Since the the real-time control occurs at 200 Hz, at every time step the controller interpolates
the lookup table and applies the correct force and steer angle correction.

\begin{table}[tb]
\begin{center}
\caption{Vehicle Parameters}\label{tb:ilcprm}
\begin{tabular}{lccc}
Parameter & Symbol & Value & Units \\\hline
Lookahead Distance       & $x_\mathrm{LA}$          &  15.2 & $\mathrm{m} $ \\
Lanekeeping Gain         & $k_{\mathrm{LK}}$         & 0.053 & $\mathrm{rad\,m^{-1}}$\\
Lanekeeping Sample Time  & $t_s$                        & 0.005 & s\\
ILC Sample Time          & $T_s$                        & 0.1   & s\\
Speed Tracking Gain                & $K_x$           & 2500 & $\mathrm{Nsm^{-1}}$\\
Q-ILC Matrix (Path)            & $T$ and $R$              &  $I$      & - \\
Q-ILC Matrix  (Path)           & $S$                       & 100\,$I$  & - \\
Q-ILC Matrix (Speed)            & $T$              &  $I$      & - \\
Q-ILC Matrix (Speed)            & $R$              &  $0$      & - \\
Q-ILC Matrix  (Speed)           & $S$                       & 1e-7\,$I$  & - \\\hline
\end{tabular}
\end{center}
\end{table}

Fig.~\ref{fig:expRes1} shows the applied iterative learning signals and resulting path tracking error over four laps using the SISO
quadratically optimal learning algorithm. The car is driven
aggressively at peak lateral/longitudinal accelerations of 8 $\mathrm{m/s^2}$. On the first lap, despite the 
incorporation of a feedforward-feedback controller operating at a high sampling rate, several spikes in tracking error are visible due to transient vehicle dynamics.

However, the iterative learning algorithm is able to significantly attenuate these transient spikes after just two or three laps. 
One of the most important features of the time series plot is that the learned steering corrections are applied slightly before a lateral path deviation was observed the prior lap (i.e. the steering corrections \textit{lead} the observed error). This is because the learning
algorithm has knowledge of the system model and knows that a steering correction must be applied a few meters early to cancel a path deviation
further down the road.

\begin{figure}[h!]
\centering
\includegraphics[width=3.5in]{figures/expRes1.eps}
\caption{Experimental results for path tracking error with Q-SISO learning controller, at peak lateral accelerations of 8 $\mathrm{m/s^2}$.}
\label{fig:expRes1}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/expRes2.eps}
\caption{Experimental results for speed tracking error with Q-SISO learning controller, at peak lateral accelerations of 8.5 $\mathrm{m/s^2}$.}
\label{fig:expRes2}
\end{figure}
Fig.~\ref{fig:expRes2} shows the iterative learning signals for the longitudinal speed control at a slightly
higher acceleration of 8.5 $\mathrm{m/s^2}$. Again, in just
two to three iterations, significant lags in the speed tracking are attenuated from $s = 800-900$ meters and $s = 1200-1300$ meters. Additionally, the controller also acts to slow the car down when $v > 0$ and the vehicle exceeds the speed profile. This is also desirable
from a racing perspective as it it prevents the vehicle from exceeding the friction limit. Oscillations in speed tracking
performance are visible in Fig.~\ref{fig:expRes1}(a) from 900 - 1100 meters and 1300 - 1400 meters. These are generally undesirable, and further tuning of the filter matrix $Q$ is possible
to remove rapid changes in the learned force input.

 Notice that Fig.~\ref{fig:expRes2} has
several regions where the speed error $v << 0$. These are straight regions of the track where is no true planned speed
because the desired longitudinal action is to fully apply the throttle and go as fast as physically possible. For convenience, the ILC is programmed to saturate the longitudinal
learning signal to 8000 Newtons, although a more elegant solution is to switch the ILC controller off on straight portions of the track.

In Fig.~\ref{fig:expRes3}, root-mean-square tracking results are shown for a range of peak vehicle accelerations. 
The results show that at lower vehicle accelerations,
the initial speed and lateral tracking errors (Iteration 0) are smaller, as the built-in feedback-feedforward controller 
performs better. However, as the speed profile becomes more aggressive, the path and speed tracking degrades in the presence
of highly transient tire dynamics. Regardless of the initial error, application of iterative learning control 
reduces the trajectory tracking errors significantly over just 2 or 3 laps. At an acceleration of 8.5 $\mathrm{m/s^2}$, for example, the RMS
lateral tracking error is around 3 cm, on the order of the expected RMS error from the GPS position sensor. 
On some tests, the RMS tracking error 
increases slightly from
Lap 2 to Lap 3, and for the case where vehicle acceleration is 9 $\mathrm{m/s^2}$, the lateral tracking error is constant from Lap 1 to Lap 2
before decreasing further in Lap 3. While not predicted in simulation, this behavior likely occurs because the repeating disturbance
from lap-to-lap is not exactly constant, especially as the vehicle approaches the handling limits. More refined tuning of the gain matrices may be able to prevent 
this RMS error increase, or the ILC algorithm can be stopped after several iterations once the tracking performance is acceptable.

Experimental results in this section were only given for the quadratically optimal controller with decoupled (SISO) dynamics. 
The PD iterative learning controller was not tested due to the relatively worse simulation performance, and the quadratically optimal controller with coupled dynamics provided no clear benefit in simulation but a much longer computation time.

\begin{figure}
\centering
\includegraphics[width=2.8in]{figures/expRes3.eps}
\caption[Experimental RMS tracking error for the Q-SISO learning controller at several levels of lateral acceleration. ]{Experimental RMS tracking error for the Q-SISO learning controller at several levels of lateral acceleration. (a) RMS lateral tracking
error as a function of lap number/iteration number. (b) RMS speed tracking error as a function of lap/iteration number. Note that lap 0 corresponds
to the baseline case where iterative learning control is not applied.}
\label{fig:expRes3}
\end{figure}

\section{Iterative Trajectory Modification}
\subsection{Effect of Tire Slip Norm on Lap Time}
\label{sec:etu}

Fig.~\ref{fig:t2} shows the complicated effect of driving at different levels of lateral and longitudinal acceleration for region \circled{2} in Fig.~\ref{fig:expErrors}. Recall that the
speed profile is generated by assuming a global value of tire friction $\mu$, which is directly related to the acceleration norm of
the desired speed profile ($\sqrt{a_x^2 + a_y^2}$) by a factor of \mbox{$g =$ 9.81 $\mathrm{m/s^2}$}. Fig.~\ref{fig:t2}(a) confirms that higher levels
of $\mu$ result in strictly faster speed profiles for the same turn.

\begin{figure}
\centering
\includegraphics[width=2.8in]{figures/t2res.eps}
\caption[Desired speed for varying levels of $\mu$ - region 2]{(a) Desired speed for varying levels of $\mu$ (b) Actual speed for varying levels of $\mu$. Asterisks correspond to 
regions where $\zeta > 1$. (c) Tire slip norm for varying levels of $\mu$. }
\label{fig:t2}
\end{figure}

However, selecting a more aggressive value of $\mu$ does not necessarily entail a faster experimental lap time. Consider the actual speeds of the vehicle in
Fig.~\ref{fig:t2}(b) when trying to experimentally follow three different speed profiles. For the case where $\mu = 0.9$, the vehicle completes
the turn without fully utilizing the tire's capability and achieves relatively low velocities. For the extreme case where $\mu = 0.95$, the
vehicle enters the turn at a high speed but then begins to slide as $\zeta = 1.6$. While not shown in the plot, the saturation
is occurring primarily at the front tires, causing an understeer that can cause the car to skid off the track. Completing the lap therefore requires a stabilizing
action from the stability controller to slow the car down and regain control of the vehicle. As a result,
when the vehicle accelerates at the end of the turn, the actual vehicle speed at $\mu = 0.95$ is significantly slower than the case where $\mu = 0.9$!
A final ``just right" value of $\mu = 0.93$ was also tested experimentally for this turn, and while the car does slide a bit at this level of driving (peak $\zeta = 1.3$),
the needed stabilizing action is significantly smaller and the vehicle exits the turn with the highest speed.

 \begin{figure}[tb]
\centering
\includegraphics[width=2.8in]{figures/t3res.eps}
\caption[Desired speed for varying levels of $\mu$ - region 3]{(a) Desired speed for varying levels of $\mu$. (b) Actual speed for varying levels of $\mu$. Asterisks correspond to 
regions where $\zeta > 1$. (c) Tire slip norm for varying levels of $\mu$. }
\label{fig:t3}
\end{figure}

Unfortunately, the best value of $\mu$ is not constant throughout the track. Fig.~\ref{fig:t3} shows the same data plotted for region \circled{3} of the 
Thunderhill Raceway. In this case, even with an atypically high value of 
$\mu = 0.97$, the slip norm of the vehicle tires is relatively low, and the value of $\mu$ that optimizes the completion time for this turn could 
be even larger.

\subsection{Naive Method: Greedy Algorithm}
\label{sec:greedy}

Section \ref{sec:etu} shows that to  minimize the overall lap time, there is a need to generalize the speed profile such that different portions of the track can be driven with different
values of $\mu$. In other words, \textbf{the problem is to learn the ``friction profile" $\mu^\star(s)$ along the path that minimizes the experimental vehicle lap time.}

The simplest approach to finding $\mu(s)$ is a greedy algorithm where a set of experimental data is collected for a variety of different
speed profiles, each corresponding to a different value of $\mu$ and therefore a different acceleration level. The greedy approach is then to discretize
the track into a number of small sections and pick the value of $\mu$ for each section that corresponds to the highest observed experimental velocity.
The final desired velocity profile is then generated using the numerical integration approach presented in \cite{subosits2015autonomous}. This method does not require a single 
value of friction across the whole track, and generates a smooth velocity profile even when the peak acceleration limits vary from point to point.

 \begin{figure}[tb]
\centering
\includegraphics[width=2.8in]{figures/turn2greedy.eps}
\caption[Greedy algorithm results for region 2.]{(a) Desired speed for varying levels of $\mu$ (b)``Greedy" value of $\mu$ as a function of distance along track. Asterisks denote
region of track where vehicle is understeering.}
\label{fig:t2g}
\end{figure}

A plot showing the results of applying the greedy algorithm for section \circled{2} is shown in Fig.~\ref{fig:t2g}. The flaw in simply
selecting $\mu$ based on the highest speeds is apparent at $s = 1050 \mathrm{m}$. The greedy algorithm suggests the vehicle mostly drive at $\mu$ = 0.95,
but then switch to driving at $\mu = 0.93$ as soon as the vehicle begins to slide. Switching to a less aggressive velocity profile at this
point is impossible to achieve in practice, because the vehicle is already fully sliding and has no control until the vehicle slows down.
As a result, the greedy algorithm fails to capture the hidden cost of a large understeer, which results in a period of time where the speed must
inevitably drop.

\subsection{Framing Trajectory Learning as Search Problem}
\label{sec:framingTL}

Given the inadequacy of the greedy algorithm, a more sophisticated approach is necessary to learn $\mu^\star(s)$ from
experimentally observed data. This section frames the desire to find the minimum time $\mu^\star(s)$ as a tree search problem.
Consider discretizing the racing path into $N$ evenly spaced segments. For example, on the Thunderhill Raceway with
$\Delta s = 5 \mathrm{m}$, $\textbf{s} = [0 \hspace{2mm} 5 \hdots s_k \hdots 4495 \hspace{2mm} 4500]$ for $k = 0 \hdots N$, with $N = 901$. For each
path distance $s_k$, there are $M_k$ velocity and $M_k$ tire slip observations from experimental data, each corresponding to a different $\mu$. 
For example, looking at Fig.~\ref{fig:t2}, if $k = 191$, $s_k =  950 \mathrm{m}$, $M_k = 3$, the velocity and slip norm $\zeta$ observations $U_k(\mu)$ and $Z_k(\mu)$ are as follows:

\begin{table}[h]
\begin{center}
\caption{$U_k(\mu)$ and $Z_k(\mu)$ for $k =$ 191}\label{tb:laptimes}
\begin{tabular}{ccc}
$\mu $&$ U_x(m/s) $& $\zeta $\\\hline
0.9 & 28.42 & 0.55\\
0.93& 28.91 & 0.63\\
0.95& 29.14 & 0.66\\\hline
\end{tabular}
\end{center}
\end{table} 
 $M_k$ is not necessarily the same for all $k$ to account for experimental trials that do not cover the full lap. For safety and time
 reasons, some parts of the track will only have experimental data collected at two different friction values, while others may have five or six.

 Nodes of the search tree are then defined as a two-element state tuple, with the first state element being $s_k$ and the second element being the current
 friction coefficient $\mu_k$. Since the car must start from the beginning of the path, the first state has $s_0 = 0$.
 The $M_k$ edges from a given node correspond to actions that the car can take at $s = s_k$. In this case, the ``action" is the next value of $\mu$,
 and the successor states are $(s_{k+1}, \mu_1) ... (s_{k+1}, \mu_{M_k})$. A diagram of the search tree is shown in Fig.~\ref{fig:stree}.

\begin{figure}
\centering
\includegraphics[width=2.8in]{figures/searchTree.png}
\caption{Sample search tree for Thunderhill race track where there are only 3 experimentally observed full laps at $\mu =$ 0.9, 0.93, 0.95.}
\label{fig:stree}
\end{figure}

Each edge is associated with a travel cost $c_t$. The travel cost for a given edge is the amount of time it takes to go from
$s = s_k$ to $s = s_{k+1}$ driving at the friction coefficient $\mu$ associated with that edge. As illustrated in Fig.~\ref{fig:costDiag}, the cost is estimated from the experimentally
observed data assuming linear acceleration between points. The travel cost from node $(s_k, \mu_k)$ to $(s_{k+1}, \mu_{k+1})$ can therefore be expressed mathematically with trapezoidal integration of the straight-line velocity profile:
\begin{align}
c_t(s_k, s_{k+1},\mu_k, \mu_{k+1}) &= \ln \frac{a_x(\Delta s) + U_k(\mu_k)}{a_x} - \ln \frac{U_k(\mu_k)}{a_x} \\
a_x &= \frac{U_{k+1}(\mu_{k+1}) - U_k(\mu_k)}{\Delta s}
\end{align}

 \begin{figure}[tb]
\centering
\includegraphics[width=3.5in]{figures/costStructure.eps}
\caption[Illustration of travel cost.]{Illustration of travel cost. Current state is (855, 0.9). Assuming $\Delta s$ is 15 meters, cost is time to travel from this node
to one of the three possible successor nodes, depending on the action taken.}
\label{fig:costDiag}
\end{figure}

In addition to the travel cost, there is also a \textit{switching} cost $c_s$ associated with switching to a different value of $\mu$. These
are determined by the observed tire slip norm measurements $Z_k(\mu)$. The switching cost is expressed mathematically as:

\begin{align}
\label{eq:swCost}
c_s(s_k, s_{k+1},\mu_k, \mu_{k+1}) &= \\
&\mathds{1}\left(\mu_{k+1} \neq \mu_{k}\right)\left(\lambda  + \infty \times\mathds{1}\left(Z_k(\mu_k) > 1\right) \right)
\end{align}

Where $\mathds{1}$ is the indicator function. The switching cost function (\ref{eq:swCost}) implies that an action will never
incur a switching cost if the selected value of $\mu$ is unchanged from the previous selection. If the value of $\mu$ does change, there
is a small switching penalty $\lambda$, chosen by trial and error to discourage the search algorithm from changing the friction profile to gain a
trivial decrease in lap time. Additionally, there is a very large (infinite) switching penalty if the search algorithm attempts to change the
friction profile while the vehicle's tires are saturated ($\zeta > 1$). This reflects the physical inability of the car to control its
trajectory while sliding and is what separates the search algorithm from the greedy algorithm. A diagram demonstrating (\ref{eq:swCost})
is shown in Fig.~\ref{fig:swCost}.

 \begin{figure}[tb]
\centering
\includegraphics[width=2.5in]{figures/switchCost.eps}
\caption[Costs when vehicle is and is not sliding.]{(a) Costs when vehicle currently is not sliding. The vehicle can switch to a more aggressive velocity profile, paying a small
switching penalty, or can continue on the current profile. (b) Costs when vehicle is currently sliding. Vehicle has no choice but to continue on current
trajectory. Asterisks denote regions where $\zeta > 1$ and the vehicle is sliding.}
\label{fig:swCost}
\end{figure}

\subsection{A* Search Algorithm and Heuristic}
\label{sec:ch5astarheuristic}

Once the tree is mathematically defined in terms of the start state, nodes, edges, and costs, the search problem is to find the sequence
of actions from the start node to any terminal node that minimizes the total cost. In our case, the start state, nodes, edges
and costs were defined in the previous section, and a terminal node is any node at the end of the path (i.e. $k = N$). The sequence
of actions in our case is the friction profile $\mathbf{\mu} = [\mu_1 \hdots \mu_k \hdots \mu_N]$, which determines how
aggressively the vehicle will drive on every part of the track. The total cost is the sum of all individual travel costs $c_t$ and switching costs
$c_s$, and has intuitive units of time.

Minimum-cost tree spanning algorithms (e.g. breadth-first search, Dijkstra's algorithm, etc.) are a well known subject and a thorough description
can be found in \cite{aibook}. For the purpose of solving this search problem, the A* search algorithm is used. 
Like breadth-first search, the A* algorithm is guaranteed to find the lowest cost path from a start node to a goal node, 
but uses a priority queue data structure to more efficiently explore the search tree. Frontier leaf nodes $n$ being
considered for exploration are ranked according to the following modified cost:

\begin{equation}
f(n) = g(n) + h(n)
\end{equation}
where $g(n)$ is the true cost to go from the start node to node $n$. The function $h(n)$ is a heuristic estimate of the cost to get from node $n$ to any goal state (i.e. to the end of the
path). For A* to be guaranteed to find the shortest path, $h(n)$ must be \textit{admissible}, meaning that $h(n)$ must underestimate the
true cost of getting to the end of the path.

In our case, we have a very intuitive heuristic function $h(n)$ for the A* implementation. In Sec. \ref{sec:greedy}, the greedy approach
was discussed. Define $\mathbf{U_\mathrm{g}} = [U_\mathrm{g}(1) \hdots U_\mathrm{g}(k) \hdots U_\mathrm{g}(N)]$ to be the highest observed
experimental speed for each index $s_k$. Because we know tracking this greedy profile is physically impossible due to vehicle sliding, time estimates
from this profile will always underestimate the true cost. We therefore define $h(n)$ as follows:

\begin{equation}
\label{eq:heuristic}
h(n) = \int^{s_N}_{s_n} \frac{1}{U_g(s)} ds
\end{equation}
where $s_n$ is the value of $s$ corresponding to node $n$ and $s_N$ is the total length of the track. While (\ref{eq:heuristic}) is an integral equation, it can be solved for the discrete
array $\mathbf{U}_\mathrm{g}$ via trapezoidal numerical integration.

\subsection{A* Implementation and Results}

Because the A* algorithm relies on experimental observations to learn $\mu^\star(s)$, experimental data was collected over several trials, with
each trial consisting of a speed profile generated with a constant $\mu$ value over the track. The $\mu$ values chosen for
experimental data collection were 0.85, 0.9, 0.92, 0.93, 0.94, 0.95, and 0.97. Ideally, each speed profile would be tested experimentally for a 
full lap at high speed. However, due to safety and time constraints associated with collecting high speed race data, only the speed profiles corresponding to $\mu = 0.92$ and $\mu = 0.94$ were observed
over the whole track. The other speed profiles were only tested on sections of the track. Fig.~\ref{fig:muCoverage} shows the experimental data coverage.

\begin{table}[h]
\begin{center}
\caption{Search Algorithm Information}\label{tb:astarparams}
\begin{tabular}{lccc}
Parameter & Symbol & Value & Units \\\hline
Track Length       & $L$           &  4500 & $\mathrm{m} $ \\
Discretization Length               & $\Delta s$  & 5 & $\mathrm{m}$\\
Number of Points  & $N$                        & 901 & -\\
Switching Cost          & $\lambda$                        & 0.05   & s\\\hline
CPU Solution Time           &                                  &  70    & s \\
Nodes Explored              &                                  &  6887  & - \\\hline
\end{tabular}
\end{center}
\end{table}

 \begin{figure}[tb]
\centering
\includegraphics[width=2.45in]{figures/muCoverage.eps}
\caption[Coverage of experimental data]{Coverage of experimental data. For speed profiles corresponding to $\mu = 0.92$ and $\mu = 0.94$, experimental data of the autonomous race vehicle was observed
over the whole track. For safety and time constraints, the other speed profiles were only tested on sections of the track.}
\label{fig:muCoverage}
\end{figure}

After collecting the experimental data, the A* algorithm was applied to learn the optimal friction profile $\mu^\star(s)$. Parameters for the algorithm
are shown in Table~\ref{tb:astarparams}. To interface more efficiently with experimental data files, the search algorithm was implemented in MATLAB using
custom code instead of a standard search algorithm library. The entire search process took approximately 70 seconds on a core i7 laptop machine,
exploring 6887 nodes in the process. However, since MATLAB is not designed for computational efficiency in tree-based search algorithms, a C++ implementation would likely
be several orders of magnitude faster.

 The resulting $\mu^\star(s)$ profile associated with the minimum lap time on Thunderhill Raceway is shown in Fig.~\ref{fig:muprof}. For  comparison, the A* solution is plotted against the greedy algorithm solution. Because of the incorporation of switching costs,  the A* algorithm switches $\mu$ values only when necessary to achieve a nontrivial increase in lap time, and only switches $\mu$ when  the vehicle is not sliding. The same results are plotted on a map of the track in Fig.~\ref{fig:mumap}. A satisfying observation is that the optimal profile reduces $\mu$ to 0.93 in section \circled{2} to be more conservative and increases $\mu$ to 0.97 in section \circled{3} to be more aggressive. This matches our observations about tire slip norm $\zeta$ in Sec.~\ref{sec:etu}.

 Predicted lap time results are shown in Table~\ref{tb:predresults}. Notice that the A* predicted lap time is slightly
 slower than the greedy algorithm prediction, which is expected given the physical infeasability of the greedy algorithm assumptions. The  results also indicate that a significant lap time improvement can be expected over a velocity profile generated with a constant $\mu$.

 \begin{figure}[tb]
\centering
\includegraphics[width=3.5in]{figures/muprofile.eps}
\caption{Minimum time $\mu(s)$ profile for Thunderhill, for both the A* solution and greedy algorithm solution.}
\label{fig:muprof}
\end{figure}

 \begin{table}[h]
\begin{center}
\caption{Lap Times}\label{tb:predresults}
\begin{tabular}{l|c}
Driver          & Lap Time \\\hline
Constant $\mu(s)$ = 0.94 & 139.2 \\
Constant $\mu(s)$ = 0.92 & 139.4 \\\hline
Greedy $\mu_g(s)$ (Prediction) & 136.4 \\
A*     $\mu^*(s)$ (Prediction) & 136.8 \\\hline
\end{tabular}
\end{center}
\end{table}

\begin{figure}[tb]
\centering
\includegraphics[width=3.5in]{figures/mumap.eps}
\caption{Minimum time $\mu^\star(s)$ profile from A* algorithm plotted on map of Thunderhill.}
\label{fig:mumap}
\end{figure}

Fig.~\ref{fig:mumap} provides interesting insights about what the A* algorithm is learning. Section \circled{2} is a 
long, sweeping turn with mostly steady-state cornering dynamics, so the optimal friction value of .93 is in the middle
of the range of possible values and representative of the average friction between the tires and the track. Sections \circled{3}
and \circled{7} represent short turns followed by a turn in the opposite direction. For these turns, the algorithm has learned it is better to drive a little
 faster than the true friction limit would dictate, because by the time the vehicle begins to understeer, the turn is already complete and
 the vehicle can reverse the steering quickly while following the desired path. Section \circled{8} occurs before a long straight section of the track where recovering
from an understeer would result in a significantly lower speed on the fastest part of the track. As a result, the algorithm's planned acceleration is more conservative. Finally, 
the section with the lowest $\mu(s)$ occurs on a part of the track with significant lateral weight transfer. In general, lateral weight transfer reduces the cornering forces available
to the vehicle, but this effect was not captured in the trajectory planning phase, which assumed a planar model with coupled left and right tires. In summary, the A*
algorithm allows the car to learn subtle but important driving behaviors that are not easily captured through simulation.

\subsection{Experimental Validation}
The best validation of the A* algorithm is to experimentally drive the optimal velocity profile $U^\star_x(s)$ generated from $\mu^\star(s)$. From Fig.~\ref{fig:muprof}, this velocity profile will have accelerations as low as 9.0 $\mathrm{m/s^2}$ on some turns, and as high as 9.7 $\mathrm{m/s}$ on others. Fig.~\ref{fig:muexpres} shows autonomous experimental data from 
driving $\mu^\star(s)$ compared to one other experiment\footnote{There was a minor change between the optimal $\mu$ profile from Fig.~\ref{fig:muprof} and what was tested experimentally. For section \circled{2}, the value of $\mu$ was set to 0.92 as opposed to 0.93. This was a safety measure taken based on preliminary testing.}. The comparison experiment is a fully autonomous test with a constant $\mu = 0.94$.

\begin{figure}[tb]
\centering
\includegraphics[width=3.4in]{figures/expAstarResults.eps}
\caption{Time difference between experimental dataset collected with A* result $\mu^*(s)$ and constant friction velocity profile at $\mu = 0.94$. Negative time distance corresponds to A* result being ahead. (b) Experimental velocities between both experimental datasets.
(c) Tire slip norm measurements for both datasets.}
\label{fig:muexpres}
\end{figure}

 The experimental results show solid performance of the learned friction profile $\mu^\star(s)$. From Fig.~\ref{fig:muexpres}(a), the lap time using the learned friction profile is roughly 1.5 seconds faster than the lap time from the constant friction profile. A significant part of this improved performance comes through more efficient friction usage.
 For example, in sections \circled{2} and \circled{8}, learning to drive more cautiously enables the tire slip norm $\zeta$ to drop closer to 1, avoiding
 a costly understeer. In sections \circled{3} and \circled{7}, the A* algorithm has learned that more aggressive driving is possible, increasing $\zeta$ closer to 1.\footnote{Note that due to time constraints, both datasets were all taken on different dates, meaning some variation in weather and tire conditions is possible for each test.}

\section{Conclusion}
This paper presents two complementary approaches to gradually improve the driving performance of an autonomous race car 
over time. Iterative learning control (ILC) formulations gradually determine the proper steering and throttle inputs to precisely track the desired racing trajectory. Experimental results at combined vehicle accelerations of 9 $\mathrm{m/s^2}$  indicate that the proposed algorithm can rapidly attenuate trajectory following errors over just two or three laps of racing (\S \ref{sec:ch4ExpRes}). This paper also presents a tree-search algorithm to minimize experimental lap times by learning different acceleration limits for each turn on the track. An A* search algorithm is devised to search through experimental data and find the best value of $\mu$ for each portion of the track in order to globally minimize the resulting lap time. Key developments of this algorithm include designing an appropriate A* heuristic to minimize the needed computation time and designing the cost function to account for the physical difficulty of altering the vehicle's trajectory while understeering or oversteering.

\section*{Acknowledgment}

This research is supported by the National Science Foun-
dation Graduate Research Fellowship Program (GRFP). The
authors would like to thank the members of the Dynamic
Design Lab at Stanford University and the Audi Electronics
Research Lab.

\bibliographystyle{IEEEtran}
\bibliography{Bibliography}

\end{document}

### 2019-02-13 journal/TCST2019/figures/getIndV.m

@@ -1,0 +1,11 @@
+function [ indv ] = getIndV( s )
+
+
+ind2 = find(s > 600 & s < 700);
+ind3 = find(s > 800 & s < 1150);
+ind4 = find(s > 1250 & s < 1570);
+indv = union(ind2, ind3); indv = union(indv, ind4);
+
+
+end
+

### 2019-02-13 journal/TCST2019/figures/plotMuCoverage.m

@@ -1,0 +1,19 @@
+clear all close all; clc;
+addpath('C:\Users\ddl\Documents\localTestData\searchData');
+
+figure;
+hold on; 
+grid on;
+
+s = [0 1300; 0 4500; 0 4500; 2300 4200; 3300 4000; 2300 3300; 2300 3300; 2300 3300; 2300 3300 ; 2300 3300; 530 1165;0 1270; 390 770; 0 1270 ];
+mu =[.9 .9; .92 .92; .94 .94; .9 .9;    .85   .85;    .97 .97;  .95  .95; .93 .93;     .92  .92;  .85   .85; .85 .85;.93 .93; .97 .97; .95 .95];
+
+
+for i = 1:14
+
+    
+    plot(s(i,:), mu(i,:), 'k','LineWidth',3)
+end
+
+xlabel('Distance Along Path (m)','FontSize',14);
+ylabel('\mu','FontSize',18)

### 2019-02-13 journal/TCST2019/figures/plotSimResults.m

@@ -1,0 +1,220 @@
+clear all; close all; clc;
+load sim50.mat
+
+subplot(2, 1, 1)
+grid on; hold on;
+plot(path.s, path.K,'k','LineWidth',2)
+xlim([0 675])
+ylim([-.022 .031])
+ylabel('Curvature (1/m)')
+subplot(2, 1, 2)
+grid on; hold on;
+plot(vp.s, vp.Ux,'k','LineWidth',2)
+load sim70.mat
+plot(vp.s, vp.Ux,'k--','LineWidth',2)
+load sim85.mat
+plot(vp.s, vp.Ux,'r-','LineWidth',2)
+load sim95.mat
+plot(vp.s, vp.Ux,'g','LineWidth',2)
+legend('|a| = 5 m/s^2','|a| = 7 m/s^2','|a| = 8.5 m/s^2','|a| = 9.5 m/s^2')
+xlim([0 675])
+ylabel('Desired Velocity')
+xlabel('Distance Along Path (m)')
+
+%%
+
+close all; clear all; clc;
+
+load sim50.mat
+subplot(2, 2, 1);
+for i = 1:N
+    pde(i) = SIM_PD(i).eRMS;
+    qse(i) = SIM_QS(i).eRMS;
+    qsm(i) = SIM_QM(i).eRMS;
+end
+
+semilogy(0:4, pde,'ko-','MarkerFaceColor','k','LineWidth',2);
+hold on; grid on;
+semilogy(0:4, qsm,'bs-','MarkerSize',8,'LineWidth',2);
+semilogy(0:4, qse,'r*-','MarkerFaceColor','r','LineWidth',2);
+legend('PD', 'Q-MIMO','Q-SISO');
+
+%xlabel('Iteration Number','FontName','Arial','FontSize',14)
+ylabel('e_{RMS}(m)','FontName','Arial','FontSize',14)
+title('|a| = 5 m/s^2','FontName','Arial','FontSize',14)
+
+load sim70.mat
+subplot(2, 2, 2);
+for i = 1:N
+    pde(i) = SIM_PD(i).eRMS;
+    qse(i) = SIM_QS(i).eRMS;
+    qsm(i) = SIM_QM(i).eRMS;
+end
+
+semilogy(0:4, pde,'ko-','MarkerFaceColor','k','LineWidth',2);
+hold on; grid on;
+semilogy(0:4, qsm,'bs-','MarkerSize',8,'LineWidth',2);
[…]

### 2019-02-13 journal/TCST2019/figures/plotTireCurve.m

@@ -1,0 +1,21 @@
+clear all; close all; clc;
+addpath(genpath('C:\Users\ddl\Desktop\nkapania\simulation\common'));
+
+veh = getVehicle('nonlinear','closest');
+Fx = 1e3*[0; 2; 4; 6; 8];
+N = 100;
+Fy = zeros(N, 1);
+alpha = linspace(0, 8*pi/180, N);
+
+figure;
+
+for j = 1:len(Fx)
+    for i = 1:N
+        Fy(i) = abs(coupledTireForces(alpha(i), Fx(j), veh, 'front'));
+    end
+    hold on; plot(alpha*180/pi, Fy,'k','LineWidth',2);
+end
+
+grid on; hold on;
+xlabel('Lateral Tire Slip \alpha (deg)','FontSize',14);
+ylabel('Lateral Tire Force Magnitude (N)','FontSize',14);

### 2019-02-13 journal/TCST2019/figures/plotTurn2.m

@@ -1,0 +1,72 @@
+close all; clear all; clc;
+
+addpath('C:\Users\ddl\Documents\localTestData\raceData')
+load p1 %mu = .9
+x0 = [840 1236];
+aF = alphaF/(8*pi/180);aR = alphaR/(6*pi/180);kF = kappaF/.1;kR = kappaR/.1;zetaF = sqrt(aF.^2 + kF.^2);zetaR = sqrt(aR.^2 + kR.^2);zeta = max(zetaF, zetaR);
+
+subplot(3, 1, 1)
+plot(s, UxDesired,'k','LineWidth',2)
+grid on; hold on;
+xlim(x0)
+
+subplot(3, 1, 2)
+plot(s, Ux,'k','LineWidth',2)
+grid on; hold on;
+ind = (FxFB~=0);
+plot(s(ind), Ux(ind),'k*')
+xlim(x0)
+
+subplot(3, 1, 3)
+plot(s, zeta,'k','LineWidth',2)
+grid on; hold on;
+xlim(x0)
+
+
+load p25 %mu = .93; 
+aF = alphaF/(8*pi/180);aR = alphaR/(6*pi/180);kF = kappaF/.1;kR = kappaR/.1;zetaF = sqrt(aF.^2 + kF.^2);zetaR = sqrt(aR.^2 + kR.^2);zeta = max(zetaF, zetaR);
+
+c = 'b';
+subplot(3, 1, 1)
+plot(s, UxDesired,c,'LineWidth',2)
+grid on; hold on;
+xlim(x0)
+
+subplot(3, 1, 2)
+plot(s, Ux,c,'LineWidth',2)
+grid on; hold on;
+ind = (FxFB~=0);
+plot(s(ind), Ux(ind),[c '*'])
+xlim(x0)
+
+subplot(3, 1, 3)
+plot(s, zeta,c,'LineWidth',2)
+grid on; hold on;
+xlim(x0)
+
+load p26 %mu = .95; 
+aF = alphaF/(8*pi/180);aR = alphaR/(6*pi/180);kF = kappaF/.1;kR = kappaR/.1;zetaF = sqrt(aF.^2 + kF.^2);zetaR = sqrt(aR.^2 + kR.^2);zeta = max(zetaF, zetaR);
+
+c = 'r';
+subplot(3, 1, 1)
+plot(s, UxDesired,c,'LineWidth',2)
+grid on; hold on;
+xlim(x0)
[…]

### 2019-02-13 journal/TCST2019/figures/plotTurn2G.m

@@ -1,0 +1,47 @@
+close all; clear all; clc;
+
+addpath('C:\Users\ddl\Documents\localTestData\searchData')
+load p1 %mu = .9
+U(:,1) = Ux;
+M(:,1) = currMu;
+ind1 = (FxFB ~= 0);
+load p24 %mu = .93; 
+U(:,2) = Ux;
+M(:,2) = .93;
+ind2 = (FxFB ~= 0);
+load p25 %mu = .95; 
+U(:,3) = Ux;
+M(:,3) = currMu;
+ind3 = (FxFB ~= 0);
+
+[C, I] = max(U');
+
+s = s + 700;
+
+x0 = [840 1236];
+subplot(2, 1, 1)
+grid on; hold on;
+
+plot(s, U(:,1),'k','LineWidth',2)
+plot(s, U(:,2),'b','LineWidth',2)
+plot(s, U(:,3),'r','LineWidth',2)
+legend('\mu = .9','\mu = .93','\mu = .95');
+
+plot(s(ind1), U(ind1,1),'k*')
+plot(s(ind2), U(ind2,2),'b*')
+plot(s(ind3), U(ind3,3),'r*')
+
+xlim(x0)
+ylabel('Actual Speed (m/s)','FontSize',14)
+
+for i = 1:len(s)
+    muGreedy(i) = M(i,I(i));
+end
+
+subplot(2, 1, 2)
+plot(s, muGreedy,'k','LineWidth',2)
+grid on; hold on;
+xlim(x0)
+
+xlabel('Distance Along Path (m)','FontSize',14)
+ylabel('\mu_{greedy}','FontSize',14)

### 2019-02-13 journal/TCST2019/figures/plotVtimeSeries.m

@@ -1,0 +1,127 @@
+clear all; close all; clc;
+
+% l(1) = load('C:\Users\ddl\Documents\localTestData\2014 5 ILC\fullautomu3_HL_2014-05-28_aa.mat');
+
+%PD Controller
+
+% l(5) = load('C:\Users\ddl\Documents\localTestData\2014 5 ILC\fullautomu3_4_HL_2014-05-28_aa.mat');
+% l(4) = load('C:\Users\ddl\Documents\localTestData\2014 5 ILC\fullautomu3_3_HL_2014-05-28_aa.mat');
+% l(3) = load('C:\Users\ddl\Documents\localTestData\2014 5 ILC\fullautomu3_2_HL_2014-05-28_aa.mat');
+% l(2) = load('C:\Users\ddl\Documents\localTestData\2014 5 ILC\fullautomu3_1_HL_2014-05-28_aa.mat');
+
+% l(1) = load('C:\Users\ddl\Documents\localTestData\2014 5 ILC\fullautomu6_HL_2014-05-28_aa.mat');
+% l(4) = load('C:\Users\ddl\Documents\localTestData\2014 5 ILC\fullautomu6_3_HL_2014-05-28_aa.mat');
+% l(3) = load('C:\Users\ddl\Documents\localTestData\2014 5 ILC\fullautomu6_2_HL_2014-05-28_aa.mat');
+% l(2) = load('C:\Users\ddl\Documents\localTestData\2014 5 ILC\fullautomu6_1_HL_2014-05-28_aa.mat');
+
+%l(1) = load('C:\Users\ddl\Documents\localTestData\2014 5 ILC\fullautomu7_HL_2014-05-28_aa.mat');
+%l(3) = load('C:\Users\ddl\Documents\localTestData\2014 5 ILC\fullautomu7_2_HL_2014-05-28_aa.mat');
+%l(2) = load('C:\Users\ddl\Documents\localTestData\2014 5 ILC\fullautomu7_1_HL_2014-05-28_aa.mat');
+
+%  l(1) = load('C:\Users\ddl\Documents\localTestData\2014 5 ILC\fullautomu8_wholetrack_HL_2014-05-28_aa.mat');
+%  l(4) = load('C:\Users\ddl\Documents\localTestData\2014 5 ILC\fullautomu8_wholetrack_3_HL_2014-05-28_aa.mat');
[…]

### 2019-09-03 conference/iv2019/manuscript.tex
modifications to tcst based on reviewer feedback

@@
 \title{\LARGE \bf
-A Hybrid Control Design for Autonomous Vehicles at Uncontrolled Intersections}
+A Hybrid Control Design for Autonomous Vehicles at Uncontrolled Crosswalks}
 
@@
 waiting) using Gaussian process dynamical models and trajectory matching from optical data.
-Several other techniques for pedestrian trajectory prediction have been proposed, including LQR \cite{Batkovic}, set-based reachability analysis \cite{Koschi2018}, and Markov processes \cite{Karasev2016}.
+Several other techniques for pedestrian trajectory prediction have been proposed, including LQR \cite{batkovic2018computationally}, set-based reachability analysis \cite{Koschi2018}, and Markov processes \cite{Karasev2016}.
 Another branch of literature focuses on modeling pedestrian crossing behavior in the form of video analysis.
@@
 	\State $a \gets k_s(v-v_\mathrm{speed limit})$ 
-	\If{$d > 0 \hspace{1mm} \mathbf{and} \hspace{1mm} \mathrm{inCrosswalk}(x_p)$}
+	\If{$d > 0 \hspace{1mm} \mathbf{and} \hspace{1mm} \mathrm{inCrosswalk}(x_p, \dot{x_p})$}
 		\If{$\mathrm{timeAdvantage}(x_p, \dot{x_p}, d, v) > \tau_\mathrm{max}$} 
@@
 \If {$state = YIELDING$}
-\If {$d > d_\mathrm{cmf} + t_{delay}v$}
+\If {$d > d_\mathrm{cmf} + t_\mathrm{delay}v$}
 	\State $a_{des} = 0$
@@
 
-\If {$\textbf{not} \hspace{1mm} \mathrm{inCrosswalk}(x_p)$} \State $state \leftarrow DRIVING$
+\If {$\textbf{not} \hspace{1mm} \mathrm{inCrosswalk}(x_p, \dot{x_p})$} \State $state \leftarrow DRIVING$
 
@@
 	\State $a = -\frac{v^2}{2d} + k_s(v - v_{des})$ 	
-	\If {$\textbf{not} \hspace{1mm} \mathrm{inCrosswalk}(x_p)$} \State $state \leftarrow DRIVING$
+	\If {$\textbf{not} \hspace{1mm} \mathrm{inCrosswalk}(x_p, \dot{x_p})$} \State $state \leftarrow DRIVING$
 
@@
 	\State $a = a_\mathrm{cmf}$ 	
-	\If {$\textbf{not} \hspace{1mm} \mathrm{inCrosswalk}(x_p) \vee d < 0$} \State $state \leftarrow DRIVING$
+	\If {$\textbf{not} \hspace{1mm} \mathrm{inCrosswalk}(x_p, \dot{x_p}) \vee d < 0$} \State $state \leftarrow DRIVING$
 
@@
 The pedestrian accepts a randomly sized time gap in the traffic stream to cross the road at constant speed $\dot{x}_p$.
+In reality, the pedestrian may accelerate or decelerate in the crosswalk, which would need to be addressed in the time advantage calculation.
 The size of the time gap is drawn from a normal distribution with mean $\mu_{gap}$ and standard deviation $\sigma_{gap}$.
@@
 Where $g_\mathrm{legality}$ and $g_\mathrm{safety}$ are terms encouraging the vehicle to stop when the pedestrian is in the crosswalk, $g_\mathrm{efficient}$ is a term encouraging high velocity when there is no pedestrian, and $g_\mathrm{smooth}$ encourages the vehicle to avoid abrupt changes in acceleration.
-For our implementation of \cite{Thornton2018}, we assumed $p_t$ was set to \textit{stopped}.
+For our implementation of \cite{Thornton2018}, we assumed $p_t$ was set to \textit{stopped}, indicating a pedestrian stopped at the side of the crosswalk making eye contact with the vehicle.
 
@@
 The minimum distance between the vehicle and the pedestrian for each trial is shown in (a) and (b) of Figures ~\ref{fig:velPlot1} and \ref{fig:velPlot2}.
-The vehicle is able to maintain a safe distance of at least four meters from the pedestrian for the case where the vehicle is Lane B.
+The vehicle is able to maintain a safe distance of at least four meters from the pedestrian for the case where the vehicle is in Lane B.
 This is because the algorithm has enough time to react to a pedestrian crossing no matter what size gap is accepted.
@@
 For the hybrid controller, this is because the need to stop at $d=0$ is explicitly encoded in the design.
-For the POMDP, this is due to the reward function terms $g_\mathrm{legality}$ and $g_\mathrm{safety}$
+For the POMDP, this is due to the reward function terms $g_\mathrm{legality}$ and $g_\mathrm{safety}$.
 
@@
 \centering
-\includegraphics[width=3.0in]{figures/G80.jpg}
+\includegraphics[width=2.5in]{figures/G80.jpg}
 \caption{Hyundai Genesis G80 experimental testbed.}
@@
 \centering
-\includegraphics[width=2.0in]{figures/crosswalkPic.png}
+\includegraphics[width=1.5in]{figures/crosswalkPic.png}
 \caption{Two lane crosswalk used for experimental testing.}

### 2019-09-03 journal/TCST2019/manuscript.tex
modifications to tcst based on reviewer feedback

@@
 \begin{abstract}
-Autonomous vehicles will generate tremendous data from the variety of sensors they employ to track the surrounding environment.
+Autonomous vehicles will generate data from the variety of sensors they employ to track the surrounding environment.
 This data is inherently valuable, as it gives algorithm designers the potential to leverage prior experience in order to improve driving performance over time.
@@
 Two complementary algorithms are demonstrated in this paper.
-The first type uses iterative learning control (ILC) to simultaneously improve lateral and longitudinal tracking of the desired racing trajectory from lap-to-lap, while the second type of algorithm is focused on altering the trajectory itself using a search method.
+The first algorithm uses iterative learning control (ILC) to simultaneously improve lateral and longitudinal tracking of the desired racing trajectory over multiple laps, while the second algorithm is focused on altering the trajectory itself using a search method.
 When driven experimentally at the limits of handling, the result is a reduction in lap time of nearly 1.4 seconds, a major improvement.
@@
 \subsection{Problem Description and Proposed Solution}
-Several methods exist in the literature to generate autonomous racing trajectories (~\cite{casanova},\cite{kelly},\cite{theodosis},\cite{kapaniadscc}), which typically comprise of a longitudinal speed profile $U_x(s)$ and a desired curvature profile $\kappa(s)$ as a function of distance along the track.
+Several methods exist in the literature to generate autonomous racing trajectories 
+(\cite{casanova},\cite{kelly},\cite{theodosis},\cite{kapaniadscc}), which typically comprise of a longitudinal speed profile $U_x(s)$ and a desired curvature profile $\kappa(s)$ as a function of distance along the track.
 These trajectories can then be tracked using one of several path following controllers, such as (\cite{mickcop},\cite{fricStudy},\cite{mickgeneral},\cite{shladover},\cite{nagai}).
@@
 Using the trajectory generated by \cite{kapaniadscc} and the trajectory following controller presented by \cite{kapania}, the
-authors generated a test run of an autonomous race vehicle (see Fig.\ref{fig:carpic}(a)) at the limits of handling on the Thunderhill Raceway course in Willows, CA (Fig.~\ref{fig:carpic}(b)).
-The results are shown in Fig.
-\ref{fig:expErrors}.
+authors generated a test run of an autonomous race vehicle (see Fig.\ref{fig:carpic}) at the limits of handling on the Thunderhill Raceway course in Willows, CA. %(Fig.~\ref{fig:carpic}(b)). 
+The results are shown in Fig. \ref{fig:expErrors}.
 
@@
 \centering
-\includegraphics[width=3.5in]{figures/pic1.png}
-\caption{(a) ``Shelley" - Stanford's autonomous self-driving race vehicle. (b) Top-down view of Thunderhill Raceway course. }
+\includegraphics[width=2.5in]{figures/pic1.png}
+\caption{``Shelley" - Stanford's autonomous self-driving race vehicle.}
 \label{fig:carpic}
@@
 \centering
-\includegraphics[width=3.3in]{figures/expErrors.eps}
+\includegraphics[width=2.8in]{figures/expErrors.eps}
 \caption{Controller tracking performance and tire slip norm on a test run at the limits of handling ($\mu =$ 0.94).(a) Desired
@@
 
-Where $\sigma$ and $\alpha$ are the longitudinal and lateral tire slip for a given tire, and $\sigma_p$ and $\alpha_p$ are empirically
+Where $\sigma$ and $\alpha$ are the maximum longitudinal and lateral tire slips over all four tires, and $\sigma_p$ and $\alpha_p$ are empirically
 determined \textit{peak slip} values resulting in maximum longitudinal and lateral tire force generation.
@@
 
-Iterative learning control (ILC) algorithms can help an autonomous vehicle follow a desired trajectory more precisely over 
-several laps of driving. However, ILC algorithms do not alter the desired trajectory itself, only the input signals that attempt
-to track the trajectory.
-This will not be sufficient at the limits of handling.
-Consider Fig.~{\ref{fig:expErrors}, where the desired speed profile was generated assuming a tire-road friction value of $\mu = 0.94$.
-While this is a reasonable assumption overall, there are several parts on the track where the vehicle exceeds the available friction and begins to understeer.
-Region \circled{2} is one example.
-The tire slip norm $\zeta$ climbs above one, and as a result, the vehicle begins
-to slide off the track, resulting in the large negative tracking error spike in Fig.~{\ref{fig:expErrors}(b).
-The vehicle's stability algorithms must kick in and slow the vehicle down, which results in the speed tracking error seen in Fig.~{\ref{fig:expErrors}(a).
-In this situation, ILC will be unable to achieve better speed tracking and path tracking performance.
-The tires are saturated and the car is slowly beginning to careen off the track.
-Simply steering more
-on the next lap will not achieve better tracking performance due to the saturation of the steering actuator.
-In order to recover, the vehicle must deviate from the planned trajectory, either by slowing down or by taking a wider radius turn.
+Iterative learning control (ILC) algorithms can help an autonomous vehicle follow a desired trajectory more precisely over several laps of driving.
+However, ILC algorithms do not alter the desired trajectory itself, only the input signals that attempt to track the trajectory.
+In cases where the trajectory results in oversteer or understeer, a trajectory modification algorithm is necessary since the vehicle may already be saturating the tires, and applying additional actuator effort will not result in better tracking.
 
@@
 For example, Theodosis and Gerdes presented an algorithm to gradually widen the radius of a turn in response to a detected understeer \cite{paulthesis}.
-The algorithm was validated
-experimentally, but assumed sufficient availability of road width. Klomp and Gordon also developed a strategy for recovering from
-vehicle understeer by solving for an optimal emergency braking profile to minimize deviation from the desired path \cite{klomp}.
-On the other hand, Brown et al.~\cite{brown2017safe} presented a model predictive control (MPC) approach that generally aimed to follow a planned vehicle trajectory at the limits.
+Klomp and Gordon also developed a strategy for recovering from
+vehicle understeer by solving for an optimal emergency braking profile to minimize deviation from the desired path \cite{klomp} Brown et al.~\cite{brown2017safe} presented a model predictive control (MPC) approach that generally aimed to follow a planned vehicle trajectory at the limits.
 However, if the vehicle was at risk of understeering or oversteering, the MPC algorithm would deviate laterally from the planned trajectory in order to maintain stability of the vehicle without driving off the road.
-
 Another approach to trajectory modification is Learning Model Predictive Control (LMPC), developed by Rosolia et al.
@@
 In LMPC, data from every lap of racing is used to build invariant sets and approximate the value function for the generally non-convex lap time minimization problem.
-In practice, this enables the controller to explore different areas of the state space on each successive lap, modifying the desired trajectory on each lap in order to continuously improve the lap time.
 
@@
 
-Due to the repetitive nature of automotive racing, iterative learning control techniques are a promising method to gradually
-eliminate \textit{trajectory tracking} errors .
-For this application of ILC, the repetitive trials are laps of racing, and the reference trajectory is the optimal speed profile and curvature developed in \cite{kapaniadscc}.
-This paper will present adaptations of two established ILC methods
-(proportional-derivative and quadratically optimal) for use in the Audi TTS racing system. Section \ref{sec:ch4dsm} presents
-both coupled and decoupled models of the vehicle lateral and longitudinal dynamics. These models are converted from state-space representations to \textit{lifted domain}
-representations required for iterative learning control, and the proportional-derivative and quadratically optimal ILC algorithms are presented in \S \ref{sec:pdcontroller} and \S \ref{sec:qopt}.
-Section \ref{sec:simres} presents simulated results of the ILC algorithms for a sample vehicle trajectory, and finally,
-experimental results showing a gradual reduction of trajectory-following errors is presented in \S \ref{sec:ch4ExpRes}.
+Due to the repetitive nature of automotive racing, iterative learning control techniques are a promising method to gradually eliminate \textit{trajectory tracking} errors.
+This paper will first present adaptations of two established ILC methods
+(proportional-derivative and quadratically optimal) for use in the Audi TTS racing system.
+The ILC algorithms will be designed for combined lateral and longitudinal control, as an extension of the lateral-only ILC first presented in \cite{kapaniaacc}
 
-For situations where the vehicle underperforms due to excessive oversteer or understeer, an A* search algorithm is presented
+Second, this paper will address situations where the vehicle underperforms due to excessive oversteer or understeer. An A* search algorithm is presented
 that iteratively modifies portions of the \textit{velocity} profile to be more conservative if a stability violation is encountered on a prior lap.
-This is accomplished by generalizing the speed profile such that each part of the track can be driven with a learned  value of friction $\mu$ and therefore a different maximum
-acceleration.
 The algorithm also modifies the velocity profile to be more aggressive if the tires are not being actuated at the limits.
@@
 The ILC algorithms we consider require a set of closed-loop system dynamics that are (a) stable to any disturbance input, and (b) expressible as an \textit{affine} discrete dynamical system.
-In our case, we have two subsystems: the steering controller and the longitudinal speed control. Stability of the steering controller
+In our case, we have two subsystems: the steering controller and the longitudinal speed control.
+Stability of the steering controller
 under lanekeeping feedback was shown in the linear case by \cite{rossetter2002} and in the saturated case by \cite{talvala}.
@@
 The more difficult task is expressing the dynamics of the two subsystems using an affine model, given the tendency for the vehicle tires to saturate at the limits.
-This paper will model the dynamics with a bicycle model with single longitudinal input $F_x$ and front steer input $\delta$, shown in Fig.\ref{fig:bikemodel}.
+This paper will model the dynamics with a bicycle model with single longitudinal input $F_x$ and front steer input $\delta$, shown in Fig.~\ref{fig:bikemodel}.
 
@@
 
-Nonlinear, coupled equations of motion are provided in (\ref{eq:fullNL})-(\ref{eq:fullNL2}):
+% Nonlinear, coupled equations of motion are provided in (\ref{eq:fullNL})-(\ref{eq:fullNL2}):
 
-\begin{align}
-\label{eq:fullNL}
-	\dfrac{de}{dt} &= \left(v+U^\mathrm{des}_x(s)\right)(\beta + \Delta\Psi) \\
-	\dfrac{dv}{dt} &= \left(v+U^\mathrm{des}_x(s)\right)\beta r + \frac{F_x}{m} - \frac{F_\mathrm{yf}(\alpha_\mathrm{f},F_x)\delta}{m}\\
-	\dfrac{d\beta}{dt} &= \frac{F_\mathrm{yf}(\alpha_\mathrm{f}, F_x) + F_\mathrm{yr}(\alpha_\mathrm{r},F_x)}{m\left(v + U^\mathrm{des}_x(s)\right)}-r \\
-	\dfrac{dr}{dt}     &= \frac{aF_\mathrm{yf}(\alpha_\mathrm{f}, F_x) - bF_\mathrm{yr}(\alpha_\mathrm{r}, F_x)}{I_z} \\
-	\dfrac{d\Delta\Psi}{dt} &= r \label{eq:fullNL2}
-\end{align}
-The system dynamics presented in (\ref{eq:fullNL}) have five states, yaw rate $r$, vehicle sideslip $\beta$, lateral path tracking error $e$, heading error $\Delta \Psi$, and speed tracking error $v$, defined by:
+% \begin{align}
+% \label{eq:fullNL}
+% 	\dfrac{de}{dt} &= \left(v+U^\mathrm{des}_x(s)\right)(\beta + \Delta\Psi) \\
+% 	\dfrac{dv}{dt} &= \left(v+U^\mathrm{des}_x(s)\right)\beta r + \frac{F_x}{m} - \frac{F_\mathrm{yf}(\alpha_\mathrm{f},F_x)\delta}{m}\\
+% 	\dfrac{d\beta}{dt} &= \frac{F_\mathrm{yf}(\alpha_\mathrm{f}, F_x) + F_\mathrm{yr}(\alpha_\mathrm{r},F_x)}{m\left(v + U^\mathrm{des}_x(s)\right)}-r \\
+% 	\dfrac{dr}{dt}     &= \frac{aF_\mathrm{yf}(\alpha_\mathrm{f}, F_x) - bF_\mathrm{yr}(\alpha_\mathrm{r}, F_x)}{I_z} \\
+% 	\dfrac{d\Delta\Psi}{dt} &= r - \kappa(s)(v+U^\mathrm{des}_x(s)) \label{eq:fullNL2}
+% \end{align}
+% The system dynamics presented in (\ref{eq:fullNL}) have five states, yaw rate $r$, vehicle sideslip $\beta$, lateral path tracking error $e$, heading error $\Delta \Psi$, and speed tracking error $v$, defined by:
 
-\begin{equation}
-v = U_x - U^\mathrm{des}_x
-\end{equation}
+% \begin{equation}
+% v = U_x - U^\mathrm{des}_x
+% \end{equation}
 
-The potential for coupling between the subsystems is apparent from  (\ref{eq:fullNL})-(\ref{eq:fullNL2}), not only directly from the state equations but also due to the complex nature of tire force generation at the handling limits.
-As shown in Fig.~\ref{fig:coupledTires}, as longitudinal
-force $F_x$ is distributed across the tires, the available lateral force decreases. At the limits of handling, this
-\textit{derating} of the lateral force may become significant. As a result, we model lateral tire force generation $F_y$ as a function
-of both lateral tire slip $\alpha$ and the applied longitudinal force $F_x$, using a modified form of the Fiala tire model \cite{rami}.
-The front and rear tire slip are themselves functions of the vehicle state and are given by:
-
-\begin{subequations}
-\begin{align}
-	\alpha_\mathrm{f} &= \beta + \frac{ar}{U_x} - \delta\\
-	\alpha_\mathrm{r} &= \beta - \frac{br}{U_x}
-\end{align}
-\end{subequations}
-
+The potential for coupling between the subsystems occurs from both the state equations and also the complex nature of tire force generation at the handling limits.
+As shown in Fig.~\ref{fig:coupledTires}, as longitudinal force $F_x$ is distributed across the tires, the available lateral force decreases.
+At the limits of handling, this \textit{derating} of the lateral force may become significant.
+As a result, we model lateral tire force generation $F_y$ as a function of both lateral tire slip $\alpha$ and the applied longitudinal force $F_x$, using a modified form of the Fiala tire model \cite{rami}.
 \begin{figure}[tb]
@@
 
-The next step is to break up the control inputs into the closed-loop feedback term and the \textit{learned} component that is
-modified on every lap by the ILC algorithm:
+The control inputs $F_x$ and $\delta$ are broken up into the closed-loop feedback term and the \textit{learned} component that is modified on every lap by the ILC algorithm:
 \begin{align}
@@
 \end{align}
-where $K_x$ is the proportional speed tracking gain and $k_p$ and $x_\mathrm{LA}$ are the lookahead gains presented in \cite{kapania}.
+where $K_x$ is the proportional speed tracking gain and $k_p$ and $x_\mathrm{LA}$ are the lookahead gains presented in \cite{kapania}. The term $v$ denotes the speed tracking error.
 
-The closed loop system dynamics are now given by: 
+The closed loop system dynamics are given by incorporating (\ref{eqn:splitTerms}) into equations of motion for the bicycle model: 
 \begin{align}
@@
 	\frac{dr}{dt}     &= \frac{aF_\mathrm{yf}(\alpha_\mathrm{f}, F_x) - bF_\mathrm{yr}(\alpha_\mathrm{r}, F_x)}{I_z} \\
-	\frac{d\Delta\Psi}{dt} &= r \label{eq:fullNLCL2}	
+	\frac{d\Delta\Psi}{dt} &= r - \kappa(s)(v+U^\mathrm{des}_x(s))\label{eq:fullNLCL2}	
 \end{align}
@@
 Since we run the iterative learning control algorithm after seeing a trial of data, we can approximate
-the dynamics in (\ref{eq:fullNLCL})-(\ref{eq:fullNLCL2}) by linearizing about the observed states and inputs $x_o, u_o$ from the first lap. The affine model is therefore given by: 
+the dynamics in (\ref{eq:fullNLCL})-(\ref{eq:fullNLCL2}) by linearizing about the observed states and inputs $x_o, u_o$ from the first lap.
+These state estimates can be obtained from a differential GPS, simulation, or a state observer.
+The affine model is therefore given by:
  \begin{align}
@@
  \begin{bmatrix} \dot{e} \\ \dot{\beta} \\ \dot{r} \\ \dot{\Delta\Psi} \\ \dot{v}\end{bmatrix} &= \begin{bmatrix} A(t) \end{bmatrix} \begin{bmatrix} e \\ \beta \\ r \\ \Delta\Psi \\ v \end{bmatrix} + \begin{bmatrix} B(t) \end{bmatrix} \begin{bmatrix} \delta_L \\ F^L_x \end{bmatrix} + d(t) \\
- \begin{bmatrix} \dot{e} \\ \dot{v} \end{bmatrix}  &= \begin{bmatrix} 1 & 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 & 1 \end{bmatrix} x
+ \begin{bmatrix} e \\ v \end{bmatrix}  &= \begin{bmatrix} 1 & 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 & 1 \end{bmatrix} x
  \end{align}
@@
 
- While this multiple-input, multiple-output (MIMO) model captures the coupled behavior of the longitudinal and lateral inputs, it is tedious to compute,
- either numerically or analytically.
-If we make the assumption that the longitudinal and lateral inputs are decoupled and make small angle approximations, we obtain the following linear state equations for the lateral dynamics:
-\begin{subequations}
-\label{eq:bmC4}
-\begin{align}
-	\dot{\beta} &= \frac{F_\mathrm{yf}+F_\mathrm{yr}}{mU_x} - r \qquad \dot{r} = \frac{aF_\mathrm{yf} - bF_\mathrm{yr}}{I_z} \label{bm1} \\
-	\dot{e} &= U_x (\beta + \Delta\Psi) \qquad \Delta\dot{\Psi} = r - U_x\kappa \label{eq:bm2} 
-\end{align}
-\end{subequations} 
-and the following first order equation for the longitudinal dynamics, assuming a simple point mass model with proportional
-speed tracking feedback:
-\begin{equation}
-\dot{v} = \frac{-K_xv + F^L_x}{m}
-\end{equation}
-The resulting state matrices $A(t)$ and $B(t)$ for (\ref{eq:ssMAT}) are then simply block diagonal matrices consisting of the
-linearized lateral dynamics and the first order longitudinal dynamics:
+ This multiple-input, multiple-output (MIMO) model captures the coupled behavior of the longitudinal and lateral inputs.
+If we make the assumption that the longitudinal and lateral inputs are decoupled and make small angle approximations, we may obtain the following resulting analytical state matrices $A(t)$ and $B(t)$ for (\ref{eq:ssMAT}):
 
@@
   \frac{-k_\mathrm{p}\tilde{C}_\mathrm{f}(t)}{mU_x(t)}  & \frac{-k_\mathrm{p}x_\mathrm{LA}\tilde{C}_\mathrm{f}(t)}{mU_x(t)}  & \frac{b\tilde{C}_\mathrm{r}(t)-a\tilde{C}_\mathrm{f}(t)}{mU_x(t)^2}-1 & \frac{-\tilde{C}_\mathrm{f}(t) - \tilde{C}_\mathrm{r}(t)}{mU_x(t)} & 0 \\
-  0 & 0 & 0 & 0 & -K_x
+  0 & 0 & 0 & 0 & -\frac{K_x}{m}
   \end{smallmatrix}\right] \\
-&B(t) =\begin{bmatrix} 0 & 0 \\	 	0 & 0 		\\ 		\frac{a \tilde{C}_\mathrm{f}(t)}{I_\mathrm{z}} & 0 		\\ 		\frac{\tilde{C}_\mathrm{f}(t)}{mU_x(t)} & 0 		\\ 0 & 1 \end{bmatrix} 
+&B(t) =\begin{bmatrix} 0 & 0 \\	 	0 & 0 		\\ 		\frac{a \tilde{C}_\mathrm{f}(t)}{I_\mathrm{z}} & 0 		\\ 		\frac{\tilde{C}_\mathrm{f}(t)}{mU_x(t)} & 0 		\\ 0 & \frac{1}{m} \end{bmatrix} 
 \end{align}
@@
 
+While beyond the scope of this paper, it should also be noted that advances in ILC research have focused specifically on ILC for nonlinear \cite{xunonlinear} and linear parameter-varying (LPV) systems \cite{de2017iterative}.
+
 \subsection{Lifted Domain Representation and ILC \newline Problem Statement}
@@
 Whether the coupled (\ref{eq:C4nl}) or decoupled (\ref{eqn:C4dyn}) dynamics are assumed, the
-final modeling step is to apply standard discretization techniques to obtain dynamics in the following form:
+final modeling step is to apply the matrix exponential discretization technique to obtain dynamics in the following form:
 \begin{align}
@@
 Where $C_y$ is the row of $C$ in (\ref{C4ad}) corresponding to the desired output and $B_u$ the column of $B$ in (\ref{C4ad})
-corresponding to the desired input. Note that
-for the case of uncoupled lateral and longitudinal dynamics, the off-diagonal sub-matrices of $P$ are $[0]$ since we have
-two SISO systems. The term $\mathbf{w}$ in (\ref{eq:liftedDomain}) is the unknown disturbance. Iterative
+corresponding to the desired input. The term $\mathbf{w}$ in (\ref{eq:liftedDomain}) is the unknown disturbance. Iterative
 learning control relies on the assumption that this disturbance is the underlying cause of the observed errors $\mathbf{e}_j$, and that the disturbance, while unknown, is constant from lap to lap.
@@
 where $k_{p\delta}$ and $k_{pF}$ are proportional gains and $k_{d\delta}$ and $k_{dF}$ are derivative gains.
-In the lifted domain representation from (\ref{eqn:ctrlLaw}), the resulting learning matrix $L$ is given by
-\begin{equation}
-	L = \left[\begin{smallmatrix} 		-(k_{p\delta}+k_{d\delta}) &          &  0                            & 0 & \hdots & 0 \\ 
-									   k_{d\delta}                     &  \ddots  &                               & \vdots & \ddots & \vdots\\ 
-									   0                       &   k_{d\delta}    &    -(k_{p\delta}+k_{d\delta}) & 0 & \hdots & 0 \\
-						               0 & \hdots & 0 & -(k_{pF}+k_{dF}) &          &  0\\
-									   \vdots & \ddots & \vdots & k_{dF} &  \ddots  &                               &\\
-									   0 & \hdots & 0  & 0 & k_{dF} & -(k_{pF} + k_{dF}) \\			
-			\end{smallmatrix}\right]
-\end{equation}
 
@@
 assists with monotonic stability by removing oscillations in the control input generated when trying to remove small reference tracking errors after several iterations.
-Since the filtering occurs when generating a control signal for the next lap, the filter $Q$ can be zero-phase.
 The plot shown in Fig.~\ref{fig:stabPlot} is for the steering ILC design only, but the same analysis is possible for the longitudinal
@@
 \centering
-\includegraphics[width=2.45in]{figures/monotonicStability.png}
+\includegraphics[width=2 in]{figures/monotonicStability.png}
 \caption[Values of convergence bound $\gamma$ vs.
@@
 To test the feasibility of the PD and Q-ILC learning algorithms, the vehicle tracking performance over multiple laps is
-simulated using the path curvature and speed profiles shown in Fig.~\ref{fig:simSetup}. To test the performance
-of the controller at varying accelerations, four speed profiles are tested. Each profile is
-generated with a different level of peak combined longitudinal/lateral acceleration, ranging from 5 $\mathrm{m/s^2}$ (below the limits)
-up to 9.5 $\mathrm{m/s^2}$ (close to exceeding the limits). For accurate results, simulations were conducted using numerical integration with
- fully nonlinear equations of motion (\ref{eq:fullNL})-(\ref{eq:fullNL2}) and coupled lateral/longitudinal dynamics \cite{rami}.
-
+simulated using the path curvature and speed profiles shown in Fig.~\ref{fig:simSetup}.
 \begin{figure}[bh]
 \centering
-\includegraphics[width= 3.35in]{figures/simSetup.eps}
+\includegraphics[width= 2.7in]{figures/simSetup.eps}
 \caption[Curvature profile used for ILC simulation.]{ (a) Curvature profile used for ILC simulation. (b) Velocity profiles generated for four different levels of longitudinal/lateral acceleration. }
@@
 \centering
-\includegraphics[width=3.35in]{figures/simRes1.eps}
+\includegraphics[width=3.5in]{figures/simRes1.eps}
 \caption[Simulated results for root-mean-square path tracking error at several values of vehicle acceleration]{Simulated results for root-mean-square path tracking error at several values of vehicle acceleration, with $T = R = I$ and $S = \mathrm{100} I$.
@@
 Fig.~\ref{fig:simRes1} shows that both quadratically optimal ILC algorithms exponentially reduce the lateral
-path tracking error as the number of learning iterations is increased. Overall, the RMS lateral tracking performance is
-better at lower vehicle accelerations.
-This is unsurprising for two reasons.
-First, lateral path deviation in general increases for the lookahead steering feedback at higher accelerations.
-Second, our estimate of the vehicle dynamics contained in $P$
-is based on linearization, and the vehicle dynamics at lower accelerations are mostly linear.
+path tracking error as the number of learning iterations is increased.
+Fig.~\ref{fig:simRes2} shows the same results as Fig.~\ref{fig:simRes1}, but for the speed tracking performance.
+The overall trends are very similar.
+In both plots, there is very little difference between the coupled MIMO formulation and decoupled SISO formulation at low accelerations.
+At higher accelerations, there are small differences, but the overall RMS errors are still quite similar.
 
-Fig.~\ref{fig:simRes2} shows the same results as Fig.~\ref{fig:simRes1}, but for the speed tracking performance. 
-The overall trends are very similar. In both plots, there is very little difference between the coupled MIMO formulation and decoupled SISO 
-formulation at low accelerations. This is expected, as the longitudinal and lateral dynamics are independent when the vehicle tires are not saturated.
-At higher accelerations, there are small differences, but the overall RMS errors are still quite similar. A reason for this is the
-nature of the speed profiles in Fig.~\ref{fig:simSetup}. The vehicle spends the majority of time either fully braking/accelerating or
-turning at a constant velocity. There are only a few small transient regions where the vehicle needs significant amounts of both 
-lateral and longitudinal acceleration. As a result, the need to account for the coupled dynamics may not be important in practice, especially
-given the larger computation time needed when $P$ is dense and not block-diagonal.
+%  A reason for this is the
+% nature of the speed profiles in Fig.~\ref{fig:simSetup}. The vehicle spends the majority of time either fully braking/accelerating or
+% turning at a constant velocity. There are only a few small transient regions where the vehicle needs significant amounts of both 
+% lateral and longitudinal acceleration. As a result, the need to account for the coupled dynamics may not be important in practice, especially
+% given the larger computation time needed when $P$ is dense and not block-diagonal.
 
@@
 This is unsurprising given that the linearized plant dynamics $P$ are not explicitly accounted for in the selection of the PD gains.
-While the point mass model for the longitudinal dynamics is a relatively simple first order model, the lateral dynamics 
-are fourth order and highly speed dependent. A simple PD approach for learning control is likely insufficient at the limits of handling.
+
+% While the point mass model for the longitudinal dynamics is a relatively simple first order model, the lateral dynamics 
+% are fourth order and highly speed dependent. A simple PD approach for learning control is likely insufficient at the limits of handling.
 
@@
 % \end{figure}
-
-To save computation time, the learned controller inputs $\delta^L$ and $F^L_x$  are calculated at a 10 Hz update rate at the end of every race lap and stored as lookup tables in the controller. 
-Since the the real-time control occurs at 200 Hz, at every time step the controller interpolates
-the lookup table and applies the correct force and steer angle correction.
 
@@
 the steering corrections \textit{lead} the observed error).
-This is because the learning
-algorithm has knowledge of the system model and knows that a steering correction must be applied a few meters early to cancel a path deviation
-further down the road.
+This is because the ILC learns to anticipate repetitive disturbances and acts proactively.
 
@@
 \end{figure}
+
 Fig.~\ref{fig:expRes2} shows the iterative learning signals for the longitudinal speed control at a slightly
@@
 two to three iterations, significant lags in the speed tracking are attenuated from $s = 800-900$ meters and $s = 1200-1300$ meters.
-Additionally, the controller also acts to slow the car down when $v > 0$ and the vehicle exceeds the speed profile.
-This is also desirable
-from a racing perspective as it it prevents the vehicle from exceeding the friction limit. Oscillations in speed tracking
-performance are visible in Fig.~\ref{fig:expRes1}(a) from 900 - 1100 meters and 1300 - 1400 meters. These are generally undesirable, and further tuning of the filter matrix $Q$ is possible
-to remove rapid changes in the learned force input.
+
+% Additionally, the controller also acts to slow the car down when $v > 0$ and the vehicle exceeds the speed profile. This is also desirable
+% from a racing perspective as it it prevents the vehicle from exceeding the friction limit. Oscillations in speed tracking
+% performance are visible in Fig.~\ref{fig:expRes1}(a) from 900 - 1100 meters and 1300 - 1400 meters. These are generally undesirable, and further tuning of the filter matrix $Q$ is possible
+% to remove rapid changes in the learned force input.
 
@@
 because the desired longitudinal action is to fully apply the throttle and go as fast as physically possible. For convenience, the ILC is programmed to saturate the longitudinal
-learning signal to 8000 Newtons, although a more elegant solution is to switch the ILC controller off on straight portions of the track.
+learning signal to 8000 Newtons. This is unlikely to affect convergence as the throttle is likely to already be fully applied.
+
+% , although a more elegant solution is to switch the ILC controller off on straight portions of the track.
 
@@
 lateral tracking error is around 3 cm, on the order of the expected RMS error from the GPS position sensor.
+
 On some tests, the RMS tracking error 
@@
 before decreasing further in Lap 3. While not predicted in simulation, this behavior likely occurs because the repeating disturbance
-from lap-to-lap is not exactly constant, especially as the vehicle approaches the handling limits. More refined tuning of the gain matrices may be able to prevent 
-this RMS error increase, or the ILC algorithm can be stopped after several iterations once the tracking performance is acceptable.
+from lap-to-lap is not exactly constant, especially as the vehicle approaches the handling limits.
 
-Experimental results in this section were only given for the quadratically optimal controller with decoupled (SISO) dynamics. 
-The PD iterative learning controller was not tested due to the relatively worse simulation performance, and the quadratically optimal controller with coupled dynamics provided no clear benefit in simulation but a much longer computation time.
+% More refined tuning of the gain matrices may be able to prevent 
+% this RMS error increase, or the ILC algorithm can be stopped after several iterations once the tracking performance is acceptable.
+
+% Experimental results in this section were only given for the quadratically optimal controller with decoupled (SISO) dynamics. 
+% The PD iterative learning controller was not tested due to the relatively worse simulation performance, and the quadratically optimal controller with coupled dynamics provided no clear benefit in simulation but a much longer computation time.
 
@@
 speed profile is generated by assuming a global value of tire friction $\mu$, which is directly related to the acceleration norm of
-the desired speed profile ($\sqrt{a_x^2 + a_y^2}$) by a factor of \mbox{$g =$ 9.81 $\mathrm{m/s^2}$}. Fig.~\ref{fig:t2}(a) confirms that higher levels
-of $\mu$ result in strictly faster speed profiles for the same turn.
+the desired speed profile $\left(\sqrt{a_x^2 + a_y^2}\right)$ by a factor of \mbox{$g =$ 9.81 $\mathrm{m/s^2}$}.
+
+% Fig.~\ref{fig:t2}(a) confirms that higher levels
+% of $\mu$ result in strictly faster speed profiles for the same turn.
 
@@
 \centering
-\includegraphics[width=2.8in]{figures/t2res.eps}
+\includegraphics[width=2.3in]{figures/t2res.eps}
 \caption[Desired speed for varying levels of $\mu$ - region 2]{(a) Desired speed for varying levels of $\mu$ (b) Actual speed for varying levels of $\mu$. Asterisks correspond to 
@@
 is occurring primarily at the front tires, causing an understeer that can cause the car to skid off the track. Completing the lap therefore requires a stabilizing
-action from the stability controller to slow the car down and regain control of the vehicle. As a result,
-when the vehicle accelerates at the end of the turn, the actual vehicle speed at $\mu = 0.95$ is significantly slower than the case where $\mu = 0.9$!
+action from the stability controller to slow the car down and regain control of the vehicle.
+
+As a result,
+when the vehicle accelerates at the end of the turn, the actual vehicle speed at $\mu = 0.95$ is significantly slower than the case where $\mu = 0.9$.
 A final ``just right" value of $\mu = 0.93$ was also tested experimentally for this turn, and while the car does slide a bit at this level of driving (peak $\zeta = 1.3$),
@@
 
- \begin{figure}[tb]
-\centering
-\includegraphics[width=2.8in]{figures/t3res.eps}
-\caption[Desired speed for varying levels of $\mu$ - region 3]{(a) Desired speed for varying levels of $\mu$. (b) Actual speed for varying levels of $\mu$. Asterisks correspond to 
-regions where $\zeta > 1$. (c) Tire slip norm for varying levels of $\mu$. }
-\label{fig:t3}
-\end{figure}
+%  \begin{figure}[tb]
+% \centering
+% \includegraphics[width=2.3in]{figures/t3res.eps}
+% \caption[Desired speed for varying levels of $\mu$ - region 3]{(a) Desired speed for varying levels of $\mu$. (b) Actual speed for varying levels of $\mu$. Asterisks correspond to 
+% regions where $\zeta > 1$. (c) Tire slip norm for varying levels of $\mu$. }
+% \label{fig:t3}
+% \end{figure}
 
-Unfortunately, the best value of $\mu$ is not constant throughout the track. Fig.~\ref{fig:t3} shows the same data plotted for region \circled{3} of the 
-Thunderhill Raceway. In this case, even with an atypically high value of 
-$\mu = 0.97$, the slip norm of the vehicle tires is relatively low, and the value of $\mu$ that optimizes the completion time for this turn could 
-be even larger.
+% Unfortunately, the best value of $\mu$ is not constant throughout the track. Fig.~\ref{fig:t3} 
+% shows the same data plotted for region \circled{3} of the 
+% Thunderhill Raceway. In this case, even with an atypically high value of 
+% $\mu = 0.97$, the slip norm of the vehicle tires is relatively low, and the value of $\mu$ that optimizes the completion time for this turn could 
+% be even larger.
 
@@
 
-The simplest approach to finding $\mu(s)$ is a greedy algorithm where a set of experimental data is collected for a variety of different
-speed profiles, each corresponding to a different value of $\mu$ and therefore a different acceleration level. The greedy approach is then to discretize
-the track into a number of small sections and pick the value of $\mu$ for each section that corresponds to the highest observed experimental velocity.
+The simplest approach to finding $\mu(s)$ is a greedy algorithm where a set of experimental data is collected for a variety of different speed profiles, each corresponding to a different value of $\mu$ and therefore a different acceleration level.
+The greedy approach is then to discretize the track into a number of small sections and pick the value of $\mu$ for each section that corresponds to the highest observed experimental velocity.
 The final desired velocity profile is then generated using the numerical integration approach presented in \cite{subosits2015autonomous}. This method does not require a single 
@@
 
-A plot showing the results of applying the greedy algorithm for section \circled{2} is shown in Fig.~\ref{fig:t2g}. The flaw in simply
-selecting $\mu$ based on the highest speeds is apparent at $s = 1050 \mathrm{m}$. The greedy algorithm suggests the vehicle mostly drive at $\mu$ = 0.95,
-but then switch to driving at $\mu = 0.93$ as soon as the vehicle begins to slide. Switching to a less aggressive velocity profile at this
-point is impossible to achieve in practice, because the vehicle is already fully sliding and has no control until the vehicle slows down.
-As a result, the greedy algorithm fails to capture the hidden cost of a large understeer, which results in a period of time where the speed must
-inevitably drop.
+A plot showing the results of applying the greedy algorithm for section \circled{2} is shown in Fig.~\ref{fig:t2g}.
+The flaw in simply selecting $\mu$ based on the highest speeds is apparent at $s = 1050 \mathrm{m}$.
+The greedy algorithm suggests the vehicle mostly drive at $\mu$ = 0.95,
+but then switch to driving at $\mu = 0.93$ as soon as the vehicle begins to slide.
+Switching to a less aggressive velocity profile at this point is impossible to achieve in practice, because the vehicle is already fully sliding and has no control until the vehicle slows down.
+
+ % As a result, the greedy algorithm fails to capture the hidden cost of a large understeer, which results in a period of time where the speed must inevitably drop.
 
@@
 path distance $s_k$, there are $M_k$ velocity and $M_k$ tire slip observations from experimental data, each corresponding to a different $\mu$. 
-For example, looking at Fig.~\ref{fig:t2}, if $k = 191$, $s_k =  950 \mathrm{m}$, $M_k = 3$, the velocity and slip norm $\zeta$ observations $U_k(\mu)$ and $Z_k(\mu)$ are as follows:
+For example, looking at Fig.~\ref{fig:t2}, if $k = 191$, $s_k =  950 \mathrm{m}$, $M_k = 3$, the velocity and slip norm $\zeta$ observations $U_k(\mu)$ and $Z_k(\mu)$ are shown in Table \ref{tb:Uk}:
 
@@
 \begin{tabular}{ccc}
+\label{tb:Uk}
 $\mu $&$ U_x(m/s) $& $\zeta $\\\hline
@@
 \centering
-\includegraphics[width=2.8in]{figures/searchTree.png}
+\includegraphics[width=2.2in]{figures/searchTree.png}
 \caption{Sample search tree for Thunderhill race track where there are only 3 experimentally observed full laps at $\mu =$ 0.9, 0.93, 0.95.}
@@
 \centering
-\includegraphics[width=3.5in]{figures/costStructure.eps}
+\includegraphics[width=3in]{figures/costStructure.eps}
 \caption[Illustration of travel cost.]{Illustration of travel cost. Current state is (855, 0.9). Assuming $\Delta s$ is 15 meters, cost is time to travel from this node
@@
 c_s(s_k, s_{k+1},\mu_k, \mu_{k+1}) &= \\
-&\mathds{1}\left(\mu_{k+1} \neq \mu_{k}\right)\left(\lambda  + \infty \times\mathds{1}\left(Z_k(\mu_k) > 1\right) \right)
+&\mathds{1}\left(\mu_{k+1} \neq \mu_{k}\right)\left(\lambda  + \infty \times\mathds{1}\left(Z_k(\mu_k) > 1\right) \nonumber \right)
 \end{align}
@@
 
-Minimum-cost tree spanning algorithms (e.g. breadth-first search, Dijkstra's algorithm, etc.) are a well known subject and a thorough description
-can be found in \cite{aibook}. For the purpose of solving this search problem, the A* search algorithm is used. 
+% Minimum-cost tree spanning algorithms (e.g. breadth-first search, Dijkstra's algorithm, etc.) are a well known subject and a thorough description
+% can be found in \cite{aibook}.
+
+ For the purpose of solving this search problem, the A* search algorithm is used. 
 Like breadth-first search, the A* algorithm is guaranteed to find the lowest cost path from a start node to a goal node, 
@@
 each trial consisting of a speed profile generated with a constant $\mu$ value over the track. The $\mu$ values chosen for
-experimental data collection were 0.85, 0.9, 0.92, 0.93, 0.94, 0.95, and 0.97. Ideally, each speed profile would be tested experimentally for a 
-full lap at high speed.
-However, due to safety and time constraints associated with collecting high speed race data, only the speed profiles corresponding to $\mu = 0.92$ and $\mu = 0.94$ were observed
-over the whole track. The other speed profiles were only tested on sections of the track. Fig.~\ref{fig:muCoverage} shows the experimental data coverage.
+experimental data collection were 0.85, 0.9, 0.92, 0.93, 0.94, 0.95, and 0.97.
+
+Due to safety and time constraints associated with collecting high speed race data, only the speed profiles corresponding to $\mu = 0.92$ and $\mu = 0.94$ were observed
+over the whole track. The other speed profiles were only tested on critical sections of the track.
+
+% Fig.~\ref{fig:muCoverage} shows the experimental data coverage.
 
@@
 
- \begin{figure}[tb]
-\centering
-\includegraphics[width=2.45in]{figures/muCoverage.eps}
-\caption[Coverage of experimental data]{Coverage of experimental data. For speed profiles corresponding to $\mu = 0.92$ and $\mu = 0.94$, experimental data of the autonomous race vehicle was observed
-over the whole track. For safety and time constraints, the other speed profiles were only tested on sections of the track.}
-\label{fig:muCoverage}
-\end{figure}
+%  \begin{figure}[tb]
+% \centering
+% \includegraphics[width=2.45in]{figures/muCoverage.eps}
+% \caption[Coverage of experimental data]{Coverage of experimental data.
+For speed profiles corresponding to $\mu = 0.92$ and $\mu = 0.94$, experimental data of the autonomous race vehicle was observed
+% over the whole track. For safety and time constraints, the other speed profiles were only tested on sections of the track.}
+% \label{fig:muCoverage}
+% \end{figure}
 
-After collecting the experimental data, the A* algorithm was applied to learn the optimal friction profile $\mu^\star(s)$. Parameters for the algorithm
-are shown in Table~\ref{tb:astarparams}. To interface more efficiently with experimental data files, the search algorithm was implemented in MATLAB using
-custom code instead of a standard search algorithm library. The entire search process took approximately 70 seconds on a core i7 laptop machine,
-exploring 6887 nodes in the process. However, since MATLAB is not designed for computational efficiency in tree-based search algorithms, a C++ implementation would likely
-be several orders of magnitude faster.
+% After collecting the experimental data, the A* algorithm was applied to learn the optimal friction profile $\mu^\star(s)$.
+Parameters for the algorithm are shown in Table~\ref{tb:astarparams}.
+To interface more efficiently with experimental data files, the search algorithm was implemented in MATLAB using custom code instead of a standard search algorithm library.
+The entire search process took approximately 70 seconds on a core i7 laptop machine, exploring 6887 nodes in the process.
+However, since MATLAB is not designed for computational efficiency in tree-based search algorithms, a C++ implementation would likely
+% be several orders of magnitude faster.
 
@@
 The same results are plotted on a map of the track in Fig.~\ref{fig:mumap}.
-A satisfying observation is that the optimal profile reduces $\mu$ to 0.93 in section \circled{2} to be more conservative and increases $\mu$ to 0.97 in section \circled{3} to be more aggressive.
+A satisfying observation is that the optimal profile reduces $\mu$ to 0.93 in section \circled{2} to be more conservative.
 This matches our observations about tire slip norm $\zeta$ in Sec.~\ref{sec:etu}.
@@
 \centering
-\includegraphics[width=3.5in]{figures/muprofile.eps}
+\includegraphics[width=2.8in]{figures/muprofile.eps}
 \caption{Minimum time $\mu(s)$ profile for Thunderhill, for both the A* solution and greedy algorithm solution.}
@@
  the vehicle can reverse the steering quickly while following the desired path. Section \circled{8} occurs before a long straight section of the track where recovering
-from an understeer would result in a significantly lower speed on the fastest part of the track. As a result, the algorithm's planned acceleration is more conservative. Finally, 
-the section with the lowest $\mu(s)$ occurs on a part of the track with significant lateral weight transfer. In general, lateral weight transfer reduces the cornering forces available
-to the vehicle, but this effect was not captured in the trajectory planning phase, which assumed a planar model with coupled left and right tires. In summary, the A*
-algorithm allows the car to learn subtle but important driving behaviors that are not easily captured through simulation.
+from an understeer would result in a significantly lower speed on the fastest part of the track. As a result, the algorithm's planned acceleration is more conservative.
+
+% Finally, 
+% the section with the lowest $\mu(s)$ occurs on a part of the track with significant lateral weight transfer. In general, lateral weight transfer reduces the cornering forces available
+% to the vehicle, but this effect was not captured in the trajectory planning phase, which assumed a planar model with coupled left and right tires. In summary, the A*
+% algorithm allows the car to learn subtle but important driving behaviors that are not easily captured through simulation.
 
@@
 Iterative learning control (ILC) formulations gradually determine the proper steering and throttle inputs to precisely track the desired racing trajectory.
-Experimental results at combined vehicle accelerations of 9 $\mathrm{m/s^2}$  indicate that the proposed algorithm can rapidly attenuate trajectory following errors over just two or three laps of racing (\S \ref{sec:ch4ExpRes}).
-This paper also presents a tree-search algorithm to minimize experimental lap times by learning different acceleration limits for each turn on the track.
+Experimental results at combined vehicle accelerations of 9 $\mathrm{m/s^2}$  indicate that the proposed algorithm can rapidly attenuate trajectory following errors over just two or three laps of racing.
+This paper also presents a trajectory modification algorithm to minimize experimental lap times by learning different acceleration limits for each turn on the track.
 An A* search algorithm is devised to search through experimental data and find the best value of $\mu$ for each portion of the track in order to globally minimize the resulting lap time.
@@
 
-This research is supported by the National Science Foun-
-dation Graduate Research Fellowship Program (GRFP). The
+This research was supported by the National Science Foundation Graduate Research Fellowship Program (GRFP). The
 authors would like to thank the members of the Dynamic

### 2019-09-03 conference/iv2019/plotScripts/velocityPlot.m

@@ -1,8 +1,8 @@
 clear all; close all; clc;
-dat1 =load('/home/nkapania/catkin_ws/src/harpagornis/bags/out1.mat');
+dat1 =load('/home/nkapania/catkin_ws/src/harpagornis/bags/out4.mat');
 dat1.lane = 3;
 
-dat2 =load('/home/nkapania/catkin_ws/src/harpagornis/bags/out2.mat');
+dat2 =load('/home/nkapania/catkin_ws/src/harpagornis/bags/out3.mat');
 dat2.lane = 2;
 
 laneWidth = 3.7;

### 2019-09-03 dissertation/images/chapter4/plotSimResults.m

@@ -33,10 +33,10 @@ for i = 1:N
     qsm(i) = SIM_QM(i).eRMS;
 end
 
-semilogy(0:4, pde,'ko-','MarkerFaceColor','k','LineWidth',2);
+semilogy(0:N-1, pde,'ko-','MarkerFaceColor','k','LineWidth',2);
 hold on; grid on;
-semilogy(0:4, qsm,'bs-','MarkerSize',8,'LineWidth',2);
-semilogy(0:4, qse,'r*-','MarkerFaceColor','r','LineWidth',2);
+semilogy(0:N-1, qsm,'bs-','MarkerSize',8,'LineWidth',2);
+semilogy(0:N-1, qse,'r*-','MarkerFaceColor','r','LineWidth',2);
 legend('PD', 'Q-MIMO','Q-SISO');
 
 %xlabel('Iteration Number','FontName','Arial','FontSize',14)
@@ -51,10 +51,10 @@ for i = 1:N
     qsm(i) = SIM_QM(i).eRMS;
 end
 
-semilogy(0:4, pde,'ko-','MarkerFaceColor','k','LineWidth',2);
+semilogy(0:N-1, pde,'ko-','MarkerFaceColor','k','LineWidth',2);
 hold on; grid on;
-semilogy(0:4, qsm,'bs-','MarkerSize',8,'LineWidth',2);
-semilogy(0:4, qse,'r*-','MarkerFaceColor','r','LineWidth',2);
+semilogy(0:N-1, qsm,'bs-','MarkerSize',8,'LineWidth',2);
+semilogy(0:N-1, qse,'r*-','MarkerFaceColor','r','LineWidth',2);
 legend('PD', 'Q-MIMO','Q-SISO');
 
 
@@ -70,10 +70,10 @@ for i = 1:N
     qsm(i) = SIM_QM(i).eRMS;
 end
 
-semilogy(0:4, pde,'ko-','MarkerFaceColor','k','LineWidth',2);
+semilogy(0:N-1, pde,'ko-','MarkerFaceColor','k','LineWidth',2);
 hold on; grid on;
-semilogy(0:4, qsm,'bs-','MarkerSize',8,'LineWidth',2);
-semilogy(0:4, qse,'r*-','MarkerFaceColor','r','LineWidth',2);
+semilogy(0:N-1, qsm,'bs-','MarkerSize',8,'LineWidth',2);
[…]

### 2019-10-13 journal/JDSMC2015/jdsmc2015arxiv.tex
adding in files for IV

\documentclass[twocolumn,10pt]{asme2ej}

\usepackage{epsfig} %% for loading postscript figures
\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage{cite}
\usepackage[cmex10]{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{array}
\usepackage{tikz}
\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
            \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}
\hyphenation{REFERENCE}

%% The class has several options
%  onecolumn/twocolumn - format for one or two columns per page
%  10pt/11pt/12pt - use 10, 11, or 12 point font
%  oneside/twoside - format for oneside/twosided printing
%  final/draft - format for final/draft copy
%  cleanfoot - take out copyright info in footer leave page number
%  cleanhead - take out the conference banner on the title page
%  titlepage/notitlepage - put in titlepage or leave out titlepage
%  
%% The default is oneside, onecolumn, 10pt, final

\title{A Sequential Two-Step Algorithm for Fast Generation of Vehicle Racing Trajectories}

\author{Nitin R. Kapania \thanks{Address all correspondence to this author.}
    \affiliation{
	Dept. of Mechanical Engineering\\
	Stanford University\\
	Stanford, CA 94305\\
	nkapania@stanford.edu
    }	
}

\author{John Subosits
    \affiliation{Dept. of Mechanical Engineering\\
	Stanford University\\
	Stanford, CA 94305\\
	subosits@stanford.edu
    }
}

\author{J. Christian Gerdes
    \affiliation{Dept. of Mechanical Engineering\\
	Stanford University\\
	Stanford, CA 94305\\
	gerdes@stanford.edu
    }
}

\begin{document}

\maketitle

\begin{abstract}
{\it The problem of maneuvering a vehicle through a race course in minimum time requires computation of 
 both longitudinal (brake and throttle) and lateral (steering wheel) control inputs.    
 Unfortunately, solving the resulting nonlinear optimal control problem is typically computationally expensive and infeasible for real-time trajectory planning.
 This paper presents an iterative algorithm that divides the path generation
 task into two sequential subproblems that are significantly easier to solve. Given an initial path through the race track, the algorithm
 runs a forward-backward integration scheme to determine the minimum-time longitudinal speed profile, subject to
 tire friction constraints. With this fixed speed profile, the algorithm updates the vehicle's path by solving a convex optimization problem 
 that minimizes the resulting path curvature while staying within track boundaries and obeying affine, time-varying vehicle dynamics constraints.
 This two-step process is repeated iteratively until the
 predicted lap time no longer improves. While providing no guarantees of convergence or a globally optimal solution, 
 the approach performs very well when validated on the Thunderhill Raceway
 course in Willows, CA. The predicted lap time converges after four to five iterations, with each iteration over the full 4.5 km race course requiring
 only thirty seconds of computation time on a laptop computer. The resulting trajectory is experimentally driven at the race circuit with
 an autonomous Audi TTS test vehicle, and the resulting lap time and racing line is comparable to both a nonlinear gradient
 descent solution and a trajectory recorded from a professional racecar driver. The experimental results indicate that the proposed method is a viable option for 
 online trajectory planning in the near future.}
\end{abstract}

\section{Introduction}
The problem of calculating the minimum lap time trajectory for a given vehicle and race track has been studied over the last several decades
in the control, optimization, and vehicle dynamics communities.  Early research by Hendrikx et al. \cite{hendrikx} in 1996 used 
 Pontryagin's minimum principle to derive coupled differential equations to solve for the minimum-time trajectory
of a vehicle lane change maneuver. The minimum lap time problem drew significant interest from professional racing teams, and
Casanova \cite{casanova} published a method in 2000 capable of simultaneously optimizing both the path and speed profile
for a fully nonlinear vehicle model using nonlinear programming (NLP). Kelly \cite{kelly} further extended the results from Casanova by considering
the physical effect of tire thermodynamics and applying more robust NLP solution methods such as Feasible Sequential Quadratic Programming. More recently,
Perantoni and Limebeer \cite{perantoni} showed that the computational expense could be significantly reduced by 
applying curvilinear track coordinates, non-stiff vehicle dynamics, and the use of smooth computer-generated
analytic derivatives. The authors simulated the optimal vehicle lap on the Catalunya race circuit with a spatial discretization
of two meters and a solve time of around 15 minutes.

More recently, the development of autonomous vehicle technology at the industry and academic level has led to research on optimal path
 planning algorithms that can be used for driverless cars. Theodosis and Gerdes published a gradient descent approach for determining time-optimal
  racing lines, with the racing line constrained to be composed of a fixed number of clothoid segments \cite{theodosis}. Given the computational expense
 of performing nonlinear optimization, there has also been a significant research effort to find approximate methods that provide fast lap times.
 Timings and Cole \cite{timings} formulated the minimum lap time problem into a model predictive control (MPC) problem by linearizing the nonlinear vehicle
 dynamics at every time step and approximating the minimum-time objective by maximizing distance traveled along the path centerline. The resulting racing line
 for a 90 degree turn was simulated next to an NLP solution. Gerdts et al.  \cite{gerdts} proposed a similar receding horizon approach, where distance along a reference path was maximized over a series of locally optimal optimization
 problems that were combined with continuity boundary conditions. One potential drawback of the model predictive control approach is that an optimization
 problem must be reformulated and solved at every time step, which can still be computationally expensive. For example, Timings and Cole reported a computation time of 900 milliseconds
 per 20 millisecond simulation step with the CPLEX quadratic program solver on a desktop PC.

 While experimental validation was reported only by \cite{theodosis} and \cite{gerdts}, all of the aforementioned methods are 
 feasible for experimental implementation, as an autonomous vehicle can apply a closed-loop controller to follow a time-optimal vehicle trajectory computed offline. 
 However, there are  significant benefits to developing a fast trajectory generation algorithm that can approximate the globally optimal trajectory in real-time. 
 If the algorithm runtime is small compared to the actual lap time, the algorithm can run as a real-time trajectory planner and find a fast racing line 
 for the next several turns of the racing circuit. This would allow the trajectory planner to modify the desired path based on the motion of competing race vehicles and 
 estimates of road friction, tire wear, engine/brake dynamics and other parameters learned over several laps of racing. Additionally, the fast trajectory algorithm can
 be used to provide a very good initial trajectory for a nonlinear optimization method.

 This paper therefore presents an experimentally validated iterative algorithm that generates vehicle racing trajectories with low computational expense. 
 To decrease computation time, the combined lateral/longitudinal optimal control problem is replaced by two sequential sub-problems 
 where minimum-time longitudinal speed inputs are computed given a fixed vehicle path, and then the vehicle path is updated given the fixed speed commands.

 The following section presents a mathematical framework for the trajectory generation
 problem and provides a linearized five-state model for the planar dynamics of a racecar following a set of speed and steering inputs, with lateral and heading error states computed with respect to a fixed path.
 Section \ref{sec:VP} describes the method of finding the minimum-time speed inputs given a fixed path.  While this task has been recently formulated as a convex optimization problem \cite{lipp}, a forward-backward
 integration scheme based on prior work \cite{subosits} is used instead.  Section \ref{sec:UPDATE} describes a method for updating the racing path given the fixed speed inputs using convex optimization, where the curvature norm
 of the driven path is explicitly minimized. The complete algorithm is outlined in Section \ref{sec:IMPLEMENT}, and the racing trajectory generated on the Thunderhill Raceway circuit
 in Willows, CA is compared to results from both a nonlinear optimization and data recorded from a professional racecar driver. In Section \ref{sec:EXP}, the racing trajectory is then tested experimentally in
 an autonomous Audi TTS testbed via a previously published closed-loop path following controller \cite{kapaniaVSD}. The resulting lap time compares well with the lap time
 recorded for the nonlinear optimization trajectory. Section \ref{sec:DISCUSSION} concludes by discussing 
 future implementation of the algorithm in a real-time path planner.

\section{Path Description and Vehicle Model}
\label{sec:PATH}
Figure~\ref{fig:worldInfo} describes the parameterization of the reference path that the vehicle will follow. The reference path is most intuitively described in Fig.~\ref{fig:worldInfo}(a) as a 
smooth curve of Cartesian East-North coordinates, with road boundaries represented by similar Cartesian curves. However, for the purposes of quickly generating
a racing trajectory, it is more convenient to parameterize the reference path as a curvature profile $K$ that is a function of distance along the path $s$ (Fig.~\ref{fig:worldInfo}(c)). Additionally, it is 
convenient to store the road boundary information as two functions $w_\mathrm{in}(s)$ and $w_\mathrm{out}(s)$, which correspond to the lateral distance from the path at $s$
 to the inside and outside road boundaries, respectively (Fig.~\ref{fig:worldInfo}(b)). This maximum lateral distance representation will be useful when constraining the generated racing path to lie within the road 
 boundaries. The transformation from the local $s$, $K$ coordinate frame to 
the global Cartesian coordinates $E$, $N$ are given by the Fresnel integrals:
\begin{subequations}
\label{eq:fresnel}
\begin{align}
	E(s) &= \int_0^s  -\sin(\Psi_\mathrm{r}(z)) dz \\
	N(s) &= \int_0^s   \cos(\Psi_\mathrm{r}(z)) dz \\
	\Psi_\mathrm{r}(s) &= \int_0^s K(z) dz \label{eq:balls}
\end{align}
\end{subequations}
where $\Psi_\mathrm{r}(s)$ is the heading angle of the reference path and $z$ is a dummy variable.

 \begin{figure}
\centering
\includegraphics[width=3.4in]{worldInfo.eps}
\caption{(a) View of a sample reference path and road boundaries, plotted in the East-North Cartesian frame (b) Lateral distance from path to inside road edge (positive) and outside road edge (negative) as a function
of distance along path. (c) Curvature as a function of distance along path.}
\label{fig:worldInfo}
\end{figure}

With the reference path defined in terms of $s$ and $K$, the 
next step is to define the dynamic model of the vehicle. For the purposes of trajectory generation, we assume the vehicle dynamics are given 
by the planar bicycle model in Fig.~\ref{fig:bikemodel}(a), with yaw rate $r$ and sideslip $\beta$ states describing the lateral dynamics. Additionally, the vehicle's offset from
the reference path is given by the path lateral deviation state $e$ and path heading error state $\Delta\Psi$ (Fig.~\ref{fig:bikemodel}(b)). Linearized equations of
motion for all four states are given by:
\begin{subequations}
\label{eq:bm}
\begin{align}
	\dot{\beta} &= \frac{F_\mathrm{yf}+F_\mathrm{yr}}{mU_x} - r \qquad \dot{r} = \frac{aF_\mathrm{yf} - bF_\mathrm{yr}}{I_z} \label{bm1} \\
	\dot{e} &= U_x (\beta + \Delta\Psi) \qquad \Delta\dot{\Psi} = r - U_xK \label{eq:bm2} 
\end{align}
\end{subequations}
where $U_x$ is the vehicle forward velocity and $F_\mathrm{yf}$ and $F_\mathrm{yr}$ are the front and rear lateral tire forces. 
The vehicle mass and yaw inertia are denoted by $m$ and $I_z$, and the geometric parameters $a$ and $b$ are shown in Fig.~\ref{fig:bikemodel}(a). Note
that while the vehicle longitudinal dynamics are not explicitly modeled, the bicycle model does allow for time-varying values of $U_x$.

\begin{figure}
\centering
\includegraphics[width=2.5in]{BikeModelSchematic.png}
\caption{(a) Schematic of bicycle model. (b) Diagram showing lateral path deviation $e$ and path heading error $\Delta\Psi$ states.}
\label{fig:bikemodel}
\end{figure}

\section{Velocity Profile Generation Given Fixed Reference Path}
\label{sec:VP}

Given a fixed reference path described by $s$ and $K$, the first algorithm step is to find the minimum-time speed profile the vehicle can achieve
without exceeding the available tire-road friction. The approach taken in this paper is a ``three-pass" approach described in complete detail by 
Subosits and Gerdes \cite{subosits}, and originally inspired by work from Velenis and Tsiotras \cite{velenis}. 
Given the lumped front and rear tires from Fig.~\ref{fig:bikemodel}(a), the available longitudinal
force $F_\mathrm{x}$ and lateral force $F_\mathrm{y}$ at each wheel is constrained by the friction circle: 
\begin{subequations}
\label{eq:tireforce}
\begin{align}
	F^2_\mathrm{xf} + F_\mathrm{yf}^2 &\leq (\mu F_\mathrm{zf})^2\\
	F^2_\mathrm{xr} + F_\mathrm{yr}^2 &\leq (\mu F_\mathrm{zr})^2
\end{align}
\end{subequations}
where $\mu$ is the tire-road friction coefficient and $F_\mathrm{z}$ is the available normal force. The first pass of the speed profile generation aims to find the maximum permissible steady 
state vehicle speed given zero longitudinal force. For the simplified case where weight transfer and topography effects are neglected, this is given by:
\begin{equation}
\label{eq:steadystate}
	U_x(s) = \sqrt{\frac{\mu g}{|K(s)|}}
\end{equation}
where the result in (\ref{eq:steadystate}) is  obtained by setting $F_\mathrm{yf} = \frac{mb}{a+b}U_x^2K$ and $F_\mathrm{zf} = \frac{mgb}{a+b}$.
 The results of this first pass for the sample curvature profile in Fig.~\ref{fig:VPgen}(a) are shown in Fig.~\ref{fig:VPgen}(b).
 The next step is a forward 
integration step, where the velocity of a given point is determined by the velocity of the previous point and the available longitudinal force $F_{x,\mathrm{max}}$ for acceleration.
This available longitudinal force is calculated in \cite{subosits} by accounting for the vehicle engine force limit and the lateral force demand on all tires due to the road curvature:
\begin{equation}
\label{eq:forwardint}
	U_x(s+\Delta s) =\sqrt{U^2_x(s)+\mathrm{2}\frac{F_{x,\mathrm{accel,max}}}{m}\Delta s}
\end{equation}
A key point of the forward integration step is that at every point, the value of $U_x(s)$ is compared to the corresponding value from (\ref{eq:steadystate}), and
the minimum value is taken. The result is shown graphically in Fig.~\ref{fig:VPgen}(c). Finally, the backward integration step occurs, where the available
longitudinal force for deceleration is again constrained by the lateral force demand on all tires:
\begin{equation}
\label{eq:backwardsint}
	U_x(s-\Delta s) = \sqrt{U^2_x(s)-\mathrm{2}\frac{F_{x,\mathrm{decel,max}}}{m}\Delta s}
\end{equation}
The value of $U_x(s)$ is then compared to the corresponding value from (\ref{eq:forwardint}) for each point along the path, and the minimum
value is chosen, resulting in the final velocity profile shown by the solid line in Fig.~\ref{fig:VPgen}(d). While treatment of three-dimensional topography 
effects are not described in this paper, the method described in \cite{subosits} and
used for the experimental data collection determines the normal and lateral tire forces $F_\mathrm{z}$ and $F_\mathrm{y}$ at each point along the path 
by accounting for weight transfer, road bank and grade.

 \begin{figure}
\centering
\includegraphics[width=3.3in]{vpgen.eps}
\caption{(a) Sample curvature profile. (b) Velocity profile given zero longitudinal force. (c) Velocity profile after forward pass (d) Final velocity profile after backward pass. }
\label{fig:VPgen}
\end{figure}

\section{Updating Path Given Fixed Velocity Profile}
\label{sec:UPDATE}
\subsection{Overall Approach and Minimum Curvature Heuristic}
The second step of the trajectory generation algorithm takes the original reference path $K(s)$ and corresponding velocity profile $U_x(s)$
as inputs, and modifies the reference path to obtain a new, ideally faster, racing line. Sharp \cite{sharp} suggests a general approach for modifying 
an initial path to obtain a faster lap time by taking the original path and velocity profile and incrementing the speed uniformly
by a small, constant ``learning rate." An optimization problem is then solved to find a new reference path and control inputs that allow the vehicle to
drive at the higher speeds without driving off the road. If a crash is detected, the speed inputs are locally reduced around the crash site and the process is repeated.

However, one challenge with this approach is that it can take several hundred iterations of locally modifying the vehicle speed profile, detecting crashes, and 
 modifying the reference path to converge to a fast lap time. An alternative approach is to modify the reference path in one step by solving a single 
 optimization problem. The lap time $t$ for a given racing line is provided by the following equation:

 \begin{equation}
t = \int_0^L\frac{ds}{U_x(s)}
\label{integrateEq}
\end{equation}

 Equation (\ref{integrateEq}) implies that minimizing the vehicle lap time requires simultaneously minimizing the total path length $L$ while maximizing
 the vehicle's longitudinal velocity $U_x$. These are typically competing objectives, as lower curvature (i.e. higher radius) paths can result in longer path lengths but higher
 vehicle speeds when the lateral force capability of the tires is reached, as shown in (\ref{eq:steadystate}). As mentioned in the introduction, time-intensive nonlinear programming
 is required to manage this trade-off and directly minimize (\ref{integrateEq}).

 The proposed approach is therefore to simplify the cost function by only minimizing the norm of the vehicle's driven curvature
 $K(s)$ at each path modification step. Path curvature can be easily formulated as a convex function with respect to the vehicle state vector $x$, 
 enabling the path modification step to be easily solved by leveraging the computational speed of convex optimization.

 However, minimizing curvature is not the same as minimizing lap time and provides
 no guarantee of finding the time-optimal solution. The proposed cost function relies on the hypothesis that a path with minimum curvature is a good approximation for
 the minimum-time racing line. Lowering the curvature of the racing line is more important than minimizing
 path length for most race courses, as the relatively narrow track width provides limited room to shorten the overall path length. Simulated and experimental results
 in Sections \ref{sec:IMPLEMENT} and \ref{sec:EXP} will validate this hypothesis by showing similar lap time performance when compared to
 a gradient descent method that directly minimizes lap time. One reason for this comparable performance is that nonlinear methods are
 typically sensitive to initial conditions and are themselves only guaranteed to find a locally optimal solution.

\subsection{Convex Problem Formulation}

Formulating the path update step as a convex optimization problem requires an affine, discretized form of the
bicycle model in (\ref{eq:bm}). The equations of motion are nonlinear because
 the front and rear lateral tire forces become saturated as the vehicle drives near the limits of tire adhesion. 
The well-known brush tire model \cite{Pacejka2012} captures the force saturation of tires as a function of lateral tire slip angle $\alpha$ as follows:
\small
\begin{eqnarray}
\label{eq:fiala}
	F_\mathrm{y\diamond}&=&\begin{cases} -C_{\diamond}\tan\alpha_\diamond + \frac{C_\diamond^2}{3\mu F_\mathrm{z\diamond}} |\tan\alpha_\diamond| \tan\alpha_\diamond \\ \hspace{10mm}- \frac{C_\diamond^3}{27\mu^2F_\mathrm{z\diamond}^2}\tan^3\alpha_\diamond,
\hspace{8mm}  |\alpha_\diamond| < \arctan{\left(\frac{3\mu F_\mathrm{z\diamond}}{C_\diamond}\right)} \\ \\ -\mu F_\mathrm{z\diamond}\text{sgn} \ \alpha_\diamond, \hspace{36mm} \mathrm{otherwise} \end{cases}
\end{eqnarray}
\normalsize
where the symbol $\diamond \in [\mathrm{f},\mathrm{r}]$ denotes the lumped front or rear tire, and $C_\diamond$ is the corresponding tire cornering stiffness. 
The linearized tire slip angles $\alpha_\mathrm{f}$ and $\alpha_\mathrm{r}$ are functions of the vehicle lateral states and the steer angle
input, $\delta$:
\begin{subequations}
\begin{align}
	\alpha_\mathrm{f} &= \beta + \frac{ar}{U_x} - \delta\\
	\alpha_\mathrm{r} &= \beta - \frac{br}{U_x}
\end{align}
\end{subequations}
The tire model (\ref{eq:fiala}) can be linearized at every point along the reference path assuming steady state cornering conditions:
\begin{subequations}
\begin{align}
	F_\mathrm{y\diamond} &= \tilde{F}_\mathrm{y\diamond} - \tilde{C}_\diamond(\alpha_\diamond - \tilde{\alpha}_\diamond) \\
	\tilde{F}_\mathrm{y\diamond} &= \frac{F_\mathrm{z\diamond}}{g} U_x^2K
\label{eqn:ftil}
\end{align}
\end{subequations}
with parameters $\tilde{F}_\mathrm{y}$, $\tilde{\alpha}$ and $\tilde{C}$ shown in Fig.~\ref{fig:fiala}. The affine, continuous bicycle model with
steering input $\delta$ is then written in state-space form as:
\begin{subequations}
\label{eq:cts}
\begin{align}
	\dot{x}(t) &= A(t) x + B(t)\delta + d(t)\\
	 x &= [e \hspace{2mm} \Delta\Psi \hspace{2mm} r \hspace{2mm} \beta \hspace{2mm} \Psi]^T
\end{align}
\end{subequations}
where we have added a fifth state, vehicle heading angle $\Psi$, defined as the time integral of yaw rate $r$.
 This makes explicit computation of the minimum curvature path simpler. The state matrices $A(t)$, $B(t)$, and affine term $d(t)$ 
 are given by:
\begin{multline}
\label{eqn:Amatrix}
A(t)  =  \\
\left[\begin{matrix}
  0 & U_x(t) & 0 & U_x(t) & 0\\ 
  0 & 0 & 1 & 0 & 0 \\ 
  0 & 0  & \frac{-(a^2\tilde{C}_\mathrm{f}(t)+b^2\tilde{C}_\mathrm{r}(t))}{U_x(t)I_\mathrm{z}} & \frac{b\tilde{C}_\mathrm{r}(t) - a\tilde{C}_\mathrm{f}(t)}{I_\mathrm{z}} & 0 \\
  0 & 0  & \frac{b\tilde{C}_\mathrm{r}(t)-a\tilde{C}_\mathrm{f}(t)}{mU_x^2(t)}-1 & \frac{-(\tilde{C}_\mathrm{f}(t) + \tilde{C}_\mathrm{r}(t))}{mU_x(t)} & 0 \\
  0 & 0 & 1 & 0 & 0
 \end{matrix}\right]
 \end{multline}
\begin{align}
B(t) =\left[0 \hspace{2 mm} 0 \hspace{3 mm} \frac{a \tilde{C}_\mathrm{f}(t)}{I_\mathrm{z}} \hspace{3 mm}  \frac{\tilde{C}_\mathrm{f}(t)}{mU_x(t)} \hspace{3mm} 0\right]^T
\end {align}
\begin{align}
\label{eqn:Dmatrix}
d(t) = \left[\begin{matrix} 0 \\
               -K(t) U_x(t) \\ 
			    \frac{a\tilde{C}_\mathrm{f}(t)\tilde{\alpha}_\mathrm{f}(t) - b\tilde{C}_\mathrm{r}(t)\tilde{\alpha}_\mathrm{r}(t) + a\tilde{F}_\mathrm{yf}(t) - b\tilde{F}_\mathrm{yr}(t)}{I_z} \\
				\frac{\tilde{C}_\mathrm{f}(t)\tilde{\alpha}_\mathrm{f}(t) + \tilde{C}_\mathrm{r}(t)\tilde{\alpha}_\mathrm{r}(t) + \tilde{F}_\mathrm{yf}(t) + \tilde{F}_\mathrm{yr}(t)}{mU_x(t)}\\
				0
				\end{matrix}\right]
\end{align}

With the nonlinear model now approximated as an affine, time-varying model, updating the path is accomplished by solving the following
 convex optimization problem:

\begin{subequations}
\label{eq:OPT}
\begin{alignat}{3}
\underset{\delta, \hspace{.5mm} x}{\text{minimize}} \quad & \sum_{k} \left(\frac{\Psi_k - \Psi_{k-1}}{s_k - s_{k-1}}\right)^2 + \lambda(\delta_k - \delta_{k-1})^2 & \quad  \label{eq:obj}\\
{\text{subject to}} \quad & \rlap{$ x_{k+1}= A_k x_k + B_k \delta_k + d_k$} \label{jenny}\\
& \rlap{$w^\mathrm{out}_k \leq e_k \leq w^\mathrm{in}_k $} \label{bobby}\\
& \rlap{$x_1 = x_T$\label{keylani}}
\end{alignat}
\end{subequations}
where $k = 1 \dots T$ is the discretized time index, and $A_k$, $B_k$, and $d_k$ are discretized versions of the continuous state-space
equations in (\ref{eq:cts}). The objective function (\ref{eq:obj}) minimizes the curvature norm of the path driven by the vehicle, as path curvature is
the derivative of the vehicle heading angle with respect to distance along the path $s$ (\ref{eq:balls}). To maintain convexity of the objective
function, the term ${s_k - s_{k-1}}$ is a constant rather than a variable, and is updated for the next iteration after the optimization has been completed (see Section \ref{sec:IMPLEMENT}). 
Additionally, there is a regularization term with weight $\lambda$ added in the cost function to ensure a smooth steering profile for experimental 
implementation.

The equality constraint (\ref{jenny}) ensures the vehicle follows the affine lateral dynamics. The inequality
 constraint (\ref{bobby}) allows the vehicle to deviate laterally from the reference path to find a new path with lower curvature, but
 only up to the road edges. Finally, the equality constraint (\ref{keylani}) is required for complete racing circuits to ensure the generated
 racing line is a continuous loop. The results of running the optimization 
 are shown for an example turn in Fig.~\ref{fig:hairpin}. The reference path starts out at the road centerline, and the optimization finds 
 a modified path that uses all the available width of the road to lower the path curvature.

  \begin{figure}
\centering
\includegraphics[width=3.7in]{fiala.eps}
\caption{Nonlinear tire force curve given by Fiala model, along with affine tire model linearized at $\alpha = \tilde{\alpha}$. }
\label{fig:fiala}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=3.25in]{updatedpath.eps}
\caption{Path update for an example turn.}
\label{fig:hairpin}
\end{figure}

\section{Algorithm Implementation and Simulated Results}
\label{sec:IMPLEMENT}
\subsection{Algorithm Implementation}
The final algorithm for iteratively
generating a vehicle racing trajectory is described in Algorithm~\ref{alg:genTraj}. The input to the algorithm is any initial path through the racing circuit, 
parameterized in terms of distance along the path $s$, path curvature $K(s)$, and the lane edge distances $w_\mathrm{in}(s)$ and 
$w_\mathrm{out}(s)$ described in Fig.~\ref{fig:worldInfo}. Given the initial path, the minimum-time speed profile $U_x(s)$ is calculated as described in 
Fig.~\ref{fig:VPgen}. Next, the path is modified by solving the previously described minimum curvature
 convex optimization problem (\ref{eq:OPT}).

 \begin{algorithm}
 \caption{Method for Iterative Trajectory Generation}
  \label{alg:genTraj}
\begin{algorithmic}[1]
\Procedure{GenerateTrajectory}{$s^\circ, K^\circ, w_\mathrm{in}^\circ, w_\mathrm{out}^\circ$}
\State $\mathrm{path}\gets (s^\circ, K^\circ, w_\mathrm{in}^\circ, w_\mathrm{out}^\circ)$
\While{$\Delta t^\star > \epsilon$}
\State $U_x \gets \mathrm{calculateSpeedProfile(path)}$
\State $\mathrm{path} \gets \mathrm{minimizeCurvature}(U_x, \mathrm{path})$
\State $t^\star \gets \mathrm{calculateLapTime}(U_x,\mathrm{path})$
\EndWhile
\State \textbf{return} $\mathrm{path},U_x$
\EndProcedure
\end{algorithmic}
\end{algorithm}
%\caption{Iterative algorithm for fast generation of vehicle trajectories. Each iteration consists of a sequential two-step approach where
%the velocity profile is generated given a fixed path and then the path is updated based on the solution from a convex optimization problem.}
%\label{algorithmDesc}
%\end{figure}

The optimization only solves explicitly for the steering input $\delta^\star$ and resulting vehicle lateral states $x^\star$ at every 
time step. Included within $x^\star$  is the optimal vehicle heading $\Psi^\star$ and lateral deviation $e^\star$ from the initial path. To obtain 
the new path in terms of $s$ and $K$, the East-North coordinates ($E_k$, $N_k$) of the updated vehicle path are updated as follows:

\begin{subequations}
\begin{align}
	E_k &\gets E_k - e^\star_k\cos(\Psi_{\mathrm{r},k})\\
	N_k &\gets N_k - e^\star_k\sin(\Psi_{\mathrm{r},k})
\end{align}
\end{subequations}
where $\Psi_\mathrm{r}$ is the path heading angle of the original path. Next, the new path is given by the following numerical approximation:
 \begin{subequations}
 \label{eq:pupdate}
\begin{align}
	s_k &= s_{k-1} + \sqrt{(E_k - E_{k-1})^2 + (N_k - N_{k-1})^2}\\ 
	K_k &= \frac{\Psi^\star_k - \Psi^\star_{k-1}}{s_k - s_{k-1}}
\end{align}
\end{subequations}
 Notice that (\ref{eq:pupdate}) accounts for the change in the total path length that occurs when the vehicle deviates from the original path.
 In addition to $s$ and $K$, the lateral distances to the track edges $w_\mathrm{in}$ and $w_\mathrm{out}$ are different for the new path
 as well, and are recomputed using the Cartesian coordinates for
the inner and outer track edges and ($E_k$, $N_k$). The two-step procedure is iterated until the improvement in lap time $\Delta t^\star$ over the prior iteration is less than a small positive 
constant $\epsilon$.

 \begin{figure}
\centering
\includegraphics[width=3.5 in]{shelleyPic.jpg}
\caption{Audi TTS used for simulation parameters and experimental validation}
\label{fig:shelleyPic}
\end{figure}

\begin{table}[h]
\begin{center}
\small
\caption{Optimization Parameters}\label{tb:params}
\begin{tabular}{lccc}
Parameter & Symbol & Value & Units \\\hline
Regularization Parameter & $\lambda$ & 1 & $\mathrm{1/m}^2$\\
Stop Criterion           & $\epsilon $ & .1 & s \\
Vehicle mass & $m$ & 1500 & kg \\
Yaw Inertia & $I_z$ & 2250 & $\mathrm{kg \cdot m}^2$\\
Front axle to CG & $a$ & 1.04 & m\\
Rear axle to CG & $b$ & 1.42 & m\\
Front cornering stiffness & $\mathrm{C}_\mathrm{f}$ & 160 & $\mathrm{kN \cdot rad}^{-1}$ \\
Rear cornering stiffness & $\mathrm{C}_\mathrm{r}$ & 180 & $\mathrm{kN \cdot rad}^{-1}$ \\
Friction Coefficient     & $\mu $                  &  0.95 & $\mathrm{-} $ \\
Path Discretization      & $\Delta s$              & 2.75 & $m$\\
Optimization Time Steps  & $T       $              & 1843 & -  \\
Max Engine Force         & -                       & 3750 & N\\\hline
\end{tabular}
\end{center}
\end{table}

\subsection{Algorithm Validation}
The proposed algorithm is tested on the 4.5 km Thunderhill racing circuit in Willows, California, USA. The vehicle parameters 
used for the lap time optimization come from an Audi TTS experimental race vehicle (Fig.~\ref{fig:shelleyPic}), and are shown along with the optimization parameters in Table 1.
 The initial path is obtained by collecting GPS data of the inner and outer track edges and estimating the ($s, K, w_\mathrm{in}, w_\mathrm{out}$)
 parametrization of the track centerline via a separate curvature estimation subproblem similar to the one proposed in \cite{perantoni}.
 The algorithm is implemented in MATLAB, with the minimum curvature optimization problem (\ref{eq:OPT}) solved using the CVX software package \cite{boydcvx}.

 \begin{figure*}
\centering
\includegraphics[width=4.5in]{racinglines.eps}
\caption{Overhead view of Thunderhill Raceway along with generated path from algorithm. Car drives in alphabetical direction around the closed circuit. Labeled regions a-h are locations of discrepancies between the 
two-step algorithm solution and comparison solutions.}
\label{racingLines}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=6.5in]{deviationFromCenterline.eps}
\caption{Lateral path deviation of racing line from track centerline as a function of distance along the centerline. Note that upper and lower bounds on $e$ are not always symmetric
due to the initial centerline being a smooth approximation. Results
are compared with racing line from a nonlinear gradient descent algorithm and experimental data recorded from a professional racecar driver.}
\label{fig:pathDeviation}
\end{figure*}

\subsection{Comparison with Other Methods}
The generated racing path after five iterations is shown in
Fig.~\ref{racingLines}. To validate the proposed algorithm, the racing line is compared with results from a nonlinear gradient descent algorithm implemented by 
Theodosis and Gerdes \cite{theodosis} and an experimental trajectory recorded from a professional racecar driver in the testbed vehicle (Fig.~\ref{fig:shelleyPic}).
 While time-intensive to compute, the gradient descent approach generates racing lines with autonomously driven lap times within one second of
lap times measured from professional racecar drivers.

To better visualize the differences between all three racing lines, Fig.~\ref{fig:pathDeviation} shows the lateral deviation from the track centerline as a function of distance along the centerline 
for all three trajectories. The left and right track boundaries $w_\mathrm{in}$ and $w_\mathrm{out}$ are plotted as well. Note that the two-step iterative algorithm provides a 
racing line that is qualitatively similar to the racing lines provided by the nonlinear gradient descent and human driver data. In particular, all three
solutions succeed at effectively utilizing all of the available track width whenever possible, and strike similar apex points on each
of the circuit's 15 corners.

However, there are several locations on the track where there is a significant discrepancy on the order of several meters between the two-step algorithm's trajectory and the 
other comparison trajectories. These locations of interest are labeled \circled{a} through \circled{h} in Fig.~\ref{racingLines}.
Note that sections \circled{a}, \circled{e}, \circled{f}, and \circled{g} all occur on large, relatively straight portions of the racing circuit.
In these straight sections, the path curvature is relatively low and differences in lateral deviation from the track centerline have a relatively small
effect on the lap time performance.

\begin{figure}[h]
\centering
\includegraphics[width=2.25in]{composite1.eps}
\caption{Racing lines from the two-step fast generation approach, nonlinear gradient descent algorithm, and experimental data taken
from professional driver. Car drives in direction of labeled arrow.}
\label{compositeFig1}
\end{figure}

Of more significant interest are the sections labeled \circled{b}, \circled{c}, \circled{d}, and \circled{h}, which all occur at turning regions of the track.
These regions are plotted in Fig.~\ref{compositeFig1} and Fig.~\ref{compositeFig2} for zoomed-in portions of the race track. While it is difficult to analyze a single turn
of the track in isolation, discrepancies can arise between the two-step fast generation method and the gradient descent as the latter method trades off 
between a minimum curvature path and the path with shortest total distance. As a result, the gradient descent method finds
 regions where it may be beneficial to use less of the available road width in order to reduce the total distance
traveled.

\begin{figure}
\centering
\includegraphics[width=2.25in]{composite2.eps}
\caption{Racing lines from the two-step fast generation approach, nonlinear gradient descent algorithm, and experimental data taken
from professional driver. Car drives in direction of labeled arrow.}
\label{compositeFig2}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=3in]{laptimes.eps}
\caption{Lap time as a function of iteration for the two-step fast trajectory generation method. Final lap time is comparable
to that achieved with the nonlinear gradient descent approach. Iteration zero corresponds to the lap time for driving the track centerline.}
\label{lapTimes}
\end{figure}

\begin{figure*}[tb]
\centering
\includegraphics[width=6.8 in]{simData.eps}
\caption{(a) Predicted time difference between a car driving both trajectories, with a positive
value corresponding the two-step algorithm being ahead. (b) Curvature profile $K(s)$ plotted vs. distance along the path $s$.
 (c) Velocity profile $U_x(s)$ plotted 
vs. distance along the path $s$ for the two-step method and nonlinear gradient descent method.}
\label{fig:simData}
\end{figure*}

In region \circled{b}, for example, the fast generation algorithm exits the turn and gradually approaches the left side in order to create space for the
upcoming right-handed corner.  The nonlinear optimization, however, chooses a racing line that stays toward the right side of the track. In this case,
the behavior of the human driver more closely matches that of the two-step fast generation algorithm. The human driver also drives closer to the 
fast generation solution in \circled{h}, while the gradient descent algorithm picks a path that exits the corner with a larger radius. In section \circled{c}, the gradient descent
algorithm again prefers a shorter racing line that remains close the the inside edge of the track, while the two-step algorithm drives all the way
to the outside edge while making the right-handed turn. Interestingly, the human driver stays closer to the middle of the road, but more closely
follows the behavior of the gradient descent algorithm. However, there are also regions of the track where the computational algorithms pick a similar path that 
differs from the human driver, such as region \circled{d}.

\subsection{Lap Time Convergence and Predicted Lap Time}
Fig.~\ref{lapTimes} shows the predicted lap time for each iteration of the fast generation algorithm, with step 0 corresponding
to the initial race track centerline trajectory. The lap time was estimated after each iteration by numerically simulating a vehicle 
 following the desired path and velocity profile using a closed-loop controller. The equations of motion for the simulation
were the nonlinear versions of (\ref{eq:bm}) with tire forces given by
the brush tire model in (\ref{eq:fiala}).

Fig.~\ref{lapTimes} shows that the predicted lap time converges monotonically over four or five iterations, with significant 
improvements over the centerline trajectory occuring over the first two iterations. The predicted minimum lap time of 136.4 seconds
is similar to the predicted lap time of 136.7 seconds from the nonlinear gradient descent approach, 
although in reality, the experimental lap time will depend significantly on unmodelled effects such 
as powertrain dynamics.

The final curvature and velocity profile  for the two-step fast generation method is compared with the equivalent profiles for the gradient
descent algorithm in Fig.~\ref{fig:simData}. Notice that the piecewise linear nature of the 
nonlinear gradient descent method is due to the clothoid constraint imposed by Theodosis and Gerdes \cite{theodosis} for ease of autonomous path following.
In general, the curvature and velocity profiles are very similar, although the fast generation algorithm results in a velocity profile
with slightly lower cornering speeds but slightly higher top speeds. The predicted time difference between a car driving both trajectories 
is shown in Fig.~\ref{fig:simData}(a), with a positive
value corresponding the two-step algorithm being ahead. The trajectory from the two-step algorithm is predicted to outperform the 
gradient descent trajectory from \circled{a}-\circled{c}, lose time from \circled{c}-\circled{e}, and gain time from \circled{e}-\circled{h}.

\begin{figure*}
\centering
\includegraphics[width=6.5 in]{expdata.eps}
\caption{Experimental data for an autonomous vehicle driving the trajectories provided by the two-step fast generation and gradient descent algorithms.(a) Relative time difference
between vehicle driving both trajectories, with a positive time corresponding to the two-step algorithm being ahead. (b) Actual
recorded velocity of vehicle. (c) Difference between actual and desired speed. Large negative values outside plotting range occur 
 on straight sections of the track where the vehicle is limited by engine power and speed tracking error is poorly defined.}
\label{fig:expdata}
\end{figure*}

\section{Experimental Validation} 
\label{sec:EXP}
While the two-step algorithm works well in simulation, the most critical validation step is to have an
autonomous race car drive the generated trajectory. This was accomplished by collecting experimental data on ``Shelley" (Fig.~\ref{fig:shelleyPic}), an
Audi TTS developed jointly by Stanford University and Audi's Electronics Research Laboratory (ERL). The TTS is equipped 
with an electronic power steering motor for autonomous steering and active brake booster and throttle by wire for
longitudinal control.

Autonomous closed-loop following of the racing trajectory is accomplished by using an integrated Differential Global Positioning System (DGPS) 
and Inertial Measurement Unit (IMU) to obtain the global vehicle position and velocity. A localization algorithm is applied to find the vehicle's position along the desired path ($s$) and lateral/heading deviation ($e$ and $\Delta\Psi$) from the path. A previously developed feedback-feedforward steering algorithm \cite{kapaniaVSD} 
is then applied to keep the vehicle following the desired path at high lateral and longitudinal accelerations. By locating the desired position along the path, the current vehicle speed can be referenced to the desired vehicle speed from Fig.~\ref{fig:simData}, 
and a simple proportional speed tracking controller with feedforward can be applied to track the desired speed profile. The same 
controller setup is also used to experimentally drive the trajectory generated by the nonlinear gradient descent algorithm. 
Closed loop control is applied at a sample rate of 200 Hz using a dSPACE MicroAutobox unit.  The experimental data 
for an autonomous lap of driving is shown in Fig.~\ref{fig:expdata} for both the two-step trajectory and the trajectory from the gradient descent.

The resulting experimental lap time for the iterative two-step algorithm was 138.6 seconds, about 0.6 seconds faster
than the experimental lap time for the gradient descent algorithm (139.2 seconds). For safety reasons, the trajectories were generated using a conservative peak road friction value of $\mu = 0.90$,
 resulting in peak lateral and longitudinal accelerations of 0.9$g$. In reality, the true friction value of the road varies slightly, but is closer to $\mu = 0.95$ on average. As a
result, both of these lap times are slightly slower than the fastest lap time recorded by a professional race car driver (137.7 seconds) and the predicted lap times from Section \ref{sec:IMPLEMENT}. A summary of all lap times is provided in Table \ref{tb:laptimes}.

\begin{table}[h]
\begin{center}
\begin{tabular}{c|cc}
    & Simulation & Experiment \\\hline
Fast Generation& 136.4 & 138.6 \\
Gradient Descent&  136.7 & 139.2 \\
Human Driver& N/A & 137.7 \\\hline
\end{tabular}
\caption{Lap Times in Seconds}\label{tb:laptimes}
\end{center}
\end{table}
The experimental data in Fig.~\ref{fig:expdata} generally matches the simulated results in Fig.~\ref{fig:simData}. The simulation predicted the trajectory from the iterative two-step algorithm would be 0.3 seconds
 quicker than that of the nonlinear algorithm, compared to the 0.6 second speed advantage observed experimentally. 
 The simulation also predicted a relative time advantage for the two-step algorithm from sections \circled{a} to \circled{c}
 and from \circled{e} to \circled{h}, a trend seen in the experimental data as well. The two-step algorithm has relatively poor performance
 from section \circled{c} to \circled{d}. This portion of the track
corresponds to the sharp right-handed turn shown in 
Fig.~\ref{compositeFig1}, where the two-step solution differs significantly from the human driver data and the gradient descent solution. This
turn also occurs on a steep downhill segment of the track, which was not accounted for by the fast generation algorithm. The experimental results
 indicate the minimum curvature heuristic is relatively poor for this particular turn when compared to a nonlinear algorithm 
 that explicitly minimizes lap time. Accounting for three-dimensional topography effects in the curvature minimization or adding a term in the cost function to minimize
 distance traveled may improve the performance in the future.

 Another reason for variation between the simulated and experimental time difference plots is variation in speed tracking. 
 The speed tracking error for both racing lines is shown in Fig.~\ref{fig:expdata}(c). Interestingly, while the same speed tracking controller was used to test both racing lines, 
 the controller has slightly better speed tracking performance when running the trajectory from the nonlinear optimization. This is possibly due to the
 longitudinal controller gains being originally tuned on the trajectory from Theodosis and Gerdes \cite{theodosis}.

\section{Discussion and Future Work}
\label{sec:DISCUSSION}

The primary benefit of the proposed algorithm is not improved lap time performance
over the nonlinear algorithm but rather a radical improvement in computational simplicity and speed. Each two-step iteration of the full
course takes only 26 seconds on an Intel i7 processor, whereas the nonlinear algorithm from \cite{theodosis} typically runs over the course of several hours on the same
machine. The most significant computational expense for the proposed algorithm is solving the convex curvature minimization problem for
all 1843 discrete time steps $T$ over the 4.5 km racing circuit.

\begin{table}[h]
\begin{center}
\caption{Iteration Computation Time}\label{tb:solvetime}
\begin{tabular}{ccc}
Lookahead (m) & $T$ & Solve Time (s) \\\hline
450& 184 &5 \\
900&  369 & 6 \\
1800& 737 & 12 \\
4500& 1843& 26 \\\hline
\end{tabular}
\end{center}
\end{table}
This computational efficiency will enable future work to incorporate the trajectory modification algorithm as an online
 ``preview" path planner, which would provide the desired vehicle trajectory for an upcoming portion of the race track. 
 Since the computation time of the algorithm is dependent on the preview distance, 
 the high-level planner would not need to run at the same sample time as the vehicle controller. Instead, the planner would operate on a separate CPU and provide a velocity profile and racing line for only
 the next 1-2 kilometers of the race track every few seconds, or plan a path for the next several hundred meters within a second.

 Table \ref{tb:solvetime} shows problem solve times for a varying range of lookahead lengths with the same discretization $\Delta s$,
 and shows that the runtime scales roughly linearly with the lookahead distance. The above solve times are listed using the CVX convex optimization solver, which 
 is designed for ease of use and is not optimized for embedded computing. Preliminary work has been successful
 in implementing the iterative two-step algorithm into C code using the CVXGEN software tool \cite{boydcvxgen}. When written
 in optimized C code, the algorithm can solve the curvature minimization problem (\ref{eq:OPT}) in less than 0.005 seconds for a lookahead distance of 650 meters.

 The possibility of real-time trajectory planning for race vehicles creates several fascinating areas of future research. An automobile's surroundings are subject to both rapid and
 gradual changes over time, and adapting to unpredictable events requires a real-time trajectory planning algorithm.
 On a short time scale, the real-time trajectory planner could find a fast but stable recovery trajectory in the event of the race vehicle entering an understeer or oversteer 
 situation. On an intermediate time scale, the fast executing two-step algorithm could continuously plan a racing line in the presence of other moving race vehicles by constraining the
 permissible driving areas to be collision-free convex ``tubes" \cite{erlien}.  Finally, the algorithm could update the racing trajectory given estimates of the friction coefficient and other vehicle parameters learned gradually over
 several laps of racing.

\section{Conclusion}
This paper demonstrates an iterative algorithm for quickly generating vehicle racing trajectories, where each iteration is comprised of
 a sequential velocity update and path update step. Given an initial path through the race track, the 
 velocity update step performs forward-backward integration to determine the minimum-time speed inputs. Holding this speed 
 profile constant, the path geometry is updated by solving a convex optimization problem to minimize path curvature. Experimental data confirms that the results
 generated by the algorithm for the Thunderhill Raceway circuit are comparable to those from a nonlinear gradient descent algorithm, with the primary advantage being a much
 faster computation time. An exciting opportunity for future research is incorporating the trajectory modification algorithm into an online path planner to provide
 racing trajectories in real time.

\begin{acknowledgment}
The authors would like to thank Marcial Hernandez for assistance with fitting an initial curvature profile from GPS point
cloud data, and Vadim Butakov and Rob Simpson from the Audi Electronics Research Lab in Belmont, CA. The authors would
also like to thank Vincent Laurense, Samuel Schacher, and Jon Pedersen for assistance with autonomous data collection.
Kapania and Subosits are both supported by Stanford Graduate Fellowships. 
\end{acknowledgment}

\bibliographystyle{asmems4}
\bibliography{jdsmc2015}

\end{document}

### 2019-10-13 unpublished/SciRob Intro Outline/misc.tex
adding in files for IV

% Document settings
\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{amsmath,amssymb,gensymb,epsfig, epstopdf}
\usepackage{xcolor}
\pagestyle{plain}
\setlength\parindent{0pt}
\setlength{\parskip}{1em}

\begin{document}

% Header
\textbf{Nitin Kapania} \\
\textbf{SciRob Intro - Outline} \\

\begin{enumerate}
	\item Autonomous vehicles rely on control systems to automate following of a desired trajectory. These control systems are presently designed around a physical model of the vehicle.

	\subitem Include references to several overview papers on path planning / control
	for autonomous vehicles

	\subitem Control systems for autonomous vehicles currently rely on a  physics-based model either explicitly in the case of predictive control algorithms or implicitly for traditional feedforward-feedback architectures

	\subitem Control systems based on physical models have clear advantages and have demonstrated great performance under known driving conditions.

	\subitem Mention advantages of physics-based models: simplicity in terms of required parameters, intuition, and correctness from known first principles.

	\item However, physics-based models have several drawbacks, particularly when operating conditions change or when dynamics are difficult to model.

	\subitem Physics-based models are designed around a particular set of parameters and a particular operating condition. These can often change in safety critical situations. Provide example of car designed around asphalt that suddenly encounters a low-friction surface.

	\subitem Designer must explicitly decide about model complexity and manually encode logic for multiple operating conditions.

	\subitem Certain physics that are critical for control can be prohibitive to model given difficulty of estimating parameters or hard-to-model physics (e.g. effect of suspension dynamics at the handling limits)

	\item Given drawbacks of physics-based models, there has been increasing interest in developing data-driven models for automated systems.

	\subitem Reference prior art here - e.g. Punjani and Abbeel, Cole, Ghazizadeh, and papers that don't use neural networks (e.g. regression models, GPs, etc.)

	\subitem Research with neural networks dates back to early 1990's, but  explosion of deep learning methods has created potential for models with even higher capacity

	\item This paper contributes a novel neural network model for an automated vehicle.

	\subitem Describe contributions and differentiation from prior art

\end{enumerate}

\end{document}

### 2019-10-13 unpublished/notation clarification/notesOnNotation.tex
adding in files for IV

--- 2018-10-02 unpublished/pedestrian/pedestrianInteraction.tex
@@
 \textbf{Nitin Kapania} \\
-\textbf{Modeling Pedestrian - AV Interaction at a Mid-Block Intersection} \\
+\textbf{Notes on Vehicle Dynamics Notation} \\
 
-\section{Problem Formulation}
-Consider the problem of an autonomous vehicle attempting to drive past an unsignalized, ``mid-block" crosswalk. A pedestrian is simultaneously walking towards the intersection.
+The vehicle dynamics notes previously introduced use the following notation:
 
 \begin{center}
-   \includegraphics[width=.8\textwidth]{Images/example.eps}
-   \label{img:example}
+\begin{tabular}{ l | r }
+ $v_y$ & Vehicle lateral velocity in body-fixed frame \\ 
+ $v_x$ & Vehicle longitudinal velocity in body-fixed frame \\ 
+ $w_z$ & Vehicle yaw rate \\   
+\end{tabular}
 \end{center}
 
-In this case, what should the autonomous vehicle do?
-Should the vehicle always stop as soon as a pedestrian is observed near a cross-walk?
-Can the AV make the decision to speed up and cross before the pedestrian is in the intersection?
-If so, how can it communicate intent to the pedestrian?
-Likewise, can it interpret signals from the pedestrian to understand whether the pedestrian is set on crossing?
+For the next two lectures, we will use the following notation:
 
-Uncontrolled mid-block intersections, like the one shown in Fig.~\ref{img:example}, pose the ambiguity of right of way.
-These laws are dictated on a state-by-state basis.
-While several states (e.g.
-Minnesota, Georgia, Maryland) have "must-stop" rules for uncontrolled intersections, other states have ``must yield" rules.
-For example, vehicles in Arkansas must yield once a pedestrian is in any portion of the roadway, while vehicles in Alabama require motorists to yield to a pedestrian either on the same half of the road or approaching from the opposite side of the roadway.
-
-Given this ambiguity, there is a clear need to study the interaction between the pedestrian and the autonomous vehicle.
-Most human drivers and pedestrians do not know the explicit laws of their state, but safely navigate pedestrian interactions nonetheless.
-
-Several research questions can be posed for the following situation:
-
-\begin{itemize}  
-	\item Can we come up with a mathematical formulation describing this situation? 
-	\item Can we superimpose this mathematical model with real-world data of pedestrian intersection crossings? 
-	\item Can this formulation, combined with real-world data, be used to predict realistic 
-	interaction scenarios?
-	\item Can we design control policies for the vehicle that maintain a high standard of safety while also enabling robust traffic flow 
-\end{itemize}
-
-\section{Prior Work}
-
-There is a wide body of literature segmenting the work on pedestrian behavior.
-In terms of purely modeling the dynamics of pedestrian motion, Batkovic et al.
-\cite{Batkovic} presented a mathematical model to predict pedestrian motion over a finite horizon, with a focus on low computational complexity for autonomous driving.
-Karasev et al.
-modeled pedestrian behavior as intent using a Markov Decision Process \cite{Karasev2016}.
-
-Another segment of the literature is focused on developing predictive models for pedestrian crossing.
-These approaches typically study the parallel issues of pedestrian crossing behavior or driver yielding behavior and date as early as 1975 \cite{Katz1975}.
-Schroeder and Rouphail \cite{Schroeder2011} explored factors associated with driver yielding behavior at unsignalized pedestrian crossings.
-Using logistic regression, the authors found that drivers are more likely to yield to assertive pedestrians who walk briskly in their approach to a crosswalk.
-Kadali and Perumal \cite{RaghuramKadali2012} studied the ``gap acceptance" behavior of pedestrians at mid-block crosswalks through a video graphic survey, and found that the gap accepted by crossing was explained by factors such as crossing direction, vehicle speed, and pedestrian age.
-Yannis et al.
-\cite{Yannis2013} also found that gap acceptance was influenced by the size of the oncoming vehicle and the presence of other pedestrians.
-Gap acceptance models were also proposed by \cite{Sun2002}.
-Lee and Aty \cite{Lee2005} studied interactions in the form of crashes, and found crashes were linked with higher daily traffic.
-
-Another branch of literature specifically studies the interaction between pedestrians and autonomous vehicles.
-A review of these studies was conducted by \cite{Rasouli}.
-Rothenbucher et al.
-\cite{Rothenbucher2016} studied the interaction between pedestrians and driverless vehicles by constructing a car seat costume to disguise a driver.
-The authors noted that pedestrians overall managed interactions at crosswalks effectively, but later mentioned increased uncertainty about the car's behavior.
-To improve the issue of trust, several researchers have developed external interfaces to more clearly broadcast the intent of the autonomous vehicle \cite{Matthews}, \cite{Lagstrom2015}.
-
-Several real-world studies of AV-pedestrian interaction noted that once a local population learned a vehicle was programmed to be perfectly safe, pedestrians would regularly take advantage of the AV and walk in front of it.
-\cite{Madigan}.
-This illustrates an issue with automated driving in that an overly conservative crossing algorithm will often be taken advantage of or cause confusion among pedestrians.
-Camara et al \cite{Camara2018} attempted to model the natural negotation for priority between a pedestrian and an AV at an intersection using game theory.
-
-\section{Proposed Research}
-
-\bibliographystyle{plain}
-\bibliography{Bibliography}
+\begin{center}
+\begin{tabular}{ l | r }
+ $U_y$ & Vehicle lateral velocity in body-fixed frame \\ 
+ $U_x$ & Vehicle longitudinal velocity in body-fixed frame \\ 
+ $r$ & Vehicle yaw rate \\   
+\end{tabular}
+\end{center}
 

### 2019-10-13 conference/iv2019/plotScripts/velocityPlotAutogen.m

@@ -1,0 +1,129 @@
+function createfigure(X1, Y1, S1, C1, X2, Y2, C2, Y3, Y4, Y5, Y6)
+%CREATEFIGURE(X1, Y1, S1, C1, X2, Y2, C2, Y3, Y4, Y5, Y6)
+%  X1:  scatter x
+%  Y1:  scatter y
+%  S1:  scatter s
+%  C1:  scatter c
+%  X2:  scatter x
+%  Y2:  scatter y
+%  C2:  scatter c
+%  Y3:  scatter y
+%  Y4:  scatter y
+%  Y5:  scatter y
+%  Y6:  scatter y
+
+%  Auto-generated by MATLAB on 18-Jan-2019 13:33:31
+
+% Create figure
+figure1 = figure;
+
+% Create subplot
+subplot1 = subplot(3,2,1,'Parent',figure1);
+hold(subplot1,'on');
+
+% Create scatter
+scatter(X1,Y1,S1,C1,'Parent',subplot1,'Marker','*');
+
+% Create ylabel
+ylabel(['Minimum Distance';'to Pedestrian(m)'],'FontSize',18);
+
+% Create title
+title('Rightmost Lane','FontSize',18);
+
+% Uncomment the following line to preserve the X-limits of the axes
+% xlim(subplot1,[0 9]);
+% Uncomment the following line to preserve the Y-limits of the axes
+% ylim(subplot1,[0 13]);
+grid(subplot1,'on');
+% Set the remaining axes properties
+set(subplot1,'FontSize',12);
+% Create axes
+axes1 = axes('Parent',figure1,...
+    'Position',[0.50266296276652 0.709264705882353 0.334659090909091 0.215735294117647]);
+hold(axes1,'on');
+
+% Create scatter
+scatter(X2,Y2,S1,C2,'Parent',axes1);
+
+% Create title
+title('Second to Right Lane','FontSize',18);
+
+% Uncomment the following line to preserve the X-limits of the axes
+% xlim(axes1,[0 9]);
+% Uncomment the following line to preserve the Y-limits of the axes
+% ylim(axes1,[0 9]);
+grid(axes1,'on');
+% Set the remaining axes properties
+set(axes1,'FontSize',12);
+% Create axes
+axes2 = axes('Parent',figure1,...
+    'Position',[0.127666277712952 0.458466981132076 0.33465909090909 0.215735294117647]);
+hold(axes2,'on');
+
[…]

### 2019-10-13 conference/iv2019/plotScripts/velocityPlot_withPOMDP.m

@@ -1,0 +1,193 @@
+clear all; close all; clc;
+ dat1 =load('/home/nkapania/catkin_ws/src/harpagornis/bags/out1.mat');
+ dat1p = load('/home/nkapania/catkin_ws/src/harpagornis/bags/out_pomdp_1.mat');
+%dat1 =load('/home/nkapania/catkin_ws/src/harpagornis/bags/out4.mat');
+%dat1p = load('/home/nkapania/catkin_ws/src/harpagornis/bags/out_pomdp_4.mat');
+
+
+dat1.lane = 3;
+
+% dat1 =load('/home/nkapania/catkin_ws/src/harpagornis/bags/out1.mat');
+% dat1p = load('/home/nkapania/catkin_ws/src/harpagornis/bags/out_pomdp_1.mat');
+dat2 =load('/home/nkapania/catkin_ws/src/harpagornis/bags/out2.mat');
+dat2p = load('/home/nkapania/catkin_ws/src/harpagornis/bags/out_pomdp_2.mat');
+dat2.lane = 2;
+
+laneWidth = 3.7;
+width = 1.5;
+
+dat1.vehicleX = dat1.lane * laneWidth + laneWidth /2 - width / 2;
+dat1p.vehicleX = dat1.vehicleX;
+dat2.vehicleX = dat2.lane * laneWidth + laneWidth /2 - width / 2;
+dat2p.vehicleX = dat2.vehicleX;
+pedestrianY = 28.0;
+
+%%
+N1 = length(dat1.inCrosswalk_bool);
+N1p = length(dat1p.inCrosswalk_bool);
+N2 = length(dat2.inCrosswalk_bool);
+N2p = length(dat2p.inCrosswalk_bool);
+
+dat1.pedXY = zeros(N1,2);
+dat1.vehXY = zeros(N1,2);
+dat1.distXY = zeros(N1,1);
+
+dat1p.pedXY = zeros(N1p,2);
+dat1p.vehXY = zeros(N1p,2);
+dat1p.distXY = zeros(N1p,1);
+
+for i = 1:N1
+    dat1.pedXY(i,:) = [dat1.xP(i) pedestrianY];
+    dat1.vehXY(i,:) = [dat1.vehicleX dat1.xV(i)];
+    dat1.distXY(i) = norm(dat1.pedXY(i,:) - dat1.vehXY(i,:));
+end
+
+for i = 1:N1p
+    dat1p.pedXY(i,:) = [dat1p.xP(i) pedestrianY];
[…]
